---
title: Data visualization with ggplot2
---

In this section, we use `ggplot2` to create basic graphs in R. The examples use the built-in `mpg` dataset from the `ggplot2` package.

```{r}
#| message: false
library(tidyverse)
```

## Scatterplots

Scatterplots are useful for studying relationships between numerical variables.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

A regression line can be added with `geom_smooth()`.

```{r}
#| message: false
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## Histograms

Histograms show the distribution of one numerical variable.

```{r}
#| message: false
ggplot(mpg, aes(x = hwy)) +
  geom_histogram()
```

You can change the bin width.

```{r}
ggplot(mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2)
```

## Density plots

Density plots show a smoother version of a distribution.

```{r}
ggplot(mpg, aes(x = hwy)) +
  geom_density()
```

## Bar charts

Bar charts are useful for categorical variables.

```{r}
ggplot(mpg, aes(x = class)) +
  geom_bar()
```

## Grouping with color

You can use color to show groups within the same graph.

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

## Faceting

Faceting creates separate panels for different groups.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class)
```

## Labels

Good graphs need clear labels.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  labs(
    title = "Fuel efficiency by engine size",
    x = "Engine displacement",
    y = "Highway miles per gallon"
  )
```

## Themes

Themes change the appearance of the graph.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  theme_minimal()
```

The goal of a graph is to make the data easier to understand. Always choose a graph that fits the type of variable and the question you want to answer.
