---
title: Event history analysis
---

## Assignments

Use the dataset `hschool`. The variables are: `caseid` (case identifier), `time` (time in months until school drop-out, or censoring), `died` (event indicator: 1 = dropped out, 0 = censored), `sex` (0 = male, 1 = female), `grades` (1 = high grades, 5 = low grades), `parttime` (1 = part-time student), `mrg` (month of marriage, coded as $12 \times (\text{year} - 1980) + \text{month}$; 99 = never married), and `stm` (month of college entry, coded the same way).

### Data description

1. How many students were married before entering college (i.e., `mrg` occurred before `stm`)?

2. How many students got married during college (i.e., `mrg` fell between `stm` and `stm + time`)?

### Linear regression on duration

3. We are interested in the effect of `sex` and `grades` on the time until school drop-out. Fit a linear regression of `time` on `sex` and `grades`, making sure you correctly restrict the sample to only those subjects for whom drop-out actually occurred. Interpret all coefficients, including the intercept — and note anything odd or hard to interpret about the intercept.

### Cox proportional hazards model

4. Fit a Cox proportional hazards model of drop-out on `sex` and `grades`. Interpret each coefficient (a) as an effect on the log hazard, and (b) as a hazard ratio.

5. Compare the statistical significance of `sex` and `grades` in the Cox model to their significance in the linear regression from question 3. Are the conclusions the same? If not, why is this not a contradiction?

### Plotting the process

6. Plot the hazard function, the cumulative hazard function, and the survival function implied by the Cox model.

### Testing the proportional hazards assumption

7. Test the proportional hazards assumption for the Cox model of question 4. Does the assumption hold?

### BONUS: discrete-time analysis

8. Construct a person-period file from `hschool`: one row per subject per month at risk, with an event indicator equal to 1 only in the month drop-out actually occurred (0 in all other months, including for censored subjects). Fit a discrete-time event history model using logistic regression of the event indicator on `sex` and `grades`. What do you conclude, and how does this compare to the conclusions from the Cox model?

## Answers

Make sure this file is in the same R Project as the `data` folder. Use relative file paths, not full paths to your computer.

```{r setup-eha}
library(tidyverse)
library(haven)
library(survival)
library(survminer)

hschool <- read_dta("data/hschool.dta")

hschool <- hschool |>
  mutate(female = sex)
```

### Data description

#### 1. Married before entering college

```{r}
hschool |>
  filter(mrg < stm) |>
  nrow()
```

#### 2. Married during college

```{r}
hschool |>
  filter(mrg >= stm, mrg <= stm + time) |>
  nrow()
```

### Linear regression on duration

#### 3. A linear regression on completed durations

```{r}
fit_lm <- lm(time ~ sex + grades, data = hschool |> filter(died == 1))
summary(fit_lm)

# Females drop out about 0.075 months faster than males, but this is
# not significant.
# People with lower grades (higher values of grades, since 1 = high,
# 5 = low) drop out faster, but this is not significant either.
# Males (sex = 0) with the highest grades (grades = 0, which is outside
# the actual scale of 1-5 and should ideally be recentered) are
# predicted to drop out at 18.55 months. Because grades = 0 does not
# occur in the data, this intercept is not substantively meaningful as
# it stands.
```

### Cox proportional hazards model

#### 4. The Cox model

```{r}
fit_cox <- coxph(Surv(time, died) ~ sex + grades, data = hschool)
summary(fit_cox)

# Log hazard interpretation:
# Females have a 0.240 higher log hazard of dropping out than males (a
# shorter time to drop-out), but this is not significant.
# People with lower grades have a significantly higher log hazard of
# dropping out (a shorter time to drop-out).
#
# Hazard ratio interpretation (exp(coef), shown directly in the
# summary() output):
# Females have a 27% higher hazard of dropping out than males, but this
# is not significant.
# People with lower grades have a 39.73% significantly higher hazard of
# dropping out for each unit increase on the grades scale (i.e., each
# step toward lower grades).
```

#### 5. Comparing the two models

```{r}
# The significance of the predictors is not the same between the linear
# regression (question 3) and the Cox regression (question 4): grades is
# significant in the Cox model but not in the linear regression. This is
# not a contradiction. The linear regression restricts the sample to
# completed durations only and assumes normally distributed, constant-
# variance residuals on the raw time scale -- assumptions that are
# typically violated for duration data. The Cox model instead uses all
# subjects (including censored ones) and does not require these
# distributional assumptions, which is why it can detect an effect the
# linear model misses.
```

```{r}
hschool |>
  count(died)
# how many cases the linear regression discards as censored
```

### Plotting the process

#### 6. Hazard, cumulative hazard, and survival

```{r}
fit_surv <- survfit(fit_cox)

# Cumulative hazard
plot(fit_surv, fun = "cumhaz", xlab = "Time (months)", ylab = "Cumulative hazard")

# Survival function
plot(fit_surv, xlab = "Time (months)", ylab = "Survival probability")

# Using survminer for nicer plots, including an approximate hazard shape
# via the cumulative hazard and survival curves:
ggsurvplot(fit_surv, data = hschool, fun = "cumhaz")
ggsurvplot(fit_surv, data = hschool)

# Note: a smoothed hazard function itself (as opposed to the cumulative
# hazard) is not directly returned by survfit(); it can be approximated
# with a kernel-smoothed estimate, e.g. using the muhaz package, if a
# separate plot of the hazard function itself (rather than the
# cumulative hazard) is required.
```

### Testing the proportional hazards assumption

#### 7. The Schoenfeld residual test

```{r}
cox.zph(fit_cox)

# The test does not reject the proportional hazards assumption at
# conventional significance levels: the proportional hazards assumption
# appears to hold for this model.
```

### BONUS: discrete-time analysis

#### 8. A person-period file and a discrete-time model

```{r}
# Method 1: expand the data manually to one row per subject per month
person_period <- hschool |>
  dplyr::select(caseid, time, sex, grades, died) |>
  rowwise() |>
  mutate(period = list(seq_len(time))) |>
  unnest(period) |>
  ungroup() |>
  rename(end_time = time) |>
  mutate(
    time = period,
    died = if_else(died == 1 & time != end_time, 0L, as.integer(died))
  ) |>
  dplyr::select(caseid, time, end_time, sex, grades, died)

fit_discrete <- glm(died ~ sex + grades, data = person_period, family = binomial)
summary(fit_discrete)

# The discrete-time logistic model leads to the same substantive
# conclusion as the Cox model: grades significantly predicts drop-out
# (lower grades associated with higher risk of dropping out in a given
# month), while sex does not significantly predict drop-out. This
# confirms that the Cox model's conclusions are not an artifact of the
# continuous-time approach; a discrete-time formulation using ordinary
# logistic regression on the person-period file gives essentially the
# same answer.
```

```{r}
nrow(hschool)
nrow(person_period)
sum(person_period$died)
sum(hschool$died)
# one row per subject per month at risk; the number of events is
# unchanged, since each subject can drop out only once
```
