---
title: Multiple linear regression
---

In this part you will learn about the following basic tasks:

- Fit a multiple linear regression model with several predictors.

- Interpret a partial regression coefficient: the effect of one predictor, holding the other predictors constant.

- Distinguish a marginal (bivariate) relationship from a partial (controlled) relationship.

- Use categorical predictors in a regression model through factor variables (dummy coding), and choose a reference category.

- Test whether groups differ, and perform post-hoc pairwise comparisons between categories.

- Include and interpret interaction terms between two quantitative predictors, between a quantitative and a categorical predictor, and between two categorical predictors.

- Include and interpret a quadratic term to model a curved (nonlinear) relationship.

A multiple regression model extends simple linear regression to more than one predictor:

$$
E(Y) = b_0 + b_1 X_1 + b_2 X_2 + \dots + b_k X_k
$$

Each partial regression coefficient $b_j$ is interpreted as the expected change in Y for a one-unit increase in $X_j$, *holding all other predictors in the model constant*. This is different from a simple regression coefficient, which describes the marginal (uncontrolled) relationship between Y and a single predictor. A predictor can be significantly related to Y in a simple regression, yet not contribute once other predictors are controlled for — this happens when the predictor's relationship with Y runs (partly) through the other predictors in the model.

$R^2$ in a multiple regression is the proportion of variance in Y explained jointly by all predictors. The overall F-test evaluates the null hypothesis that all slopes are simultaneously zero (equivalently, that $R^2 = 0$ in the population).

## Categorical predictors

A categorical predictor with more than two categories cannot be entered directly as a single numeric variable; instead, it is represented by a set of dummy variables, one for each category except a reference category. R handles this automatically: declare the variable as a factor with `factor()`, or use a character or factor variable directly in the model formula, and R constructs the dummy variables for you. No special formula syntax is required.

Each coefficient for a category then represents the difference in expected Y between that category and the reference category, holding other predictors constant. To test all pairwise differences between categories (not just differences with the reference category), use post-hoc pairwise comparisons.

## Interactions and quadratic terms

An interaction term between two predictors allows the effect of one predictor to depend on the level of the other. For two quantitative predictors, this is included as their product; for a quantitative and a categorical predictor, this allows different slopes in each category; for two categorical predictors, this allows the difference between categories of one factor to differ across categories of the other factor. A significant interaction means that the "no interaction" (additive) model is not adequate; an interaction that is not significant means there is no evidence that the effect of one predictor depends on the other.

A special case of interaction is a quadratic term, $X^2$, added together with $X$ itself. This allows a curved relationship. If the coefficient of $X^2$ is positive, the curve is U-shaped; if negative, it is inverted-U shaped. The turning point (the value of X at which the curve changes direction) is found at $X = -b_1 / (2 b_2)$. Always check this turning point against the actual range of the data — a turning point far outside the observed range of X, or based on very few observations, should not be over-interpreted.

## R

You will learn the R functions listed below to fit and inspect multiple linear regression models.

| Function | Description |
|----|----|
| `lm()` | Fit a linear regression model with one or more predictors |
| `summary()` | Display coefficients, standard errors, F-test, and $R^2$ / adjusted $R^2$ |
| `factor()` | Declare a variable as categorical, so it is entered as a set of dummy variables |
| `relevel()` | Change the reference category of a factor |
| `emmeans()` | Estimated marginal means and pairwise comparisons between categories |
| `anova()` | Compare nested models (e.g., with and without a set of predictors) |
| `I()` | Include a computed term (e.g., a squared term) directly in a formula |
| `poly()` | Include polynomial terms (e.g., quadratic) in a formula |

Examples:

Fit a multiple regression with several quantitative predictors:

```{r, eval=FALSE}
library(tidyverse)
library(haven)

data <- read_dta("data/sesam.dta")

fit <- lm(postnumb ~ age + prenumb, data = data)
summary(fit)
```

The `summary()` output reports `Adjusted R-squared` in addition to `Multiple R-squared`; for multiple regression, report the adjusted value, since it corrects for the number of predictors in the model.

Compare a controlled (partial) coefficient to the uncontrolled (marginal) coefficient from a simple regression:

```{r, eval=FALSE}
fit_simple <- lm(postnumb ~ prenumb, data = data)
summary(fit_simple)
```

Use a categorical predictor with several categories. Declare it as a factor and include it in the formula; R automatically creates dummy variables using the first level as the reference category:

```{r, eval=FALSE}
data <- data |>
  mutate(religion = factor(religion))

fit_rel <- lm(donation ~ religion + inc, data = data)
summary(fit_rel)
```

Change the reference category of a factor:

```{r, eval=FALSE}
data <- data |>
  mutate(religion = relevel(religion, ref = "3"))

fit_rel2 <- lm(donation ~ religion + inc, data = data)
summary(fit_rel2)
```

Post-hoc pairwise comparisons between all categories of a factor, adjusting for control variables in the model:

```{r, eval=FALSE}
library(emmeans)

emmeans(fit_rel, pairwise ~ religion)
```

Include an interaction between two quantitative predictors:

```{r, eval=FALSE}
fit_int <- lm(percdon ~ age * inc, data = data)
summary(fit_int)
```

The `*` operator automatically includes the main effects and the interaction term (`age + inc + age:inc`). Use `:` instead of `*` to include only the interaction term itself.

Include an interaction between a categorical and a quantitative predictor (different slopes per category):

```{r, eval=FALSE}
fit_int2 <- lm(percdon ~ male * inc, data = data)
summary(fit_int2)
```

Include an interaction between two categorical predictors:

```{r, eval=FALSE}
fit_int3 <- lm(percdon ~ male * religion, data = data)
summary(fit_int3)
```

Include a quadratic term to model a curved relationship:

```{r, eval=FALSE}
fit_quad <- lm(prestg80 ~ educ + I(educ^2) + race + region + sex, data = data)
summary(fit_quad)
```

Find the turning point of a quadratic relationship from the fitted coefficients:

```{r, eval=FALSE}
b_educ <- coef(fit_quad)["educ"]
b_educ2 <- coef(fit_quad)["I(educ^2)"]

-b_educ / (2 * b_educ2)
```

Overlay a scatterplot with both a linear and a quadratic fit, to visually compare the two:

```{r, eval=FALSE}
ggplot(data, aes(x = educ, y = prestg80)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "blue") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "red")
```

## Assignments

### Predicting a post-test score

1. Use the dataset `sesam`. We predict `postnumb` (a post-test score) from `age` and `prenumb` (a pre-test score).

    - Fit a regression of `postnumb` on `age` and `prenumb`. Interpret each coefficient, the intercept, and the adjusted $R^2$. Is `age` a significant predictor once `prenumb` is controlled for?

    - Compare this result to a simple regression of `postnumb` on `age` alone (from the previous chapter). Why can a predictor be significantly related to the outcome in a simple regression, yet not contribute in a multiple regression?

    - Fit a regression of `postnumb` on `prenumb` alone. Compare the coefficient for `prenumb` and the $R^2$ to the model that also includes `age`. What do you conclude?

### Donations, religion, and income

Use the dataset `donations`. Treat the "miscellaneous" category of `religion` as missing, and exclude the corresponding observations from the analyses below (it is too heterogeneous a group to interpret meaningfully).

2. Test whether average donations differ between religious groups, controlling for income (`inc`). Fit a regression of `donation` on `religion` (as a factor) and `inc`. Is the overall F-test for the model significant? Which categories differ significantly from the reference category?

3. Is there a difference in average donation between two specific religious categories that are *not* both compared to the reference category in the previous output? Use `emmeans()` for pairwise comparisons between all categories, or refit the model with a different reference category, to answer this.

4. Yearly donations appear to have been recorded in thousands. Construct a new variable `propdon`, the proportion of monthly income donated: divide `donation` (converted to a monthly amount) by monthly income. Report the mean and standard deviation of `propdon`. What do these values tell you?

5. Excluding the miscellaneous religion category permanently from the dataset, fit a regression of `propdon` on `religion` and obtain post-hoc pairwise comparisons between all categories. Between which categories is there no evidence of a difference?

6. Use `percdon` (percentage of income donated) as outcome. Fit a regression on `inc` alone, and then on `inc` together with `religion`. Is there evidence that donations increase with income, even after controlling for religion? Interpret the size of the effect.

7. Fit a regression of `percdon` on `male`, `religion`, and their interaction. Is there evidence that the effect of religion on donations differs between men and women?

8. Fit a regression of `percdon` on `inc`, `religion`, and their interaction. Is there evidence that the effect of income on donations varies between religious denominations?

9. Construct the product of `age` and `lifevent` (an indicator for having experienced a serious health problem). Fit a regression of `percdon` on `age`, `lifevent`, and their interaction (both by explicitly creating the product term, and using formula syntax). Is there evidence that the effect of `lifevent` on donations is larger for older people than for younger people?

10. Fit a regression of `percdon` on `age`, `inc`, and their interaction. Is there evidence that the effect of income on donations varies with age?

11. Fit a regression of `percdon` on `male`, `inc`, and their interaction. Is there evidence that the effect of income on donations differs between men and women?

12. Fit a regression of `percdon` on `male`, `inc`, `religion`, and all their two-way and three-way interactions. Is there evidence that gender differences in the effect of income on donations vary by religion?

### Occupational prestige and education

Use the dataset `gss91_extended`. We study the relationship between years of education (`educ`) and occupational prestige (`prestg80`).

13. Make a scatterplot of `prestg80` against `educ`. Overlay it with both a linear and a quadratic fitted line. Which fits the data better, based on visual inspection?

14. Fit a regression model of `prestg80` on `educ`, a squared term for `educ`, and the control variables `race`, `region`, and `sex`. Is the squared term statistically significant? What do you conclude about the shape of the relationship?

15. Using the coefficients for `educ` and `educ`$^2$, compute the turning point of the curve: the value of `educ` at which expected prestige is at its minimum or maximum. Is the curve U-shaped or inverted-U shaped? Check how many observations in the data actually fall at or below this turning point, and discuss whether the turning point should be interpreted, or whether it is more appropriate to describe the relationship only over the range where most of the data lie.

## Answers

Make sure this file is in the same R Project as the `data` folder. Use relative file paths, not full paths to your computer.

```{r setup-multiple-regression}
library(tidyverse)
library(haven)
library(emmeans)

sesam <- read_dta("data/sesam.dta")
donations <- read_dta("data/donations.dta")
gss91 <- read_dta("data/gss91_extended.dta")
```

### Predicting a post-test score

#### 1. Marginal versus partial effects

```{r}
fit_multi <- lm(postnumb ~ age + prenumb, data = sesam)
summary(fit_multi)

# Adj. R-squared = 0.45
#
# E(postnumb) = 7.42 + 0.12 age + 0.78 prenumb      (SE_age = 0.11, SE_prenumb = 0.06)
#
# Age does not contribute in predicting postnumb (not significant). The
# marginal correlation between age and postnumb is significant, but here
# the effect of age is conditional on that of prenumb: once you control
# for prenumb, age does not matter.
# If pretest score goes up by 1, posttest score significantly goes up by
# 0.78, irrespective of age.
# People aged 0 with a pretest score of 0 have a posttest score of 7.42;
# interpreting the constant is not meaningful here.
# The F-test p-value is 0.000, so not all coefficients are 0 (R-squared
# is significant). 45% of the variation in the post-test is explained by
# the pre-test and age.
```

```{r}
fit_age_only <- lm(postnumb ~ age, data = sesam)
summary(fit_age_only)

# In the simple regression, age is significantly related to postnumb.
# This is because age is correlated with prenumb; once prenumb is added
# to the model, the (partial) effect of age drops out. A predictor can be
# marginally related to Y through its association with other predictors,
# without having an independent (partial) effect once those predictors
# are controlled for.
```

The correlation between the two predictors is what drives this.

```{r}
sesam |>
  summarise(r_age_prenumb = cor(age, prenumb, use = "complete.obs"))
# older children score higher on the pre-test, which is why age looks
# informative until the pre-test is in the model
```

```{r}
fit_prenumb_only <- lm(postnumb ~ prenumb, data = sesam)
summary(fit_prenumb_only)

# Very little has changed compared to the model with both predictors:
# the coefficient for prenumb and R-squared are almost the same with or
# without age in the model. This confirms that age adds essentially
# nothing once prenumb is controlled for.
```

A formal test of whether adding `age` improves the model.

```{r}
anova(fit_prenumb_only, fit_multi)
# comparing nested models: a non-significant F confirms that age adds
# nothing beyond prenumb
```

### Donations, religion, and income

#### 2. Donations by religion, controlling for income

```{r}
table(donations$religion, useNA = "ifany")
# check the codes before excluding the miscellaneous category
```

```{r}
donations <- donations |>
  filter(religion != 6) |>
  mutate(religion = factor(religion))

levels(donations$religion)
# the first level becomes the reference category
```

```{r}
fit_rel <- lm(donation ~ religion + inc, data = donations)
summary(fit_rel)

# There are differences in the average donations of the various religious
# groups. The null hypothesis that R-squared = 0 (p-value of the F-test)
# is rejected. From the coefficients we learn that there are significant
# differences in average donations between catholics and the reference
# category (none), between reformed and none, and between rereformed and
# none. There is no evidence of a difference in average donation between
# islam and none.
```

The individual t-tests answer a different question from the joint test of whether religion matters at all.

```{r}
anova(fit_rel)
# the F-test for religion tests all its dummy variables simultaneously
```

#### 3. Pairwise comparisons between categories

```{r}
emmeans(fit_rel, pairwise ~ religion)

# Pairwise comparisons show whether specific pairs of categories differ
# significantly, not just each category versus the reference category.
# Alternatively, releveling the factor to "reformed" as reference and
# refitting the model gives the same information for that specific
# comparison:

donations <- donations |>
  mutate(religion = relevel(religion, ref = "3"))

fit_rel_releveled <- lm(donation ~ religion + inc, data = donations)
summary(fit_rel_releveled)

# reset back to the original reference category for later use
donations <- donations |>
  mutate(religion = relevel(religion, ref = "1"))
```

Note that `emmeans()` adjusts for multiple testing by default (Tukey), while releveling and reading off a t-test does not. With several categories the two can disagree on borderline comparisons; the adjusted comparison is the safer one to report.

#### 4. Proportion of income donated

```{r}
donations <- donations |>
  mutate(propdon = ((donation * 1000) / 12) / (inc * 1000))

donations |>
  summarise(
    mean_propdon = mean(propdon, na.rm = TRUE),
    sd_propdon = sd(propdon, na.rm = TRUE)
  )

# Yearly donations have been recorded in thousands, hence
# the *1000 conversion. The mean propdon = 0.0177, and the sd = 0.02235.
# The mean proportion of monthly income donated is about 0.02. The sd is
# relatively large compared to the mean.
```

The two factors of 1000 cancel, so the expression simplifies. Writing it out in full is nevertheless clearer about what is being converted.

```{r}
donations |>
  mutate(propdon_short = donation / (12 * inc)) |>
  summarise(max_difference = max(abs(propdon_short - propdon), na.rm = TRUE))
# the difference should be zero
```

A standard deviation larger than the mean, for a variable that cannot be negative, points to a strongly skewed distribution.

```{r}
ggplot(donations, aes(x = propdon)) +
  geom_histogram(bins = 30)
# most people donate a small proportion, a few donate a great deal more
```

#### 5. Proportion donated by religion

The miscellaneous category was already removed in assignment 2, so this filter changes nothing; it is kept for clarity.

```{r}
donations <- donations |>
  filter(religion != 6)

nrow(donations)
```

```{r}
fit_propdon <- lm(propdon ~ religion, data = donations)
summary(fit_propdon)

emmeans(fit_propdon, pairwise ~ religion)

# No significant differences are found between islam and none, and
# between catholics and reformed (the confidence interval for these
# pairwise differences includes 0).
```

#### 6. Percentage donated and income

```{r}
fit_percdon_inc <- lm(percdon ~ inc, data = donations)
summary(fit_percdon_inc)

fit_percdon_full <- lm(percdon ~ inc + religion, data = donations)
summary(fit_percdon_full)

# There is evidence that donations increase with income, even controlling
# for religion (b = 0.424, t = 5.49, p < 0.001). For each unit (1000)
# increase in monthly income, the percentage of income donated increases
# by about 0.4 percentage points.
```

Note what this coefficient says: richer people donate a larger *share* of their income, not merely a larger amount. That is a substantively stronger claim than a positive effect on `donation` would be.

```{r}
bind_rows(
  tibble(model = "inc only", b_inc = coef(fit_percdon_inc)["inc"],
         adj_r2 = summary(fit_percdon_inc)$adj.r.squared),
  tibble(model = "inc + religion", b_inc = coef(fit_percdon_full)["inc"],
         adj_r2 = summary(fit_percdon_full)$adj.r.squared)
)
# the income effect barely changes when religion is added, so religion does
# not confound the income effect
```

#### 7. Does the religion effect differ by gender?

```{r}
fit_male_rel <- lm(percdon ~ male * religion, data = donations)
summary(fit_male_rel)

# There is no evidence that the effect of religion on donations differs
# for men and women (the interaction terms are not significant).
```

With several interaction terms, test them jointly rather than one at a time.

```{r}
anova(lm(percdon ~ male + religion, data = donations), fit_male_rel)
# a single F-test for the whole set of interaction terms
```

#### 8. Does the income effect differ by religion?

```{r}
fit_inc_rel <- lm(percdon ~ inc * religion, data = donations)
summary(fit_inc_rel)

# There is no evidence that the effect of income on donations varies
# between the religious denominations (the interaction terms are not
# significant).
```

```{r}
anova(lm(percdon ~ inc + religion, data = donations), fit_inc_rel)
# joint test of the interaction terms
```

#### 9. Age and a serious health problem

```{r}
donations <- donations |>
  mutate(age_lifevent = age * lifevent)

fit_age_life_manual <- lm(percdon ~ age + lifevent + age_lifevent, data = donations)
summary(fit_age_life_manual)

fit_age_life_formula <- lm(percdon ~ age * lifevent, data = donations)
summary(fit_age_life_formula)

# There is no evidence that the effect of having experienced a serious
# health problem on donations is larger for older than for younger
# people (the interaction term is not significant).
```

```{r}
all.equal(coef(fit_age_life_manual), coef(fit_age_life_formula), check.attributes = FALSE)
# the two routes give identical estimates; the formula syntax is just shorter
# and cannot go wrong about which main effects to include
```

#### 10. Does the income effect vary with age?

```{r}
donations <- donations |>
  mutate(age_inc = age * inc)

fit_age_inc_manual <- lm(percdon ~ age + inc + age_inc, data = donations)
summary(fit_age_inc_manual)

fit_age_inc_formula <- lm(percdon ~ age * inc, data = donations)
summary(fit_age_inc_formula)

# There is no evidence that the effect of monthly income on donations
# varies with age (the interaction term is not significant).
```

In a model with an interaction, the coefficient for `inc` is the effect of income at `age` = 0, not an average effect. Centring the predictors makes the main effects interpretable again.

```{r}
lm(
  percdon ~ I(age - mean(age, na.rm = TRUE)) * I(inc - mean(inc, na.rm = TRUE)),
  data = donations
) |>
  summary()
# now the main effects refer to a person of average age and average income
# the interaction term itself is unchanged
```

#### 11. Does the income effect differ between men and women?

```{r}
donations <- donations |>
  mutate(male_inc = male * inc)

fit_male_inc_manual <- lm(percdon ~ male + inc + male_inc, data = donations)
summary(fit_male_inc_manual)

fit_male_inc_formula <- lm(percdon ~ male * inc, data = donations)
summary(fit_male_inc_formula)

# There is no evidence that the effect of monthly income on donations
# differs for men and women (the interaction term is not significant).
```

#### 12. A three-way interaction

```{r}
fit_full_interaction <- lm(percdon ~ male * inc * religion, data = donations)
summary(fit_full_interaction)

# There is no evidence that possible gender differences in the effect of
# monthly income on donations vary with religion (the three-way
# interaction terms are not significant).
```

```{r}
anova(
  lm(percdon ~ (male + inc + religion)^2, data = donations),
  fit_full_interaction
)
# a single test of the three-way interaction terms against the model with
# only two-way interactions
```

```{r}
donations |>
  count(male, religion) |>
  arrange(n) |>
  slice(1:5)
# a three-way interaction subdivides the data very finely
# with small cells, non-significance mainly reflects a lack of power rather
# than evidence that no such interaction exists
```

### Occupational prestige and education

#### 13. Linear and quadratic fits

```{r}
ggplot(gss91, aes(x = educ, y = prestg80)) +
  geom_point()

ggplot(gss91, aes(x = educ, y = prestg80)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "blue") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "red")

# The quadratic (red) line explains more variance than the linear (blue)
# line: R-squared = 0.291 versus 0.27 for the linear fit. Whether this
# difference is significant needs to be tested formally, e.g. through an
# interaction of educ with itself (a squared term) in a regression model.
```

Overplotting hides how many observations sit at each level of education.

```{r}
ggplot(gss91, aes(x = educ, y = prestg80)) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "red")
# jittering and transparency reveal where the data actually are
# note how few observations lie at the low end of educ
```

#### 14. A quadratic model with controls

```{r}
fit_quad <- lm(
  prestg80 ~ educ + I(educ^2) + factor(race) + factor(region) + factor(sex),
  data = gss91
)
summary(fit_quad)

# The educ-squared term is significant (t(1407) = 6.46, p < 0.0001), so
# there is evidence of a quadratic relationship, controlling for race,
# region, and sex.
```

```{r}
fit_linear <- lm(
  prestg80 ~ educ + factor(race) + factor(region) + factor(sex),
  data = gss91
)

anova(fit_linear, fit_quad)
# the same conclusion as a nested model comparison
```

#### 15. The turning point

```{r}
b_educ <- coef(fit_quad)["educ"]
b_educ2 <- coef(fit_quad)["I(educ^2)"]

turning_point <- -b_educ / (2 * b_educ2)
turning_point

# -b[educ] / (2 * b[educ2]) = -(-1.081) / (2 * 0.130) = 4.2
#
# Since b[educ2] > 0, the relation is U-shaped, as suggested by the
# scatterplot (though that plot did not include statistical controls).
# We conclude that around 4 years of education corresponds to the lowest
# predicted prestige.

gss91 |>
  summarise(n_at_or_below = sum(educ <= turning_point, na.rm = TRUE))

# Only a small number of subjects have educ values at or below the
# turning point (about 12 out of roughly 1500). It is highly unlikely
# that prestige is actually higher with 1 or 2 years of education than
# with 4 years; the model most plausibly applies for 4 and more years of
# education. We therefore abstain from interpreting the turning point
# itself, and conclude instead that, over the range where most of the
# data lie, prestige increases in education at an increasing rate.
```

```{r}
gss91 |>
  summarise(
    n = n(),
    at_or_below = sum(educ <= turning_point, na.rm = TRUE),
    percent = 100 * mean(educ <= turning_point, na.rm = TRUE)
  )
# the proportion of the sample on the descending part of the curve
```

```{r}
ggplot(gss91, aes(x = educ)) +
  geom_histogram(binwidth = 1) +
  geom_vline(xintercept = turning_point, linetype = "dashed")
# the dashed line marks the turning point
# almost the entire distribution lies to the right of it
```

The general lesson: a quadratic term is fitted to the whole range of X, but the fit is only trustworthy where the data are dense. A turning point located in a sparse region is an artefact of the functional form you imposed, not a finding about the world.

