---
title: Binary logistic regression
---

## Assignments

Sociologists often claim that labor market outcomes (e.g., income, promotion, reentry after unemployment, ...) depend not only on human capital factors (formal qualifications, labor market experience, ...), but also on social capital: it is not only what you know, but also who you know.

Use the dataset `unemploy`. Construct the outcome `work`, coded 1 (or TRUE) if `studytim` (the unemployment spell length) is less than 53 (weeks), and 0 (or FALSE) otherwise. Human capital is captured by `edu` (education), `age`, and `exp` (number of previous unemployment spells).

1. Fit a linear probability model of `work` on `age`, `edu`, and `exp` and interpret all 
coefficients. 

2. Compute heteroskedasticity-consistent (robust) standard errors for the previous models. You can do that using `modelsummary(your_model_object, vcov = "robust")`. Are the substantive conclusions affected?

3. Obtain the predicted probabilities from the linear probability model (use the `predict()` from base R, or the values from `your_model_object$fitted.values`). Count how many predicted probabilities fall outside the range 0 to 1. 


4. Fit a logistic regression of `work` on `age`, `edu`, and `exp`. (Use `glm`, which has the basically the same syntax as `lm()`. But you should specify the argument `family = binomial()`.) Interpret each coefficient:
  a. in terms of log-odds
  b. in terms of odds-ratios/odds
  c. in terms of marginal effects (on the probability scale). You can compute marginal effects using `marginaleffects::avg_slopes()`.  Compare these effects, and the substantive conclusions they imply, to the coefficients of the linear probability model from question 1.

5. Compute the marginal effect of `age` for a person with the lowest education level (`edu = 1`) and the largest number of prior unemployment spells (`exp = 3`). Use `marginaleffects::slopes()` with the `newdata` argument to specify the values of `edu` and `exp` (see example [here](https://marginaleffects.com/chapters/slopes.html#adjustment-variables)). Interpret this effect.

6. Compute the marginal effect of `age` at the means of `edu` and `exp`. (Use `newdata = "mean"` in `marginaleffects::slopes()`.) Compare this to the average marginal effect of `age` found in question 4c. Why can these two values differ?

7. The dataset comprises measures for three forms of social recourses (strong, medium, and weak ties): `scweak`, `scmedium`, and `scstrong`. Fit a logit model of `work` on `age`, `edu`, `exp`, and the three social capital measures.
  a. Test, using a Wald test and a likelihood ratio test, whether social capital as a set significantly contributes to explaining re-employment.
  b. Test whether the coefficients for `scweak` and `scstrong` are different. (Hint: use `marginaleffects::hypotheses()`)

8. Fit a probit regression of `work` on `age`, `edu`, and `exp`. (The code is almost the same as for the logit model, but with `family = binomial(link = "probit")`) Compare the estimates of this model with the estimates of the linear probability model question 1 and the average marginal effects using the logistic regression model estimated in 4c. What do you find? 


::: {.content-visible when-format="html"}

## Answers

::: {.callout-note appearance="simple" collapse=true}
## See answers


```{r setup-logistic}
library(tidyverse)
library(haven)
library(modelsummary)
library(lmtest)
library(marginaleffects)

unemploy <- read_dta("data/unemploy.dta")

unemploy <- unemploy |>
  mutate(work = studytim < 53)
```

### 1

```{r}
fit_lpm <- lm(work ~ age + edu + exp, data = unemploy)
summary(fit_lpm)
```

If age increases by one year, the probability of re-employment significantly decreases by 3.25 percentage points. If education increases, the probability of re-employment significantly increases by 5.6 percentage points per unit of education. If the number of previous unemployment spells (exp) is higher, the probability of re-employment does not decrease significantly at the 5% level, but does decrease significantly at the 7.7% level (by about 4.8 percentage points per additional spell).

### 2

Model with robust SEs:
```{r}
modelsummary(fit_lpm, vcov = "robust", stars=TRUE)
```

We can also compare the model with and without robust SEs side by side:
```{r}
modelsummary(list("original" = fit_lpm, "robust SES" = fit_lpm), 
vcov = list("classical", "robust"), stars=TRUE)
```

The coefficients are exactly the same (as they should); only the standard errors change very slightly. The substantive conclusions are robust to this correction. Note that it never hurts to use robust SEs, even if the model is homoskedastic, because they are valid in both cases. So it is a good idea to use robust SEs by default.

### 3

```{r}
pred_probs <- predict(fit_lpm)

sum(pred_probs < 0 | pred_probs > 1)
```

None of the observations have predicted probabilities outside the  range 0 to 1. In practice, this is very often the case, even though the linear probability model does not guarantee it in general.

### 4
```{r}
fit_logit <- glm(work ~ age + edu + exp, data = unemploy, family = binomial)
summary(fit_logit)
```

a. Logg odds:  If age increases by one year, the log odds of re-employment significantly go down by 0.143. If education increases, the log odds of re-employment significantly
increase by 0.259. If the number of previous unemployment spells is higher, the log odds
of re-employment do not decrease significantly at the 5% level, but do decrease significantly at the 7.4% level (by 0.216).

b. Odds ratios. We need to exponentiate the coefficients:
```{r}
# Solution 1:
exp(coef(fit_logit))

# Solution 2
modelsummary(fit_logit, exponentiate = TRUE, stars = TRUE)
```

The odds of getting a job within a year (the probability of getting a job divided by the probability of not getting a job) change by (exp(b) - 1) * 100% for a one-unit increase in x:
  - the odds increase by (exp(0.259) - 1) * 100 = 29.5% for one unit
    of education
  - the odds change by (exp(-0.143) - 1) * 100 = -13.3%, i.e. decrease
    by 13.3%, for each year older
  - the odds decrease by about 19.5% for each previous unemployment
    spell (exp(-0.216))
The advantage of the odds ratio interpretation is that it applies exactly for every subject, irrespective of the value of the focal variable and of the values of the other predictors -- unlike a marginal effect, which depends on the specific values of all predictors.


c. Marginal effects
```{r}
avg_slopes(fit_logit)
```

With each year of age, the probability of re-employment decreases on average by about 3.1 percentage points. The effects of the other predictors follow similarly. These average marginal effects are very close to the coefficients of the linear probability model in question 1. In practice, this is very often the case.

We can even compare the average marginal effects to the coefficients of the linear probability model side by side usign `modelsummary()`:

```{r}
modelsummary(list("LPM"= fit_lpm, 
                   "Logit - Marginal Effects" = avg_slopes(fit_logit)), 
              stars = TRUE) 
```

### 7
```{r}
slopes(fit_logit, variables = "age", newdata = datagrid(edu = 1, exp = 3))
```

For the lowest educated people with the largest number of prior unemployment spells, a one-year increase in age is associated with a significant decrease in the probability of re-employment of about 3.0 percentage points.

### 8

```{r}
slopes(fit_logit, variables = "age", newdata = "mean")
```

For a person with average education and an average number of prior unemployment spells, a one-year increase in age is associated with a significant decrease in the probability of re-employment of about 3.6 percentage points. This differs from the average marginal effect in question 4c because there we averaged the marginal effect over all observed combinations of education and prior unemployment spells, rather than evaluating the effect once at the mean values of these predictors.


### 7.

```{r}
fit_soccap <- glm(work ~ age + edu + exp + scweak + scmedium + scstrong,
                data = unemploy, family = binomial)

# Wald test 
hypotheses(fit_soccap, joint = c("scweak", "scmedium", "scstrong"))

# Likelihood ratio test 
lrtest(fit_logit, fit_soccap)
```

At a 5% significance level,there is no evidence that social capital affects re-employment, controlling for human capital.

Are the effects of `scweak` and `scstrong` different?

```{r}
hypotheses(fit_soccap, "scweak = scstrong")
```
We have evidence that the effects of `scweak` and `scstrong` are indeed different.


### 8

```{r}
fit_probit <- glm(work ~ age + edu + exp, data = unemploy, family = binomial(link = "probit"))

modelsummary(list("LPM"= fit_lpm, 
                   "Logit - Marginal Effects" = avg_slopes(fit_logit), 
                   "Probit" = fit_probit),
              stars = TRUE) 
```

The coefficients are larger, but the conclusions are the same. Indeed, the conclusions between the three models are very often very similar.

:::
:::