---
title: Model testing
---

## Assignments

### sesam

Use the dataset `sesam`.

1. Fit a restricted model of `postnumb` on `age` and `prenumb`, and a full model that adds `prelet`, `prerelat`, and `peabody`. Does adding this set of three predictors significantly improve the overall fit? Test this in two ways:
  a. A Wald test using the F-statistic. There are several ways to do this in R:
    - `anova(fit_restricted, fit_full)` 
    - `car::linearHypothesis(fit_full, c("prelet = 0", "prerelat = 0", "peabody = 0"))`
    - `marginaleffects::hypotheses(fit_full, joint = c("prelet", "prerelat", "peabody"))`
    - Any of the three will work, but we recommend the third one because `marginaleffects` is a super useful package, so it is worth getting acquainted with it.
  b. A likelihood ratio test comparing the two models using `lmtest::lrtest()`.

2. Test whether the effect of `prenumb` on `postnumb` while controlling for `age` is significantly different for boys and girls. (Hint: it's an interaction)


### donations

Use the dataset `donations`. As before, treat the "miscellaneous" category of `religion` as missing. Construct `propdon`, the proportion of monthly income donated, and `percdon`, this proportion expressed as a percentage. Use the latter as outcome variable. Construct a dummy variable `old` for subjects older than 40.

3. Is there evidence that older people (over 40) donate a larger share of their income than younger people, controlling for religion?

4. Is there evidence that men donate a larger share of their income than women, controlling for religion? 

5. Using the model from the previous question, test whether religion as a whole contributes to explaining `percdon`, controlling for gender. Note that this amounts to testing the  hypothesis that all religion coefficients are zero. 


::: {.content-visible when-format="html"}

## Answers

::: {.callout-note appearance="simple" collapse=true}
## See answers


Make sure this file is in the same R Project as the `data` folder. Use relative file paths, not full paths to your computer.

```{r setup-model-testing}
library(tidyverse)
library(haven)
library(car)
library(lmtest)
library(modelsummary)
library(marginaleffects)
```

### sesam

```{r}
sesam <- read_dta("data/sesam.dta")
```

#### 1

Let's fit and have a look at the the restricted and full models.
```{r}
fit_restricted <- lm(postnumb ~ age + prenumb, data = sesam)
fit_full <- lm(postnumb ~ age + prenumb + prelet + prerelat + peabody, data = sesam)

modelsummary(list("Restricted" = fit_restricted, "Full" = fit_full), stars = TRUE)
```

The increase in R-squared from the restricted to the full model is small. Let's formally test it:

```{r}
# a. Wald test using an F-statistic. All the same results:
anova(fit_restricted, fit_full)
linearHypothesis(fit_full, c("prelet = 0", "prerelat = 0", "peabody = 0")) 
hypotheses(fit_full, joint = c("prelet", "prerelat", "peabody")) 

# b. Likelihood ratio test
lrtest(fit_restricted, fit_full)
```

We conclude that the full model fits significantly better than the restricted model. The joint null hypothesis that prelet, prerelat, and peabody are all zero is rejected.

#### 2

We can of course informally compare the coefficients for prenumb in the two separate models for males and females, but this does not give us a formal test of whether the difference is significant. We should simply fit a model with an interaction between prenumb and sex, controlling for age. 
```{r}
# Creating a numeric female dummy just for cleaner labels in the regression table
# You should get the same results using, e.g., factor(sex)
sesam <- sesam |> 
           mutate(female = as.numeric(sex==2))

fit_male <- lm(postnumb ~ prenumb + age, data = sesam |> filter(sex == 1))
fit_female <- lm(postnumb ~ prenumb + age, data = sesam |> filter(sex == 2))
fit_interaction <- lm(postnumb ~ prenumb * female + age, data = sesam)

modelsummary(list("Males" = fit_male, "Females" = fit_female, "Interaction" = fit_interaction), stars = TRUE)
```

The interaction coefficient is not significant at the 5% level, so we conclude that there is no evidence that the effect of prenumb on postnumb differs between boys and girls.

::: {.callout-tip}
There is way of combining estimates from separate models and then formally comparing coefficients across models. It is called "seemingly unrelated estimation" (suest) and was originaly developed in a Stata package by a former SaSR03 teacher, Jeroen Weesie. (see [here](https://www.stata.com/products/stb/journals/stb52.pdf), p. 34). There is a [R package](https://www.trentonmize.com/software/suest_r/), which must be installed from Github, implementing this method. We are skiping suest in the course, but provide the relevant code below for illustration. Nte that it also uses function from the `marginaleffects` package the we previously loaded. 

```{r}
# The lines below are commented out for technical reasons, but you can run this code in your own R session if you want to try it out. 
# install.packages("remotes")
# remotes::install_github("tdmize/suest")
library(suest)

combined <- suest(fit_male, fit_female, model_names = c("Male", "Female"))
effects <- avg_comparisons(combined, variables = "prenumb")
effects

hypotheses(effects, hypothesis = difference ~ pairwise)
```

Note that the estimate (0.224) is the difference between the coefficients of prenumb estimated separately for females (0.897) and males (.674). The difference is deemed not significant at the 5% level. We conclude the same thing with this method as with the interaction coefficient shown above: the effect of prenumb on postnumb does not significantly differ between boys and girls.

Bonus: Also note that the difference between coefficients estimated with different models (0.224) is not identical to the the interaction coefficient in the pooled model (0.208), even though they are supposed to measure the same thing ("by how much the effect of prenumb differs between males and females"). Change the interaction model above to also include an interaction term between sex and age, see what happens with the sex*prenumb coefficient, and reflect 
:::


### donations

Prepare data:

```{r}
donations <- read_dta("data/donations.dta")

donations <- donations |>
mutate(
  religion = na_if(religion, 6),
  religion = factor(religion,
                    levels = 1:5,
                    labels = c("none", "catholic", "reformed",
                              "rereformed", "islam")),
  old = age > 40,
  propdon = (donation / 12) / inc,
  percdon = propdon * 100
)
```

#### 3

```{r}
fit_old <- lm(percdon ~ old + religion, data = donations)
summary(fit_old)
```

Older people (over 40) donate a larger share of their income than younger people. Controlling for religion, the regression coefficient for old is positive (about 0.82) and significant.

#### 4

```{r}
fit_male_don <- lm(percdon ~ male + religion, data = donations)
summary(fit_male_don)
```

Controlling for religion, men donate a larger share of their income than women: the coefficient for male is positive (about 0.32), but not significant at the 5% level. It is significant at about the 6.5% level. There is therefore no strong evidence that men donate significantly more than women; this result should be interpreted with caution.


#### 5

As in question 1, there are multiple ways of doing this. Here we will do a Wald test using the F-statistic using `marginaleffects::hypotheses()`. Note that we need to use the coefficient names as they appear in the model summary.
```{r}
hypotheses(fit_male_don, joint = c("religioncatholic", "religionreformed", "religionrereformed", "religionislam"))
```

The F-statistic is significant, which means that the fit improves between a model without religion and a model with religion. We conclude that religion as a whole does significantly contribute to explaining `percdon`, controlling for gender.

:::
:::