---
title: Multiple linear regression
---

## Assignments

### Predicting a post-test score

1. Use the dataset `sesam`. We predict `postnumb` (a post-test score) from `age` and `prenumb` (a pre-test score).

    - Fit a regression of `postnumb` on `age` and `prenumb`. Interpret each coefficient, the intercept, and the $R^2$. Is `age` a significant predictor once `prenumb` is controlled for?

    - Fit a regression of `postnumb` on `prenumb` alone. Compare the coefficient for `prenumb` and the $R^2$ to the model that also includes `age`. What do you conclude?

### Donations, religion, and income

Use the dataset `donations`, which contains information on voluntary donations.. Treat the "miscellaneous" category of `religion` as missing, and exclude the corresponding observations from the analyses below (it is too heterogeneous a group to interpret meaningfully).

2. Test whether average donations differ between religious groups, controlling for income (`inc`). Fit a regression of `donation` on `religion` (as a factor) and `inc`. Which categories differ significantly from the reference category?

3. Is there a difference in average donation between two specific religious categories that are *not* both compared to the reference category in the previous output? For example, is there a significant difference between rereformed and reformed?
  - One way of testing this is re-running the model with a different reference category. A better way is to use the function `avg_comparisons()` from package `marginaleffects`, which you have to install and load. You can obtain pairwise comparisons between all levels of a categorical (i.e. factor) variable with `avg_comparisons(your_model_object, variables = list(your_variable_name =  "pairwise"))`.

4. Donations and income have been recorded in thousands. Look at the variable labels (`glimpse(data_frame_name$variable_name)`) to check in which time units they are recorded. Construct a new variable `propdon`, the proportion of monthly income donated. This will be outcome variable for the next questions. Report the mean and standard deviation of `propdon`. What do these values tell you?

5. Fit a regression of `propdon` on `religion` and obtain pairwise comparisons between all categories. Between which categories is there evidence of a difference?

6. Consider the following hypothesis: donations increase with income when controlling for 
religion, i.e., people gives a higher fraction of their income, the higher their income (in 
theoretical terms, charity is a "luxury good"). Fit a regression model to test this and interpret the results. 

7. Fit a regression of `propdon` on `male`, `religion`, and their interaction. Is there evidence that the effect of religion on donations differs between men and women?

8. Fit a regression of `percdon` on `inc`, `religion`, and their interaction. Is there evidence that the effect of income on donations varies between religious denominations?

9. Consider whether life events (`lifevent`) have a larger effect on charity among older people than among younger people. Beware: this formulation does not mean that we want you to 
dichotomize age. It is just an informal way of expressing that the effect of life events 
increases with age. Is there evidence that the effect of `lifevent` on `propdon` is larger for older people than for younger people?

10. Test whether the effect of income on `propdon` varies with age. (For young people a smaller share of their income is discretionary due to mortgage, children, etc.)

11. Test whether the effect of income on `propdon` varies by gender. 

12. Test whether the gender difference in income effects varies with religion (Note: this 
is actually a 3-way interaction).  

### Occupational prestige and education

Use the dataset `gss91_extended`. @e want to investigate whether the effect of education (`educ`) on occupational prestige (`prestg80`) is nonlinear.

13. Make a scatterplot of `prestg80` against `educ`. Overlay it with both a linear and a quadratic fitted line. 
  - Tipo: for the quadratic line, use the formula `y ~ x + I(x*x)` in `geom_smooth()`.  

14. Fit a regression model of `prestg80` on `educ`, a squared term for `educ`, and the control variables `race`, `region`, and `sex`. Make sure that the control variables are interpreted as categorical variables, not numeric. Is the squared term statistically significant? What do you conclude about the shape of the relationship?

15. Using the coefficients for `educ` and `educ`$^2$, compute the turning point of the curve: the value of `educ` at which expected prestige is at its minimum or maximum. Is the curve U-shaped or inverted-U shaped? Check how many observations in the data actually fall at or below this turning point, and discuss whether the turning point should be interpreted, or whether it is more appropriate to describe the relationship only over the range where most of the data lie.


::: {.content-visible when-format="html"}

## Answers

::: {.callout-note appearance="simple" collapse=true}
## See answers



```{r setup-multiple-regression}
library(tidyverse)
library(haven)
library(janitor)
library(modelsummary)
library(marginaleffects)
library(skimr)
```

# Predicting a post-test score
```{r}
sesam <- read_dta("data/sesam.dta")

fit_multi <- lm(postnumb ~ age + prenumb, data = sesam)
summary(fit_multi)

# E(postnumb) = 7.42 + 0.12 age + 0.78 prenumb      (SE_age = 0.11, SE_prenumb = 0.06)
```

Age does not contribute in predicting postnumb (not significant). The #marginal correlation between age and postnumb is significant, but here the effect of age is conditional on that of prenumb: once you control for prenumb, age does not matter. If pretest score goes up by 1, posttest score significantly goes up by 0.78, irrespective of age. People aged 0 with a pretest score of 0 have a posttest score of 7.42; interpreting the constant is not meaningful here.
About 45% of the variation in the post-test is explained by the pre-test and age.

```{r}
fit_prenumb_only <- lm(postnumb ~ prenumb, data = sesam)

modelsummary(list(fit_multi, fit_prenumb_only), stars = TRUE)
```

Very little has changed compared to the model with both predictors: the coefficient for prenumb and R-squared are almost the same with or without age in the model. 

### Donations, religion, and income

#### 2. Donations by religion, controlling for income

```{r}
donations <- read_dta("data/donations.dta")

donations |>  
  tabyl(religion)

print_labels(donations$religion)
```

Let's convert the `misc` values to missing, then convert the variable to factor. 
```{r}
donations <- donations |>
            mutate(religion = na_if(religion, 6),
                   religion = factor(religion,
                                     levels = 1:5,
                                     labels = c("none", "catholic", "reformed",
                                                "rereformed", "islam"))) 
```

```{r}
fit_rel <- lm(donation ~ religion + inc, data = donations)
summary(fit_rel)
```

There are differences in the average donations of the various religious groups. The null hypothesis that R-squared = 0 (p-value of the F-test) is rejected. From the coefficients we learn that there are significant differences in average donations between catholics and the reference
category (none), between reformed and none, and between rereformed and none. There is no evidence of a difference in average donation between islam and none.


#### 3. Pairwise comparisons

```{r}
avg_comparisons(fit_rel, variables = list(religion = "pairwise"))
```

There is a significant difference in donations between rereformed and reformed.  

#### 4. Proportion of income donated

```{r}
glimpse(donations$donation)
glimpse(donations$inc)
```
Donation is recorded per year, income is recorded per month. So we have to divide donation by 12 top get the monthly amount. As both are in thousands, it does not make a difference whether or not we multiply both by 1000. 
 
```{r}
donations <- donations |>
  mutate(
    propdon = (donation / 12) / inc,
  )

donations |>
  skim(propdon)
```

The mean proportion of monthly income donated is about 0.02. The sd is relatively large compared to the mean, which suggests substantial variation in the proportion of income that is donated.


#### 5. Proportion donated by religion

```{r}
fit_propdon <- lm(propdon ~ religion, data = donations)
summary(fit_propdon)

avg_comparisons(fit_propdon, variables = list(religion = "pairwise"))
```

No significant differences are found between islam and none, and between catholics and reformed (the 95% confidence interval for these pairwise differences includes 0). There is evidence of difference for all other comparisons. 

#### 6. Percentage donated and income

```{r}
fit_propdon_inc_rel <- lm(propdon ~ inc + religion, data = donations)
summary(fit_propdon_inc_rel)

```


There is evidence that donations increase with income, even controlling for religion (b = 0.00424, t = 5.49, p < 0.001). For each unit (1000) increase in monthly income, the proportion of income donated increases by about 0.004 (or the percentage increases by 0.4 percentage points).


#### 7. Does the religion effect differ by gender?

```{r}
fit_male_rel <- lm(propdon ~ male * religion, data = donations)
summary(fit_male_rel)
```

There is no evidence that the effect of religion on donations differs for men and women (the interaction terms are not significant).

#### 8. Does the income effect differ by religion?

```{r}
fit_inc_rel <- lm(propdon ~ inc * religion, data = donations)
summary(fit_inc_rel)
```

There is no evidence that the effect of income on donations varies between the religious denominations.


#### 9. Age and life events

```{r}
fit_age_life <- lm(propdon ~ age*lifevent, data = donations)
summary(fit_age_life)
```

There is no evidence that the effect of a life event on donations is larger for older than for younger people.


#### 10. Does the income effect vary with age?

```{r}
fit_age_inc <- lm(propdon ~ age * inc, data = donations)
summary(fit_age_inc)
```

There is no evidence that the effect of monthly income on donations varies with age.

#### 11. Does the income effect differ between men and women?

```{r}
fit_male_inc <- lm(propdon ~ male * inc, data = donations)
summary(fit_male_inc)
```

There is no evidence that the effect of monthly income on donations differs for men and women.

#### 12. A three-way interaction

```{r}
fit_inc_male_religion <- lm(propdon ~ male * inc * religion, data = donations)
summary(fit_inc_male_religion)

```

There is no evidence that possible gender differences in the effect of monthly income on donations vary with religion (the three-way interaction terms are not significant).


### Occupational prestige and education

```{r}
gss91 <- read_dta("data/gss91_extended.dta")
```
#### 13. Linear and quadratic fits

```{r}
ggplot(gss91, aes(x = educ, y = prestg80)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "blue") +
  geom_smooth(method = "lm", formula = y ~ x + I(x*x), se = FALSE, color = "red")
```

#### 14. Quadratic model with controls

To fit a quadratic line  we can either create the squared term beforehand (i.e. create `educ2 = educ*educ`, then include both `educ` and `educ2` in the equation) or specify the squared term with the syntax `I(educ*educ)`. (Note that simply including `educ*educ`, as we would do for an interaction of different variables, does not work. Thus the need for `I()`). 

```{r}
fit_quad <- lm(
  prestg80 ~ educ + I(educ^2) + factor(race) + factor(region) + factor(sex),
  data = gss91
)
summary(fit_quad)
```

The educ-squared term is significant, so there is evidence of a quadratic relationship, controlling for race, region, and sex.


#### 15. Turning point


```{r}
b_educ <- coef(fit_quad)["educ"]
b_educ2 <- coef(fit_quad)["I(educ^2)"]

turning_point <- -b_educ / (2 * b_educ2)
turning_point
```

Since the coefficient for the squared term is greater than zero, the relation is U-shaped. This was already suggested by the scatterplot (though that plot did not include statistical controls). We conclude that around 4 years of education corresponds to the lowest predicted prestige.

```{r}
gss91 |>
  filter(educ <= 4)
```

Only a small number of subjects have educ values at or below the turning point (about 12 out of roughly 1500). It is highly unlikely that prestige is actually higher with 1 or 2 years of education than# with 4 years; the model most plausibly applies for 4 and more years of education. We therefore abstain from interpreting the turning point itself, and conclude instead that, over the range where most of the data lie, prestige increases in education at an increasing rate.

:::
:::