---
title: Ordinal logistic regression
---

## Assignments

Use the dataset `hin95_voting`.

### Fitting the model and interpreting coefficients

1. Construct `age` (from `birthdate`), a dummy `female`, and a recoded church-attendance frequency `chfreq` (visits per year). Fit an ordered logistic regression of `vote` (voting intention, an ordinal variable) on `education`, `age`, `female`, and `chfreq`. Interpret the coefficient for `chfreq` as an effect on the latent voting-intention variable, and as a change in log odds.

2. Exponentiate the coefficient for `chfreq` to obtain an odds ratio. Interpret this odds ratio.

3. Compute the average marginal effects of all predictors. Interpret the marginal effect of `chfreq` on each of the four response categories of `vote`. Why is it not surprising that some of these marginal effects are positive and others negative?

### Predicted probabilities

4. Obtain predicted probabilities for each of the four response categories, for every subject. Verify, for one particular subject, that these probabilities sum to 1 (within rounding error).

### Testing a hypothesis on religion

5. Fit a restricted ordinal model without `chfreq`, and compare it to the full model with `chfreq` using a likelihood ratio test. What happens if you attempt this directly, without first checking the two models are based on the same sample? Explain why this happens, and how to fix it. After fixing the sample issue, what do you conclude about the relationship between church attendance frequency and voting intention?

### Checking linearity

6. Fit three models: (a) `chfreq` entered linearly, (b) `chfreq` entered as a fully flexible categorical predictor, and (c) `chfreq` entered with both a linear and a quadratic term. Compare (a) to (b), and (a) to (c), using likelihood ratio tests. Is there evidence that the linear specification is inadequate?

### Interactions

7. Fit a model with an interaction between `female` and `chfreq` (in addition to the main effects of `education`, `age`, `female`, and `chfreq`). Is there evidence that the effect of church attendance on voting intention differs between men and women? Report the implied effect of `chfreq` separately for men and for women.

8. Confirm your conclusion from the previous question with a likelihood ratio test comparing the model with and without the interaction term.

### Explained variance

9. Compute a pseudo-$R^2$ for the latent variable underlying `vote`, using the variance of the fitted linear predictor and the fixed error variance of the ordered logit model ($\pi^2/3$).

### Proportional odds assumption

10. Test the proportional odds (parallel lines) assumption for the model of `vote` on `education`, `age`, `female`, and `chfreq`. Is the assumption supported by the data? What would this mean for the interpretation of the model if the assumption were rejected?

## Answers

Make sure this file is in the same R Project as the `data` folder. Use relative file paths, not full paths to your computer.

```{r setup-ordinal}
library(tidyverse)
library(haven)
library(MASS)
library(lmtest)
library(marginaleffects)
library(brant)

hin95_voting <- read_dta("data/hin95_voting.dta")

hin95_voting <- hin95_voting |>
  mutate(
    age = 95 - birthdate %% 100,
    female = as.integer(sex == 2),
    chfreq = case_match(
      church,
      1 ~ 0,
      2 ~ 1,
      3 ~ 12,
      4 ~ 52,
      .default = NA_real_
    ),
    vote = factor(vote, ordered = TRUE)
  )
```

### Fitting the model and interpreting coefficients

#### 1. The ordered logit model

```{r}
fit <- polr(vote ~ education + age + female + chfreq, data = hin95_voting, Hess = TRUE)
summary(fit)

# (Latent) voting intention increases with religiosity: for each
# additional church visit per year, the latent voting-intention variable
# increases by about 0.011 units, holding education, age, and gender
# constant.
#
# Equivalently: if church attendance frequency goes up by 1, the log
# odds of being in a higher (more likely to vote) response category
# rather than a lower one increase by 0.011, holding the other
# predictors constant, and this applies uniformly across all cutpoints
# under the proportional odds assumption.
```

#### 2. The odds ratio for `chfreq`

```{r}
exp(coef(fit))

# If church attendance frequency goes up by 1, the odds of being in a
# higher rather than a lower response category of vote increase by
# about 1.13%, holding education, age, and gender constant.
```

#### 3. Average marginal effects

```{r}
avg_slopes(fit)

# If church attendance frequency goes up by 1:
#   - the probability of "certainly voting" increases by about
#     0.2 percentage points
#   - the probability of "probably voting" decreases by about
#     0.1 percentage points
#   - the probability of "probably not voting" decreases by about
#     0.08 percentage points
#   - the probability of "certainly not voting" decreases by about
#     0.06 percentage points
#
# This pattern is not surprising: an increase in the latent
# voting-intention variable shifts probability mass away from the lower
# (less likely to vote) categories toward the higher (more likely to
# vote) category. Because probabilities across all categories must sum
# to 1, some marginal effects are necessarily negative while others are
# positive.
```

### Predicted probabilities

#### 4. Predicted probabilities per response category

```{r}
pred_probs <- predict(fit, type = "probs")

nrow(hin95_voting)
nrow(pred_probs)

# polr() drops every case with a missing value on one of the predictors,
# so there are fewer predictions than rows in the dataset. The row names
# of the prediction matrix tell us which rows the model actually used.

hin95_voting_pred <- hin95_voting |>
  slice(as.integer(rownames(pred_probs))) |>
  bind_cols(as_tibble(pred_probs) |> rename_with(~ paste0("pr", seq_along(.))))

hin95_voting_pred |>
  filter(irespnr == 1000) |>
  dplyr::select(starts_with("pr"))

hin95_voting_pred |>
  filter(irespnr == 1000) |>
  dplyr::select(starts_with("pr")) |>
  rowSums()

# The predicted probabilities sum to 1 for each subject (within
# rounding error). Only cases without missing values on any of the
# predictors receive a prediction.
```

### Testing a hypothesis on religion

#### 5. A likelihood ratio test on matching samples

```{r}
fit_a <- polr(vote ~ education + age + female + chfreq, data = hin95_voting, Hess = TRUE)
fit_b <- polr(vote ~ education + age + female, data = hin95_voting, Hess = TRUE)

# This comparison is not valid:
# tryCatch(lrtest(fit_a, fit_b), error = function(e) e)

# lrtest refuses (or gives a misleading result) if attempted directly,
# because the two models are fit to different samples: fit_a excludes
# cases with a missing value on chfreq, while fit_b does not. A valid
# likelihood ratio test requires both models to be fit to exactly the
# same set of observations.

hin95_voting_complete <- hin95_voting |>
  filter(!is.na(chfreq))

fit_b_matched <- polr(vote ~ education + age + female, data = hin95_voting_complete, Hess = TRUE)
fit_a_matched <- polr(vote ~ education + age + female + chfreq, data = hin95_voting_complete, Hess = TRUE)

lrtest(fit_a_matched, fit_b_matched)

# After matching the samples, we conclude that the higher the
# religiosity (church attendance frequency), the higher the voting
# intention (LR chi2(1) = 23.27, p < 0.0001).
```

### Checking linearity

#### 6. Linear, flexible, and quadratic specifications

```{r}
fit_linear <- polr(vote ~ education + age + female + chfreq,
                    data = hin95_voting_complete, Hess = TRUE)

fit_flexible <- polr(vote ~ education + age + female + factor(chfreq),
                      data = hin95_voting_complete, Hess = TRUE)

fit_quadratic <- polr(vote ~ education + age + female + chfreq + I(chfreq^2),
                       data = hin95_voting_complete, Hess = TRUE)

lrtest(fit_linear, fit_flexible)
lrtest(fit_linear, fit_quadratic)

# Neither the fully flexible specification nor the quadratic
# specification significantly improves the fit over the linear
# specification of chfreq, so treating chfreq as linear is an adequate
# simplification.
```

### Interactions

#### 7. An interaction between `female` and `chfreq`

```{r}
fit_interaction <- polr(vote ~ education + age + female * chfreq,
                         data = hin95_voting_complete, Hess = TRUE)
summary(fit_interaction)

# No support for an interaction between religiosity and gender
# (Wald z = -0.25, p = 0.801).
#
# Effect of chfreq on the latent vote variable:
#   = b_chfreq                          for males
#   = b_chfreq + b_female:chfreq        for females
#
# The effect of church attendance on voting intention is essentially the
# same for men and women.
```

```{r}
b <- coef(fit_interaction)

b["chfreq"]
b["chfreq"] + b["female:chfreq"]
# the implied effect of chfreq for men and for women
```

#### 8. The same conclusion from a likelihood ratio test

```{r}
fit_no_interaction <- polr(vote ~ education + age + female + chfreq,
                            data = hin95_voting_complete, Hess = TRUE)

lrtest(fit_interaction, fit_no_interaction)

# The likelihood ratio test confirms the Wald test: there is no evidence
# that the effect of chfreq on vote differs between men and women.
```

### Explained variance

#### 9. A pseudo-$R^2$ for the latent variable

```{r}
fit_r2 <- polr(vote ~ education + age + female + chfreq,
               data = hin95_voting_complete, Hess = TRUE)

# The linear predictor (fitted latent variable, without the error term)
X <- model.matrix(fit_r2)[, -1]
lp <- as.vector(X %*% coef(fit_r2))

var_lp <- var(lp)
r2 <- var_lp / (var_lp + pi^2 / 3)
r2

# R2 = variance(model) / (variance(model) + variance(error)), where
# variance(error) = pi^2/3 = 3.29 by definition for the ordered logit
# model. This gives a pseudo-R-squared describing the proportion of
# variance explained in the latent (underlying continuous) voting
# intention variable.
```

### Proportional odds assumption

#### 10. The Brant test

```{r}
brant(fit_r2)

# The Brant test rejects the proportional odds (parallel lines)
# assumption for at least one of the predictors. If this assumption is
# rejected, the effect of that predictor is not the same across all
# cutpoints of vote, and the ordered logit coefficients should be
# interpreted with caution -- a model that relaxes the proportional odds
# assumption (e.g. a generalized/partial proportional odds model) may be
# more appropriate, though such extensions are beyond the scope of this
# course.
```
