---
title: Simple linear regression
---

## Assignments

### Anscombe's quartet

1. Use the dataset `anscombe`. This is a built-in dataset that you can load with `data(anscombe)` and contains four pairs of variables, (`x1`,`y1`), (`x2`,`y2`), (`x3`,`y3`), and (`x4`,`y4`).

    - Fit four separate bivariate regressions: `y1` on `x1`, `y2` on `x2`, `y3` on `x3`, and `y4` on `x4`. Compare the intercepts, slopes, standard errors, and $R^2$ values of the four models. What do you notice?
      - Tip: we recommend using  the function `modelsummary()`, from the package of the same name, to produce regression tables, especially when comparing models. It supports a wide range of models (including the more complex models you will learn later in the programme) and makes it easy to produce tables (almost) ready for publication. 

    - Make a scatterplot with a fitted regression line for each of the four pairs. Inspect the four plots side by side. 
      - Tip: You can of course make separate plots, but you can also restructure the dataset from wide to long, then use `facet_wrap()` or `facet_grid()` when building a ggplot. 

    - What is your conclusion? Why is it not sufficient to only report regression coefficients and $R^2$?

### Crime and poverty

Use the dataset `state_crime`. This dataset contains, for each U.S. state, the murder rate (`murder`) and a poverty score (`poverty`).

2. Make a scatterplot of `murder` against `poverty`, with state names (`state`) as labels. Which state stands out as an outlier?

3. Fit a regression of `murder` on `poverty` using all states. Report and interpret the intercept, the slope, and $R^2$.

4. Refit the regression excluding the outlying state, without removing the observation from the dataset (use a `filter()` inside the `data` argument, or subset only for this call). Compare the intercept, slope, and $R^2$ to the results in the previous question. What is the effect of this single observation on the results?

### Income and age

Use the dataset `sbs399`. Formulate a hypothesis about the relation between response variable hourly income (`hrinc`) and predictor variable `age`.

5. Make a scatterplot of `hrinc` against `age`. Does the relationship look linear? Are there any outliers? 

6. Fit a regression of `hrinc` on `age` (i.e. `hrinc` as the oucome and `age` as predictor). Interpret the intercept, the slope, and $R^2$.

7. Compute the mean of `hrinc` separately for men and women (e.g. using and `summarise()` with `.by`). What is the difference between the two means?

8. Fit a regression of `hrinc` on `sex`. Compare the regression coefficient for `sex` to the difference in means found in the previous question.


### Sesame street

The data `sesam` are part of a larger data set that evaluates the impact of the first year of the Sesame Street television series. Sesame Street (SS) is mainly concerned with teaching  preschool related skills to children in the 3–5-year age range. Use the following variables: 

- `sex` 
- `peabody`: vocabulary maturity before watching SS 
- `postnumb`:  knowledge of numbers after a year of SS 

9. Use `postnumb` as the response variable. Is the relation between `postnumb` and `peabody` linear? If not, think about a possible linearizing transformation of the response variable.

10. Fit a linear regression of `postnumb` on `peabody`. Report and interpret the intercept, slope, and $R^2$. Is the null hypothesis of no relation between `peabody` and `postnumb` rejected?

11. Fit a regression of  `postnumb` on `female`. Is there evidence that boys and girls differ, on average, on `postnumb`?

::: {.content-visible when-format="html"}

## Answers

::: {.callout-note appearance="simple" collapse=true}
## See answers

Make sure this file is in the same R Project as the `data` folder. Use relative file paths, not full paths to your computer.

```{r setup-regression}
library(tidyverse)
library(haven)
library(modelsummary)
```

### Anscombe's quartet

```{r}
data(anscombe)

fit1 <- lm(y1 ~ x1, data = anscombe)
fit2 <- lm(y2 ~ x2, data = anscombe)
fit3 <- lm(y3 ~ x3, data = anscombe)
fit4 <- lm(y4 ~ x4, data = anscombe)

modelsummary(list(fit1, fit2, fit3, fit4), stars = TRUE)

# Tip: modelsummary allows for a lot of customization
# If you want to show the p.value in the table, use:
# modelsummary(list(fit1, fit2, fit3, fit4), stars = TRUE, statistic = c("std.error", "p.value"))W
```

Coefficients, standard errors, and R-squared are (essentially) identical

Scatterplots with fitted regression line:

```{r}
# Solution 1: separate plots
p1 <- ggplot(anscombe, aes(x = x1, y = y1)) +
      geom_point() +
      geom_smooth(method = "lm")

p2 <- ggplot(anscombe, aes(x = x2, y = y2)) +
      geom_point() +
      geom_smooth(method = "lm")
# And so on for models 3 and 4...

# Solution 2: restructure, then use facets
anscombe |>
  pivot_longer(
    everything(),
    names_pattern = "(x|y)(\\d)",
    names_to = c(".value", "set")
  ) |> 
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~set)

```

The actual data patterns are quite different!


### Crime and poverty

```{r}
state_crime <- read_dta("data/state_crime.dta")
```

Spotting the outlier:

```{r}
ggplot(state_crime, 
      aes(x = poverty, 
          y = murder)) +
  geom_point()

ggplot(state_crime, 
        aes(x = poverty, 
            y = murder, 
            label = state)) +
  geom_text()
```

The outlier is Washington DC, with an exceptionally high murder rate.

Regression using all states: 

```{r}
fit_crime <- lm(murder ~ poverty, data = state_crime)
summary(fit_crime)

# E(murder) = -10.14 + 1.32 poverty       (SE = 0.28)

```

A 1 unit increase in the poverty scale yields a predicted increase of 1.32 units in the murder rate. R-squared = 0.32: 32% of the variance in murder rate is explained by the poverty rate. Note that the intercept is negative, which would mean a negative murder rate at a poverty score of 0. This of course does not make sense substantively, it is just a linear extrapolation to a value outside the range of the poverty score. 

Regression excluding the outlier:
```{r}
fit_crime_no_dc <- lm(murder ~ poverty, data = state_crime |> filter(state != "DC"))
summary(fit_crime_no_dc)

# The two models side by side
modelsummary(list(fit_crime, fit_crime_no_dc))

# E(murder) = -0.86 + 0.58 poverty        (SE = 0.10)
```

A 1 unit increase in the poverty scale now gives a predicted increase of only 0.58 units in the murder rate. R-squared = 0.40: 40% of the variance in murder rate is now explained by the poverty rate. 

### Income and age

```{r}
sbs399 <- read_dta("data/sbs399.dta")
```

Scatterplot of income against age. We expect a positive relation between age and hourly income.

```{r}
ggplot(sbs399, aes(x = age, y = hrinc)) +
  geom_point()
```

There are a number of observations with a score of 100 on the variable hrinc. Missings or hrinc cutted off at 100? We don't know. These observations may be outliers in Y-space. Is the relation linear? Maybe quadratic.


```{r}
fit_age <- lm(hrinc ~ age, data = sbs399)
summary(fit_age)
```

If age increases by 1 year, hourly income significantly rises by 0.29 EUR. Age explains 6.5% of the variance in income. The intercept of 8.02 is the predicted income at age 0, which is meaningless in itself.


Mean income for men and women:
```{r}
sbs399 |>
  summarise(mean_hrinc = mean(hrinc, na.rm = TRUE),
            .by = sex) 
```

Income is 3.4 EUR lower for women.

Regression of income on sex:
```{r}
fit_sex <- lm(hrinc ~ sex, data = sbs399)
summary(fit_sex)
```

The regression coefficient for sex is -3.41, which is the difference in expected hourly income between men and women as computed before. (Note that the output of the previous question was printed with fewer decimals, but R stores it with much higher precision. Fee free to explore how to check that the simple difference corresponds to the regression coefficient up to several decimals). 


### Sesame street

```{r}
sesam <- read_dta("data/sesam.dta")
```

Scatterplot:
```{r}
ggplot(sesam, aes(x = peabody, y = postnumb)) +
  geom_point()
```

The relation looks nonlinear: it is increasing, but decreasingly so. A quadratic transformation of peabody might help.

Regression:
```{r}
fit_peabody <- lm(postnumb ~ peabody, data = sesam)
summary(fit_peabody)

# E(postnumb) = 10.50 + 0.42 peabody       (SE = 0.04)
```

The null hypothesis of no relation between peabody and postnumb is rejected. Note: a significant slope does not mean the linear model is appropriate. It only means a straight line fits better than a horizontal one.

```{r}
fit_sex <- lm(postnumb ~ sex, data = sesam)
summary(fit_sex)
```

Difference is 0.02, not significant. There is no evidence that the average postnumb scores for boys and girls differ. 

:::
:::