---
title: Aggregating data
---

In this chapter you learn about the two related tasks:

- Create a new dataset with summary statistics such as group means.
- Add to a dataset new variables with aggregate statistics such as group means.

To illustrate, consider the following toy dataset with 5 pupils with respondent identifier `id`, from 2 school classes identified by the variable `class`, recording the father's education of the pupil in the variable `fedu`.

| id  | class | fedu |
|-----|-------|------|
| 1   | 1     | 3    |
| 2   | 1     | 4    |
| 21  | 2     | 2    |
| 22  | 2     | 3    |
| 23  | 2     | 3    |

The dataset with *class-level summary statistics* should contain 2 cases, one for each class, with the mean of `fedu` in the respective classes, new variable `mean_fedu`, and the number of observations, new variable `n`, in each class.

| class | n   | mean_fedu |
|-------|-----|-----------|
| 1     | 2   | 3.5       |
| 2     | 3   | 2.6667    |

The original individual-level dataset *enriched with class-level summary statistics* should read:

| id  | class | fedu | mean_fedu | n   |
|-----|-------|------|-----------|-----|
| 1   | 1     | 3    | 3.5       | 2   |
| 2   | 1     | 4    | 3.5       | 2   |
| 21  | 2     | 2    | 2.6667    | 3   |
| 22  | 2     | 3    | 2.6667    | 3   |
| 23  | 2     | 3    | 2.6667    | 3   |

Thus, the variable `mean_fedu` records contextual, class-level, information; the mean education of the parents in a pupil's class is treated as an individual attribute.

## R

In this part you need to study the following R functions:

| Function      | Description                                     |
|---------------|-------------------------------------------------|
| `summarise()` | Create a dataset with summary statistics        |
| `mutate()`    | Add new variables to the existing dataset       |
| `.by`         | Apply `summarise()` or `mutate()` within groups |
| `n()`         | Count the number of observations in a group     |

We present a short code fragment to accomplish the data manipulation described in the preamble.

Add the mean of `fedu` by class as a variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    mean_fedu = mean(fedu, na.rm = TRUE),
    .by = class
  )
```

Note that we could also have used the function `group_by()` to create a grouping structure, and then `mutate()` would have been applied within the groups. The `.by` argument is a convenient (and more recent) alternative that applies the operation separately within groups without permanently changing the grouping structure of the dataset. So the code below is a valid alternative to the code above:

```{r, eval=FALSE}
data <- data |> 
  group_by(class) |>
  mutate(
    mean_fedu = mean(fedu, na.rm = TRUE)
  ) |>
  ungroup()
```

(Read more about `group_by()` and `.by` [here](https://dplyr.tidyverse.org/reference/dplyr_by.html).)

Add variable `nclass`: number of observations in each class:

```{r, eval=FALSE}
data <- data |>
  mutate(
    nclass = n(),
    .by = class
  )
```

Create a N = 2 contextual dataset with the N and mean of `fedu`, grouped by class:

```{r, eval=FALSE}
class_data <- data |>
  summarise(
    n = n(),
    mean_fedu = mean(fedu, na.rm = TRUE),
    .by = class
  )
```

Make sure you understand the difference between `mutate()` and `summarise()`. The former *enriches the individual-level dataset* with class-level summary statistics, while the latter reduces the dataset to one row per group, thus creating a *new dataset with class-level summary statistics*.


## Assignments


### 1. People nested within neighborhoods within cities

We consider a simple synthetic dataset `ex_aggr.dta` of people nested within neighborhoods nested within cities.

- `city` city code
- `neighb` neighborhood code
- `employ` employment status, 1 = employed, 0 = unemployed
- `ethnic` 1 = native Dutch; 2 = western immigrant; 3 = nonwestern immigrant

a. Create a separate dataset recording, for each city, the number of people, the fraction of unemployed and the ethnic composition, i.e., the frac8tions of native Dutch, western immigrants, and nonwestern immigrants.
b. Add variables to the original dataset with the percentages of unemployed and the ethnic composition of the neighborhood, i.e., the fractions, or percentages, of Dutch, western, and nonwestern people in the neighborhood. Consider how to also add the information aggregated at the city level.
c. Add a variable representing the percentage of ethnic ingroup in the neighborhood.
d. What is the average percentage of ethnic ingroup in the neighborhood for native Dutch, western immigrants, and nonwestern immigrants? 

### 2. Education within couples

Use `hin95_ext.dta`. The dataset contains data on couples (married or cohabiting), as well as people living alone. The households are indicated by the identifier `hrespnr`. Keep only the couples, i.e., households with exactly two members. The variable `educ` records the level of education of the respondent.

a. Add a variable for the mean education of the household. Thus, this variable should have the same value for both partners.
b. Add a variable for the minimum education of the household. Add a variable for the maximum education of the household.
c. Add a variable recording the absolute difference of the levels of education of husband and wife.


::: {.content-visible when-format="html"}

## Answers

::: {.callout-note appearance="simple" collapse=true}
## See answers

```{r setup-aggregate}
library(tidyverse)
library(haven)

ex_aggr <- read_dta("data/ex_aggr.dta")
hin95_ext <- read_dta("data/hin95_ext.dta")
```

### 1. People nested within neighborhoods within cities

Create the city-level dataset.
```{r}
city_data <- ex_aggr |>
  summarise(
    n = n(),
    fr_unemployed = mean(employ == 0),
    fr_dutch      = mean(ethnic == 1),
    fr_western    = mean(ethnic == 2),
    fr_nonwestern = mean(ethnic == 3),
    .by = city
  ) 

city_data
# mean() of a logical condition gives the fraction for which it is TRUE,
# because TRUE counts as 1 and FALSE as 0
```

```{r}
city_data |>
  mutate(
    total = fr_dutch + fr_western + fr_nonwestern
  )
# the three ethnic fractions should sum to 1 in every city
```


Add neighborhood-level variables to the individual-level dataset.

```{r}
ex_aggr <- ex_aggr |>
  mutate(
    n_neighb = n(),
    pc_unemployed_neighb = 100 * mean(employ == 0),
    pc_dutch_neighb = 100 * mean(ethnic == 1),
    pc_western_neighb = 100 * mean(ethnic == 2),
    pc_nonwestern_neighb = 100 * mean(ethnic == 3),
    .by = c(city, neighb)
  )
# grouping by c(city, neighb) instead of just neighb is safe even if neighborhood codes repeat across cities
```

```{r}
ex_aggr |>
  arrange(city, neighb) |> 
  slice(1:20)
# the same neighborhood percentages are repeated for everyone in it
```

The city-level information is added in exactly the same way, only the grouping changes.

```{r}
ex_aggr <- ex_aggr |>
  mutate(
    n_city = n(),
    pc_unemployed_city = 100 * mean(employ == 0),
    pc_dutch_city = 100 * mean(ethnic == 1),
    pc_western_city = 100 * mean(ethnic == 2),
    pc_nonwestern_city = 100 * mean(ethnic == 3),
    .by = city
  )
# could also have used the city_data dataset to join the information, using left_join(), which we present in the next chapter
```


Add the percentage of the respondent's own ethnic group in the neighborhood.

```{r}
ex_aggr <- ex_aggr |>
  mutate(
    pc_ingroup = recode_values(ethnic,
                                1 ~ pc_dutch_neighb,
                                2 ~ pc_western_neighb,
                                3 ~ pc_nonwestern_neighb
                              )
        )
# Dutch people (ethnic == 1) get the percentage of Dutch in their neighborhood, and so on
# could of course also have used case_when()
```

Average percentage of ethnic ingroup in the neighborhood, by ethnic group.
```{r}
ex_aggr |>
  summarise(
    mean_pc_ingroup = mean(pc_ingroup),
    .by = ethnic
  )
```


### 2. Education within couples

Select only couples.
```{r}
hin95_ext_couples <- hin95_ext |>
                     filter(n() == 2, .by = hrespnr)
```

Add the household mean, minimum, and maximum of education.

```{r}
hin95_ext_couples <- hin95_ext_couples |>
                      mutate(
                        mean_educ = mean(educ, na.rm = TRUE),
                        min_educ = min(educ, na.rm = TRUE),
                        max_educ = max(educ, na.rm = TRUE),
                        .by = hrespnr
                      )
```

```{r}
# all three created variables take the same value for both partners in a household
hin95_ext_couples |>
  select(hrespnr, educ, mean_educ, min_educ, max_educ) |>
  arrange(hrespnr) |>
  slice(1:10)
```

Add the absolute difference in education between the partners.

```{r}
hin95_ext_couples <- hin95_ext_couples |>
                      mutate(
                        diff_educ = max_educ - min_educ
                      )
# no need for abs() here
```

```{r}
hin95_ext_couples |>
  select(hrespnr, educ, min_educ, max_educ, diff_educ) |>
  arrange(hrespnr) |>
  slice(1:10)
# diff_educ is zero for couples with equal education
```

Summary stats for created variables:

```{r}
hin95_ext_couples |>  
  skimr::skim(mean_educ, min_educ, max_educ, diff_educ)
```

:::
:::