---
title: Documenting data
---

In this short section you will learn about the following basic tasks:

- Rename variables and give them clear names.

- Use meaningful names for categorical values.

- Convert categorical variables to factors.

- Code missing values as `NA`.

- Change the order of variables in a dataset.

## R

Learn the following R commands to document the data, variables, and values:

| Function       | Description                        |
|----------------|------------------------------------|
| `rename()`     | Rename variables                   |
| `mutate()`     | Modify or create variables         |
| `factor()`     | Convert a variable to a factor     |
| `na_if()`      | Change a specific value to `NA`    |
| `replace_na()` | Replace `NA` with a specific value |
| `relocate()`   | Change the order of variables      |

Rename variables:

```{r, eval=FALSE}
data <- data |> 
  rename(educ = v101) 
# rename v101 to educ

data <- data |> 
  rename(sex = v1, edu = v2) 
# rename multiple variables
```

Convert a categorical variable to a factor:

```{r, eval=FALSE}
data <- data |>
  mutate(
    sex = factor(
      sex,
      levels = c(1, 2),
      labels = c("male", "female")
    )
  )
# assign meaningful labels to categories
```

Code missing values:

```{r, eval=FALSE}
data <- data |> 
  mutate(educ = na_if(educ, 99)) 
# change the value 99 to NA 

data <- data |> mutate(across(c(educ, feduc), ~ na_if(.x, 9))) 
# change the value 9 to NA in educ and feduc
```

Replace missing values with a numeric code:

```{r, eval=FALSE}
data <- data |> 
  mutate(educ = replace_na(educ, 99)) 
# change NA to 99
```

Change the order of variables:

```{r, eval=FALSE}
data <- data |> 
  relocate(hrespnr, sex, age, edu) 
# put these variables first 
```

Be careful when replacing missing values with numeric codes such as 99. In R, missing values should usually be coded as `NA`. Numeric missing-value codes can easily be treated as real values in analyses if they are not recoded properly.

## Assignments

See `dh_doc.Rmd` for solutions.

In these assignments we use the dataset `sbs399_unlabeled.dta`.

1.  Rename the variable `hrespnr` to `hid` and the variable `hrinc` to `hinc` in one command.

2.  The variable `sex` is coded numerically. Convert `sex` into a factor with meaningful labels. Use `male` and `female` as category labels.

3.  The variables `sexrol1` and `sexrol2` measure family attitudes. Convert both variables into factors with the following category labels:

    1.  fully disagree

    2.  slightly disagree

    3.  indifferent

    4.  slightly agree

    5.  fully agree