---
title: Creating a system file from raw data
---

In this part you learn about reading raw data into R. Raw data are often stored as text files that can be opened using a simple text editor. Typically the data are in the form of a data matrix in which the rows are cases, columns are variables, and values are separated by commas, tabs, or spaces.

In modern R workflows, raw data are usually read from files such as CSV, tab-delimited text files, Excel files, or Stata files. Occasionally, it is also useful to create a small dataset manually in R, for example for examples, tests, or teaching.

## R

In this part you need to learn and read about the following R functions:

| Function       | Description                         |
|----------------|-------------------------------------|
| `read_csv()`   | Read comma-separated text data      |
| `read_tsv()`   | Read tab-separated text data        |
| `read_table()` | Read whitespace-separated text data |
| `read_dta()`   | Read a Stata dataset                |
| `write_csv()`  | Save data as a CSV file             |
| `write_rds()`  | Save data as an R object            |
| `tribble()`    | Manually create a small dataset     |
| `View()`       | Inspect a dataset in RStudio        |

Examples:

Read a CSV file:

```{r, eval=FALSE}
library(readr)
data <- read_csv("data/mydata.csv")
```

Read a tab-delimited text file:

```{r, eval=FALSE}
data <- read_tsv("data/mydata.txt")
```

Read a whitespace-separated text file:

```{r, eval=FALSE}
data <- read_table("data/mydata.dat")
```

Read a Stata dataset:

```{r, eval=FALSE}
library(haven)
data <- read_dta("data/mydata.dta")
```

Create a small dataset manually:

```{r, eval=FALSE}
library(tibble)
data <- tribble(
  ~hrespnr, ~city,        ~female, ~edu,
  101,      "Amsterdam",  1,       3,
  103,      "Den Haag",   0,       5,
  107,      "Rotterdam",  0,       2,
  109,      "Utrecht",    1,       3
)
```

Inspect the dataset:

```{r, eval=FALSE}
View(data)
```

Save the dataset as a CSV file:

```{r, eval=FALSE}
write_csv(data, "data/mydata.csv")
```

Save the dataset as an R object:

```{r, eval=FALSE}
write_rds(data, "data/mydata.rds")
```

## Assignments

### 1. Manual entering data

We want to create an R dataset with the following variables:

| hrespnr | city      | female | edu |
|---------|-----------|--------|-----|
| 101     | Amsterdam | 1      | 3   |
| 103     | Den Haag  | 0      | 5   |
| 107     | Rotterdam | 0      | 2   |
| 109     | Utrecht   | 1      | 3   |

Do this in two ways.

First, create the dataset manually in R using `tribble()`.

Second, create the same dataset as a CSV file, save it in your `data/` folder, and read it into R using `read_csv()`.

Save both newly created datasets.

## Answers

Make sure this file is in the same R Project as the `data` folder. Use relative file paths, not full paths to your computer.

```{r setup-enterdata}
library(tidyverse)
```

### 1. Manual entering data

Enter the data by hand with `tribble()`.

```{r}
mydata <- tribble(
  ~hrespnr, ~city,        ~female, ~edu,
  101,      "Amsterdam",  1,       3,
  103,      "Den Haag",   0,       5,
  107,      "Rotterdam",  0,       2,
  109,      "Utrecht",    1,       3
)

mydata
# tribble() lays the data out row by row, so the code looks like the table
# the column names are preceded by a tilde
```

Check what R made of the variable types.

```{r}
glimpse(mydata)
# city is a character variable; the other three are numeric (dbl)
# note that hrespnr and female are identifiers and codes, not quantities,
# even though R stores them as numbers
```

Save this version.

```{r}
write_rds(mydata, "data/mydata_tribble.rds")
# .rds keeps the object exactly as it is, including variable types
```

Now build the same dataset as a CSV file. Writing the file from R is the reproducible way to do it; you could equally well type it in a text editor and save it as `data/mydata.csv`.

```{r}
csv_text <- "hrespnr,city,female,edu
101,Amsterdam,1,3
103,Den Haag,0,5
107,Rotterdam,0,2
109,Utrecht,1,3
"

write_file(csv_text, "data/mydata.csv")
# the first line holds the variable names, each following line is one case
# values are separated by commas, without spaces after them
```

Look at the file as plain text, the way a text editor would show it.

```{r}
read_file("data/mydata.csv") |>
  cat()
# a CSV file is just text: no formatting, no types, only characters
```

Read the file back into R.

```{r}
mydata_csv <- read_csv("data/mydata.csv")

mydata_csv
# read_csv() reports which type it guessed for each column
```

Verify that both routes produced the same dataset.

```{r}
all.equal(mydata, mydata_csv)
# TRUE means the two datasets are identical in content and types
```

Save the version read from CSV.

```{r}
write_rds(mydata_csv, "data/mydata_csv.rds")
```

A remark on types. `read_csv()` guesses the type of each column from its contents, whereas `tribble()` takes the types from what you typed. Here both give the same result, but a guess can go wrong: an identifier with leading zeros, such as a postcode `0234`, would be read as the number 234. You can override the guess.

```{r}
mydata_typed <- read_csv(
  "data/mydata.csv",
  col_types = cols(
    hrespnr = col_character(),
    city = col_character(),
    female = col_double(),
    edu = col_double()
  )
)

glimpse(mydata_typed)
# hrespnr is now text, which is what an identifier really is
```

Finally, a note on why `.rds` and `.csv` are both useful. A CSV file can be opened by any program and read by anyone, but it stores only text and loses the variable types. An `.rds` file preserves the R object exactly, but only R can read it.

```{r}
list.files("data", pattern = "^mydata")
# the files created in this assignment
```
