---
title: Using functions
---

In this part you will learn about the use of functions (such as "take the square root", "taking a substring of a string", or "draw a random number" — the computer version of "flipping coins" — to accomplish various tasks in constructing or modifying variables. Functions may be embedded in arithmetic expressions, and they may also be composed. For instance, we may take the logarithm of a random variate with a uniform distribution on \[0, 1\]. We present the information, mostly in the context of assignments, under a number of headers:

- Numeric functions

- String functions

- Other kind of functions

## Examples

Absolute difference `dage` of the age of husband (`hage`) and wife (`wage`):

```{r, eval=FALSE}
data <- data |>   
  mutate(dage = abs(wage - hage)) 
```

Equivalent, using `if_else()`:

```{r, eval=FALSE}
data <- data |>   
  mutate(dage = if_else(wage > hage, wage - hage, hage - wage)) 
```

The natural logarithm of `1 + wagerate`. Note that `log(x)` is missing for `x <= 0`.

```{r, eval=FALSE}
data <- data |>   
  mutate(lnwage = log(1 + wagerate)) 
```

The maximum wage rate of the husband (`hwage`) and wife (`wwage`):

```{r, eval=FALSE}
data <- data |>   
  mutate(maxwage = pmax(hwage, wwage)) 
# pmax stands for "parallel max". It compares the given columns for each row and
# returns the maximum value for *each row*. 
```

Income chopped at 100000:

```{r, eval=FALSE}
data <- data |>   
  mutate(rinc = pmin(inc, 100000)) 
```

Length, number of characters, in the string variable `cityname`:

```{r, eval=FALSE}
data <- data |>   
  mutate(nch = str_length(cityname)) 
# or nchar(cityname)
```

Display the number of full weeks in a year. The R Console can be used as an interactive calculator.

```{r, eval=FALSE}
floor(365 / 7) 
```

Embellish unrounded output.

```{r, eval=FALSE}
paste("number of full weeks in year =", round(365 / 7, 2)) 
```

### Numeric functions

1.  The following functions in R are important for the creation of new variables or the modification of existing variables. Make sure that you understand the functions. If you are uncertain, think of examples, and verify. In particular, check how to access the help pages for their definitions. Be sure to understand that these functions exist, roughly what they are good for, and how to find the details at the moment you really want to use them.

| R               | Description                     |
|-----------------|---------------------------------|
| `sqrt(x)`       | square-root of x                |
| `abs(x)`        | the absolute value of x         |
| `pmin(x, y)`    | the minimum of x and y          |
| `pmax(x, y, z)` | the maximum of x, y, and z      |
| `round(x)`      | x rounded to nearest integer    |
| `floor(x)`      | x rounded down                  |
| `ceiling(x)`    | x rounded up                    |
| `x %% y`        | remainder after dividing x by y |
| `exp(x)`        | the exponential function        |
| `log(x)`        | the natural logarithm           |
| `log10(x)`      | the logarithm to the base 10    |

Functions and other operations may be mixed. For instance, you may write:

```{.r}         
data <- data |>   
  mutate(lp = (log(p) / log(1 - p)) - 4) 
```

Functions may also be nested, or composed. For instance:

```{.r}         
data <- data |>   
  mutate(fx = log(1 + exp(x))) 
```

Use parentheses if you are unsure about the priority of operators. What comes first, division or subtraction? Many inexperienced users will not remember (it is division). We recommend using parentheses rather than relying on your memory.

For each of these functions, what do you think will happen if the argument, or any of the arguments, is missing? Under what other conditions can each of these functions return a missing value? Experiment, or read the help pages.

A warning about `round()`: R rounds halves to the nearest even number, so `round(2.5)` is 2 and `round(3.5)` is 4. See: 
```{r}
round(c(0.5, 1.5, 2.5, 3.5, 4.5))
```


2.  Check the meaning of the following mathematical R functions.

| R                        | Description                                 |
|--------------------------|---------------------------------------------|
| `x %in% c(a, b, c, ...)` | logical: x equals a, or b, or c, ...        |
| `between(x, a, b)`       | logical: x ≥ a and x ≤ b                    |
| `if_else(x, y, z)`       | if x, then y else z                         |
| `case_when()`            | recode values using multiple conditions     |
| `cut(x, breaks)`         | recode a continuous variable into intervals |

### String functions

In R, string manipulation is commonly performed using functions from the `stringr` package, which is part of the `tidyverse`. Make sure that you understand the purpose of the following functions. If you are uncertain, think of examples and verify them.

| R                  | Description                                  |
|--------------------|----------------------------------------------|
| `str_length(s)`    | length of string                             |
| `str_to_lower(s)`  | convert string `s` to lowercase              |
| `str_to_upper(s)`  | convert string `s` to uppercase              |
| `str_detect(s, t)` | test whether string `t` occurs in string `s` |
| `str_locate(s, t)` | position of string `t` in string `s`         |
| `str_sub(s, n, m)` | substring of `s` from position `n` to `m`    |
| `str_trim(s)`      | drops leading and trailing blanks from `s`   |
| `as.numeric(s)`    | convert string `s` to a number               |
| `as.character(x)`  | convert object `x` to a string               |


As a more demanding task, split a string variable `str` into the first word (`w1`) and the remainder of the string (`rest`), where the first word is defined as all characters before first space.

```{.r}          
data <- data |>   
  mutate(     
    first_space = str_locate(str, " ")[, 1],     
    w1   = str_sub(str, end = first_space-1),     
    rest = str_sub(str, start = first_space)  
  ) 
```

Does this code yield correct results if `str` has leading spaces? What if it has no spaces? Can you fix it? Can you think of a way to extract the second word? 

For more complex operations with strings, it is useful to know the basics of regular expressions, which we do not cover in this course. 

The functions `as.numeric()` and `as.character()` convert between strings and numeric values. When `as.numeric()` cannot interpret a string as a number, it returns `NA`.

In the assignments below you will apply string functions to manipulate dates and name-address information.

### Other functions in R

There are plenty of other types of R functions. We do not expect you to remember all details of these or any of the other functions you encounter here, just that you get an understanding of the kinds of tasks that may be accomplished with functions, and how to access information about them when needed.

####  Random number generation

- `runif(1)` draws a random number from the uniform distribution (by default, between 0 and 1). Each time you invoke `runif()`, the function returns a different value. 

- `sample(1:5, 1)` returns 1, 2, ..., 5 each with the same probability.

- `rbinom(1, 1, 0.75)` generates a 1 with probability 0.75, and a 0 with probability 0.25. `rbinom(1, 1, 0.5)` generates 0 and 1 with equal probability. (Check by generating a large number of random numbers — change argument `n` — and checking how many 1s and 0s were produced)

- `rnorm(1, 1.3, 2.5)` generates a random number from a normal distribution with mean = 1.3 and standard deviation = 2.5.

For instance, to randomly select each observation in the data with probability 0.5:

```{.r}         
data |>   
  mutate(selected = rbinom(n(), 1, 0.5)) |>
  filter(selected == 1) 
```

See `?Distributions` for additional random number generators.

#### Missing value handling

The function `is.na()` returns `TRUE` if a value is missing and `FALSE` otherwise.

For instance, to select all observations that are not missing on the variables `age`, `medu`, and `gender`:

```{.r}         
data |>   
  mutate(hasmiss = is.na(age) | is.na(medu) | is.na(gender)) |> 
  filter(!hasmiss) 

# Or equivalently 
data |>   
  mutate(notmiss = !is.na(age) & !is.na(medu) & !is.na(gender)) |> 
  filter(notmiss) 

# As always with R, other solutions are possible
```

#### Miscellaneous


- `row_number()` returns the observation sequence number, that is, 1 in the first observation, 2 in the second observation, etc. One common application is to generate a case identifier:

```{.r}          
data <- data |>   
  mutate(respnr = row_number()) 
```

- `n()` returns the number of observations. Within `.by` or `group_by()`, it returns the number of observations in the group.

To create a variable with number of respondents in each household, indexed by `hrespnr`:

```{.r}          
data |>   
  mutate(n_residents = n(), .by = hrespnr)
```

## Assignments


1.  Use `sbs399.dta`. We focus on the variable `hrinc` (hourly income).
  - What is the value of the mean and median of the untransformed variable?

  - For each of the following transformations, find the mean and the median: `exp(hrinc)`, `hrinc*hrinc`, `sqrt(hrinc)`, `log(hrinc)`, `hrinc` (i.e., untransformed, identity), `1/sqrt(hrinc)`, `1/hrinc`, and `1/(hrinc*hrinc)`.

  - For which transformation, including the no-transformation, is the variable most symmetrically distributed? You may gauge the symmetry of a distribution by its skewness. Recall: a positive skewness indicates a distribution tilted to the left, a negative skewness indicates a distribution tilted to the right. Compute the skewness using the function `skewness()` from the package `parameters`.


2.  We focus on the variable `hrinc` in the original `sbs399` data again. The values are decimal numbers such as 3.5. Generate transformations of this variable using functions:
  - round `hrinc` to the nearest integer (so any value between 3.5 and 4.5 should be rounded to 4);

  - round the variable down (so any value between 3 and 4 is rounded to 3);

  - round the variable up (so any value between 3 and 4 is rounded to 4);

  - round in multiples of 10 (so any value between 5 and 15 is rounded to 10);

  - clip to the range \[10,50\], i.e., values below 10 are set to 10; values above 50 are set to 50; values between 10 and 50 are returned unchanged. Check the range (min and max) of the clipped variable. 

  - Inspect the first 10 values of the original variable and the above transformations. 

3.  The dataset `sbs399` contains the educational level of the father and mother of respondents (`feduc` and `meduc` respectively). We define the educational resources of the parents (`peduc`) as the maximum of the educational levels of the parents. Give code to create `peduc`. Do you think that this conceptualization is also appropriate if education is seen as a proxy for attitudes?

4. Now use the `home_work` dataset, which includes 3 cases only. You want to compute the geographic distance `d(h,w)` between a subject's home address and a subject's work address. The data contain the geographic coordinates `(home_x, home_y)` and `(work_x, work_y)` of the home and work location, expressed as east-west and north-south distances from a reference point in the Netherlands, in units of 1 km, rounded. Give R code that computes the distance `d` between home and work, assuming a flat earth. Recall that by Pythagoras' theorem

$$d(h,w) = \sqrt{(home_x - work_x)^2 + (home_y - work_y)^2}$$

  (Did you dream of ever using this in the social sciences?) Round the distance in meaningful units, as locations are rounded in 1 km units, it does not make sense that distance is expressed in a finer unit, say meters.[^1]


5.  Use the dataset `hin95_birthdates`.
  - This dataset contains the respondent's birth date `rbirth`, a numeric variable recording the birth dates in the format `ddmmyy`. Note that the number has 5 digits if `dd < 10`, and 6 digits otherwise. Construct three variables `d_rbirth`, `m_rbirth`, and `y_rbirth` with the day, month, and year in which the respondent was born. Use the functions `%%` and `floor()`.

  - List the first twenty observations to verify your variable construction.

  - To check your work, you may actually reconstruct the original coding from the three variables, and verify that the reconstructed variable matches the original variable.

6.  The file `hin95_birthdates` contains the respondent's birth date as a string variable `rbirth_str` in the format `"ddmmyy"`. The string is of length 5 if `dd < 10`, and of length 6 otherwise. Use string functions to construct three variables with the day, month, and year in which the respondent was born.

7.  The dataset `hin95_naw` contains name, address, and city information. This is a random selection of address information that we purchased from the Dutch mail agency for the "Households in the Netherlands in 1995" project. In this assignment we do some manipulations of this dataset with address information.
  - The string variable `postcode` contains postcodes in the Dutch format `1234 AB`. Construct a new numeric variable `npcode` that contains the numeric part. Thus, from `1234 AB` we want to extract the part `1234`. This is useful to merge information that is being organized with respect to geographic information, e.g., for contextual analyses. First take the substring of the first initial characters, i.e., convert `1234 AB` to a string `1234`. Then convert this string to a number, thus the string `1234` to the number `1234`. Be sure to understand the difference. What is the mean value of `npcode`? How do you interpret this number?

[^1]: For the interested reader: How do you obtain such location information yourself? If the dataset contains address information or postal codes, it is possible to use geocoding tools to obtain geolocations. Such tools can add new variables with cleaned address information and geographic coordinates. Using these geolocations you may compute, as explained above, the distance as the crow flies assuming a flat earth. If you pedantically want to account for the curvature of the earth, R also has packages that compute distances using geographic coordinates on the surface of the earth. More advanced students can even take into account that due to its spinning, the earth has to a high degree of accuracy an ellipsoid shape, not a ball shape. Then again, for a small country as the Netherlands such fine points are fully immaterial. Additional tools and packages provide access to other map services, such as the travel distance and travel time between two locations, distinguishing travel by car, by public transportation, by bike, and walking. One can also add variables with facilities, e.g., the number of schools within a 2 km radius around a location.

::: {.content-visible when-format="html"}

## Answers

::: {.callout-note appearance="simple" collapse=true}
## See answers


Make sure this file is in the same R Project as the `data` folder. Use relative file paths, not full paths to your computer.

```{r setup-functions}
library(tidyverse)
library(haven)
library(skimr)
library(parameters)

sbs399 <- read_dta("data/sbs399.dta")
```

### 1. Transformations of hourly income

The mean and the median of the untransformed variable.

```{r}
sbs399 |>
  skim(hrinc)
# mean = 18.2
# median = 16.1
```

Create all eight transformations:

```{r}
transformed <- sbs399 |>
  transmute(
    identity = hrinc,
    exponential = exp(hrinc),
    square = hrinc * hrinc,
    square_root = sqrt(hrinc),
    logarithm = log(hrinc),
    inverse_root = 1 / sqrt(hrinc),
    inverse = 1 / hrinc,
    inverse_square = 1 / (hrinc * hrinc)
  )
# transmute() keeps only the newly created variables. 
# Equivalent to cerating the variables with mutate(), then selecting only them
# with select()
```

Compute the mean, median, and skewness for each transformation.

```{r}
transformed |>
  skim()

transformed |>
  skewness() 
```



### 2. Rounding the hourly wage rate

```{r}
sbs399 <- sbs399 |>
  mutate(
    hrinc_round = round(hrinc),
    hrinc_down = floor(hrinc),
    hrinc_up = ceiling(hrinc),
    hrinc_tens = round(hrinc / 10) * 10,
    hrinc_clip = pmin(pmax(hrinc, 10), 50))
# round()   rounds to the nearest integer
# floor()    rounds down
# ceiling()  rounds up
# dividing by 10, rounding, and multiplying by 10 gives multiples of 10
# pmin(pmax(x, 10), 50) clips the values to the interval [10, 50]

# Check range
sbs399 |> 
  skim(hrinc_clip)

# List first 10 observations
sbs399 |> 
  select(starts_with("hrinc")) |> 
  slice_head(n = 10)
```



### 3. Educational resources of the parents

```{r}
sbs399 <- sbs399 |>
  mutate(peduc = pmax(feduc, meduc))
# pmax() compares the two variables element by element, respondent by respondent
# max(feduc, meduc) would return one single number for the whole dataset
```

Check the new variable.

```{r}
sbs399 |>
  select(feduc, meduc, peduc) |>
  slice(1:10)
# peduc should equal the larger of the two parental education levels
```

### 4. Distance between home and work

```{r}
home_work <- read_dta("data/home_work.dta")
```

```{r}
home_work |>
  mutate(
    d = sqrt((home_x - work_x)^2 + (home_y - work_y)^2),
    d_km = round(d)
  )
# Pythagoras' theorem on a flat earth
# the coordinates are rounded to whole kilometres, so reporting d to the
# nearest kilometre is the finest unit that is still meaningful
```



### 5. Splitting a numeric birth date

```{r}
hin95_birthdates <- read_dta("data/hin95_birthdates.dta")
```

```{r}
hin95_birthdates <- hin95_birthdates |>
  mutate(
    d_rbirth = floor(rbirth / 10000),
    m_rbirth = floor((rbirth %% 10000) / 100),
    y_rbirth = rbirth %% 100
  )
# rbirth has the form ddmmyy
# dividing by 10000 and rounding down leaves the day
# the remainder after dividing by 10000 is mmyy; dividing that by 100 leaves the month
# the remainder after dividing by 100 is the year
```

List the first twenty observations.

```{r}
hin95_birthdates |>
  select(rbirth, d_rbirth, m_rbirth, y_rbirth) |>
  slice(1:20)
# verify that the three components match the original number
```

Reconstruct the original variable and compare.

```{r}
hin95_birthdates |>
  mutate(
    check = d_rbirth * 10000 + m_rbirth * 100 + y_rbirth
  ) |>
  summarise(
    n_different = sum(check != rbirth, na.rm = TRUE)
  )
# the number of mismatches should be zero
```

### 6. Splitting a string birth date

```{r}
hin95_birthdates <- hin95_birthdates |>
  mutate(
    d_str = str_sub(rbirth_str, 1, str_length(rbirth_str) - 4),
    m_str = str_sub(rbirth_str, -4, -3),
    y_str = str_sub(rbirth_str, -2, -1)
  )
# the year is always the last two characters and the month the two before that,
# so counting from the end with negative positions works for strings of
# length 5 as well as length 6
# whatever remains at the front is the day
```

List the first twenty observations.

```{r}
hin95_birthdates |>
  select(rbirth_str, d_str, m_str, y_str) |>
  slice(1:20)
# compare with the numeric solution of the previous assignment
```

Convert the components to numbers and compare with assignment 5.

### 7. Postcodes and city names

```{r}
hin95_naw <- read_dta("data/hin95_naw.dta")
```

Extract the numeric part of the postcode.

```{r}
hin95_naw <- hin95_naw |>
  mutate(
    pcode_str = str_sub(postcode, 1, 4),
    npcode = as.numeric(pcode_str)
  )
# first take the substring "1234", which is still a string
# then convert that string to the number 1234
```


```{r}
hin95_naw |>
  skim(npcode)
# the mean of npcode is 4639. It is meaningless.
```

:::
:::