---
title: Dealing with missing data
---

As one of the more serious of the ten plagues of the social sciences, we now turn to missing values. We will not discuss extensively how to deal with missing values in a statistically appropriate way. Suppose you want to fit a regression model of `Y` on `X1` and `X2`. What should you do if `Y`, `X1`, or `X2` are sometimes missing? Some possible approaches:

- listwise deletion: discard all observations that are missing on one of the variables. This method is also known as "complete observations only", and is the default approach used by many statistical functions in R.

- pairwise deletion: correlations and covariances for two variables are computed from all observations that are not missing on those two variables. This method is available in functions such as `cor()` through the argument `use = "pairwise.complete.obs"`.

- single imputation methods make the data complete by imputing "reasonable" values for the missing values. The keyword here is of course "reasonable". A widely used method is the imputation of missing values of `x` with the mean of `x`. A somewhat better single imputation method uses predictions from a regression model, or random draws from the forecasted values of a regression model. The hot deck method imputes a missing value in observation `i` on `x` by the non-missing value of `x` in the observation `j` that most closely resembles observation `i`.

- multiple imputation methods impute missing values by multiple values, representing the distribution of plausible values for the missing values, conditional on the observed values and a model for the joint distribution of all variables. Time permitting, multiple imputation methods will be discussed later in the programme. In R, the package `mice` is commonly used for multiple imputation.

## R

In this part you need to learn about the following R functions and packages:

| Function       | Description                                     |
|----------------|-------------------------------------------------|
| `na_if()`      | Recode values to missing       |
| `replace_na()` | Recode missing to values         |
| `is.na()`      | Test whether values are missing                 |
| `vis_miss()`   | Description and visualization of missing values |

Examples:

Create a dummy variable for whether `educ` is missing:

```{r, eval=FALSE}
data <- data |>
  mutate(educ_mv = is.na(educ))
# educ_mv is a logical dummy (TRUE/FALSE)
```

Create a dummy `male`, set to missing if `sex` is missing:

```{r, eval=FALSE}
data <- data |>
  mutate(
    male = if_else(sex == 1, 1, 0)
  )
# assumes sex == 1 is male
# if_else() propagates NA from its condition: when sex is NA, the condition
# sex == 1 is also NA, so male becomes NA automatically.
```


List `edu` for observations for which `age` and `country` are not missing:

```{r, eval=FALSE}
data |>
  filter(!is.na(age), !is.na(country)) |>
  select(edu)
```


Recode the values 98 and 99 of `edu` into missing values:

```{r, eval=FALSE}
data <- data |>
  mutate(
    edu = na_if(edu, 98),
    edu = na_if(edu, 99)
  )
```

Recode missing values to 99:

```{r, eval=FALSE}
data <- data |>
  mutate(
    edu = replace_na(edu, 99)
  )
```

Count the number of missing values in a variable:

```{r, eval=FALSE}
data$educ |>
  is.na() |>
  sum()
```

Count the number of missing values per variable in a series of variables:

```{r, eval=FALSE}
data |>
  summarise(
    across(starts_with("good"), \(x) sum(is.na(x)))
  )
```

`skimr::skim()` gives the same information, together with other descriptive
statistics, for any set of selected variables:

```{r, eval=FALSE}
library(skimr)

data |>
  select(starts_with("good")) |>
  skim()
# the n_missing and complete_rate columns report missingness per variable
```

Display missing value patterns and their frequency in a series of variables:

```{r, eval=FALSE}
library(visdat)

data |>
  select(good01:good05) |>
  vis_miss()
```

## Assignments

### 1

Use the dataset `hin95_educ`.

How many missing values are there in the variable `educ`?

Create a dummy variable indicating whether `educ` is missing.

Create a copy of `educ` with missing values set to 99.

Create a new imputed variable, replacing the missing values by the unconditional mean of `educ` over the observed values.

Create another new imputed variable, replacing the missing values in `educ` by the gender-specific conditional means of `educ` over the observed values. Thus, for men we impute the mean of `educ` for men, and for women we impute the mean of `educ` for women.

Compare the means and standard deviations of the three versions of the original variable `educ` and the two imputed versions. Reflect on eventual differences.

### 2

(Advanced, assumes you already know regression analysis) You may also replace the missing values in a variable `y` by the predicted values of `y` on some appropriate explanatory variables. As an example, perform a regression of a respondent's `educ` on the education of the respondent's father, `feduc`, the education of the respondent's mother, `meduc`, and gender, `sex`. Replace missing values in `educ` by the predicted values of this regression.
Explain why `educ` still has missing values. Any suggestions how to proceed?

### 3.  

Consider the variables `rs280` to `rs286` in the dataset `hin95`.

- Create a variable counting the number of non-missing values in this series of variables. How many observations have no missing values on any of these variables? How many observations have exactly one missing value? How many observations are always missing in this series?

- Now select the observations without missing values. Generate a variable with the mean of these variables. What is the mean of this mean-variable?

- In the original dataset, create a variable with the mean over the available values. Thus, if for an observation the variables `v1`, `v3`, and `v8` are observed and the other variables are missing, for this observation you should return `(v1 + v3 + v8) / 3`. Use `rowMeans()` with care.

### 4.  

In a hypothetical dataset you have variables `male` (respondent's gender, 0 = female and 1 = male), `feduc` (father's education), and `meduc` (mother's education). You want to create a variable `ssedu` that represents the level of education of the same-sex parent, i.e., for men the father and for women the mother. Provide code, and discuss what would happen if any of the three variables are missing.

## Answers
```{r setup-missing}
library(tidyverse)
library(haven)
library(skimr)
library(visdat)

hin95_educ <- read_dta("data/hin95_educ.dta")
hin95 <- read_dta("data/hin95.dta")
```

### 1. Describing and imputing a single variable

How many missing values are there in `educ`?

```{r}
hin95_educ$educ |> 
  is.na() |>
  sum()
# number of missing values

hin95_educ$educ |>
  is.na() |>
  mean()
# the same information as a proportion
```

Create a dummy variable indicating whether `educ` is missing.
```{r}
hin95_educ <- hin95_educ |>
  mutate(
    educ_mv = is.na(educ)
  )
# educ_mv is TRUE when educ is missing
```

```{r}
table(hin95_educ$educ_mv)
# check the dummy against the count above
```

Create a copy of `educ` with missing values set to 99.

```{r}
hin95_educ <- hin95_educ |>
  mutate(
    educ99 = replace_na(educ, 99)
  )
# note that 99 is now an ordinary value: any mean or regression using educ99
# will treat it as a real education level, which it is not
```

Replace the missing values by the unconditional mean.

```{r}
hin95_educ <- hin95_educ |>
  mutate(
    educ_imp1 = replace_na(educ, mean(educ, na.rm = TRUE))
  )
# mean imputation over the observed values
```

Replace the missing values by the gender-specific conditional means.

```{r}
hin95_educ <- hin95_educ |>
  mutate(
    educ_imp2 = replace_na(educ, mean(educ, na.rm = TRUE)),
    .by = sex
  )
# .by = sex makes mean() compute a separate mean within each sex
# respondents with a missing sex form their own group and therefore keep
# their missing value on educ_imp2
```

Compare the three versions.
```{r}
hin95_educ |>
  select(educ, educ_imp1, educ_imp2) |>
  skim()
```

### 2. Regression imputation

Fit the regression on the complete cases.

```{r}
m_educ <- lm(educ ~ feduc + meduc + sex, data = hin95_educ)
summary(m_educ)
# lm() applies listwise deletion by default
```

Replace the missing values by the predicted values.

```{r}
hin95_educ <- hin95_educ |>
  mutate(
    educ_hat = predict(m_educ, newdata = hin95_educ),
    educ_imp3 = if_else(is.na(educ), educ_hat, educ)
  )
# predict() returns NA for any respondent with a missing predictor
# only the observed values of educ are left untouched
```

How many missing values remain?

```{r}
hin95_educ |>
  summarise(
    still_missing = sum(is.na(educ_imp3)),
    of_which_missing_predictor = sum(
      is.na(educ) & (is.na(feduc) | is.na(meduc) | is.na(sex))
    )
  )
# the two numbers should be equal
```


### 3. A series of variables

Count the number of non-missing values per respondent.

```{r}
hin95 <- hin95 |>
  mutate(
    n_missing = (is.na(rs280) + is.na(rs281) + is.na(rs282) + 
                 is.na(rs283) + is.na(rs284) + is.na(rs285) + is.na(rs286)),

    n_valid = 7 - n_missing # 7 is the number of variables in the series
  )

# more compact solution:
hin95 <- hin95 |>
  mutate(
    n_valid2 = rowSums(across(rs280:rs286, \(x) !is.na(x))),
    n_missing2 = rowSums(across(rs280:rs286, \(x) is.na(x)))
  )

```

Distribution of the number of missing answers:
```{r}
table(hin95$n_missing)
```

Select the complete observations and compute the mean.
```{r}
hin95_complete <- hin95 |>
      filter(n_missing == 0) |> 
      mutate(
        rs_mean = rowMeans(pick(rs280:rs286))
      ) 

hin95_complete  |>    
    skim(rs_mean)
# no na.rm is needed here because these rows have no missing values
```


Compute the mean over the available values, for all respondents.

```{r}
hin95 <- hin95 |>
  mutate(
    rs_mean_available =  rowMeans(pick(rs272:rs286), na.rm = TRUE)
    ) 
# Note that rowMeans(na.rm = TRUE) returns NaN, not NA, for a row with no valid values

hin95 |> 
  skim(rs_mean_available)
```


### 4. Education of the same-sex parent


```{r, eval=FALSE}
data <- data |>
  mutate(
    ssedu = if_else(male == 1, feduc, meduc)
  )
# male == 1 gives NA when male is missing, so ssedu becomes NA
# for men the father's education is taken, for women the mother's
```

