---
title: Dealing with missing data
---

As one of the more serious of the ten plagues of the social sciences, we now turn to missing values. We will not discuss extensively how to deal with missing values in a statistically appropriate way. Suppose you want to fit a regression model of `Y` on `X1` and `X2`. What should you do if `Y`, `X1`, or `X2` are sometimes missing? Some possible approaches:

- listwise deletion: discard all observations that are missing on one of the variables. This method is also known as "complete observations only", and is the default approach used by many statistical functions in R.

- pairwise deletion: correlations and covariances for two variables are computed from all observations that are not missing on those two variables. This method is available in functions such as `cor()` through the argument `use = "pairwise.complete.obs"`.

- single imputation methods make the data complete by imputing "reasonable" values for the missing values. The keyword here is of course "reasonable". A widely used method is the imputation of missing values of `x` with the mean of `x`. A somewhat better single imputation method uses predictions from a regression model, or random draws from the forecasted values of a regression model. The hot deck method imputes a missing value in observation `i` on `x` by the non-missing value of `x` in the observation `j` that most closely resembles observation `i`.

- multiple imputation methods impute missing values by multiple values, representing the distribution of plausible values for the missing values, conditional on the observed values and a model for the joint distribution of all variables. Time permitting, multiple imputation methods (Enders, 2010; van Buuren, 2020) will be discussed later in the programme. In R, the package `mice` is commonly used for multiple imputation.

## R

In this part you need to learn about the following R functions and packages:

| Function       | Description                                     |
|----------------|-------------------------------------------------|
| `na_if()`      | Change numeric values to missing values         |
| `replace_na()` | Change missing values to numeric values         |
| `is.na()`      | Test whether values are missing                 |
| `vis_miss()`   | Description and visualization of missing values |

Examples:

Create a dummy variable for whether `educ` is missing:

```{r, eval=FALSE}
data <- data |>   
  mutate(educ_mv = is.na(educ)) 
```

Create a dummy `male`, set to missing if `sex` is missing:

```{r, eval=FALSE}
data <- data |>   
  mutate(     
    male = if_else(       
      is.na(sex),       
      NA_real_,       
      if_else(sex == 1, 1, 0)     
      )   
    ) 
```

Create a dummy with `case_when()`:

```{r, eval=FALSE}
data <- data |>   
  mutate(     
    male = case_when(       
      sex == 1 ~ 1,       
      !is.na(sex) ~ 0,       
      TRUE ~ NA_real_     
      )   
    ) 
```

Tabulate `edu`, including missing values:

```{r, eval=FALSE}
table(data$edu, useNA = "ifany") 
```

List `edu` for observations for which `age` and `country` are not missing:

```{r, eval=FALSE}
data |>   
  filter(!is.na(age), !is.na(country)) |>   
  select(edu) 
```

Count the number of missing values among a list of variables:

```{r, eval=FALSE}
data <- data |>   
  mutate(     
    nmiss = rowSums(       
      is.na(select(., good01:good05))     
      )   
    ) 
```

Count the number of valid, non-missing, values among a list of variables:

```{r, eval=FALSE}
data <- data |>   
  mutate(     
    nvalid = rowSums(       
      !is.na(select(., good01:good05))     
      )   
    ) 
```

Recode the values 98 and 99 of `edu` into missing values:

```{r, eval=FALSE}
data <- data |>   
  mutate(     
    edu = na_if(edu, 98),     
    edu = na_if(edu, 99)   
    ) 
```

Recode missing values to 99:

```{r, eval=FALSE}
data <- data |>   
  mutate(     
    edu = replace_na(edu, 99)   
    ) 
```

Display frequencies of missing values in a series of variables:

```{r, eval=FALSE}
colSums(   
  is.na(     
    select(data, starts_with("good"))   
    ) 
  ) 
```

Display missing value patterns and their frequency in a series of variables:

```{r, eval=FALSE}
library(visdat)  

vis_miss(   
  select(data, good01:good05) 
  ) 
```

## Assignments

See `dh_missing_assignments.Rmd` for solutions.

### 1.  Use the dataset `hin95_educ`.

How many missing values are there in the variable `educ`?

Create a dummy variable indicating whether `educ` is missing.

Create a copy of `educ` with missing values set to 99.

Create a new imputed variable, replacing the missing values by the unconditional mean of `educ` over the observed values.

Create another new imputed variable, replacing the missing values in `educ` by the gender-specific conditional means of `educ` over the observed values. Thus, for men we impute the mean of `educ` for men, and for women we impute the mean of `educ` for women.

### 2.  Advanced, assumes you already know regression analysis.

You may also replace the missing values in a variable `y` by the predicted values of `y` on some appropriate explanatory variables.

As an example, perform a regression of a respondent's `educ` on the education of the respondent's father, `fedu`, the education of the respondent's mother, `medu`, and gender, `sex`.

Replace missing values in `educ` by the predicted values of this regression.

Explain why `educ` still has missing values. Any suggestions how to proceed?

### 3.  Consider the variables `rs272` to `rs286` in the dataset `hin95`.

- Create a variable counting the number of non-missing values in this series of variables. How many observations have no missing values on any of these variables? How many observations have exactly one missing value? How many observations are always missing in this series?

- Study the patterns in the missing values. For example, is the missing value pattern monotonic, meaning that if variable A is missing, variable B is also missing, but beware, not necessarily the reverse. You may use tools such as `vis_miss()` or tables based on `is.na()`.

- Now select the observations without missing values. Generate a variable with the mean of these variables. What is the mean of this mean-variable?

- Also create a variable with the mean over the available values. Thus, if for an observation the variables `v1`, `v3`, and `v8` are observed and the other variables are missing, for this observation you should return `(v1 + v3 + v8) / 3`. Use `rowMeans()` with care.

### 4.  In a dataset you have variables `male`, respondent's gender, 0 = female and 1 = male, `fedu`, father's education, and `medu`, mother's education.

You want to create a variable `ssedu` that represents the level of education of the same-sex parent, i.e., for men the father and for women the mother.

Give code, and carefully discuss how to deal with missing values in the three variables.

## Answers
Make sure this file is in the same R Project as the `data` folder. Use relative file paths, not full paths to your computer.

```{r setup-missing}
library(tidyverse)
library(haven)
library(visdat)

hin95_educ <- read_dta("data/hin95_educ.dta")
hin95 <- read_dta("data/hin95.dta")
```

### 1. Describing and imputing a single variable

How many missing values are there in `educ`?

```{r}
sum(is.na(hin95_educ$educ))
# number of missing values

mean(is.na(hin95_educ$educ))
# the same information as a proportion
```

Create a dummy variable indicating whether `educ` is missing.

```{r}
hin95_educ <- hin95_educ |>
  mutate(
    educ_mv = is.na(educ)
  )
# educ_mv is TRUE when educ is missing
```

```{r}
table(hin95_educ$educ_mv)
# check the dummy against the count above
```

Create a copy of `educ` with missing values set to 99.

```{r}
hin95_educ <- hin95_educ |>
  mutate(
    educ99 = replace_na(educ, 99)
  )
# note that 99 is now an ordinary value: any mean or regression using educ99
# will treat it as a real education level, which it is not
```

Replace the missing values by the unconditional mean.

```{r}
hin95_educ <- hin95_educ |>
  mutate(
    educ_imp1 = replace_na(educ, mean(educ, na.rm = TRUE))
  )
# mean imputation over the observed values
```

Replace the missing values by the gender-specific conditional means.

```{r}
hin95_educ <- hin95_educ |>
  mutate(
    educ_imp2 = replace_na(educ, mean(educ, na.rm = TRUE)),
    .by = sex
  )
# .by = sex makes mean() compute a separate mean within each sex
# respondents with a missing sex form their own group and therefore keep
# their missing value on educ_imp2
```

Compare the three versions.

```{r}
hin95_educ |>
  summarise(
    n = n(),
    mean_observed = mean(educ, na.rm = TRUE),
    mean_imp1 = mean(educ_imp1, na.rm = TRUE),
    mean_imp2 = mean(educ_imp2, na.rm = TRUE),
    sd_observed = sd(educ, na.rm = TRUE),
    sd_imp1 = sd(educ_imp1, na.rm = TRUE),
    sd_imp2 = sd(educ_imp2, na.rm = TRUE)
  )
# the means barely change, but the standard deviations shrink
# every imputed case sits exactly on a mean, so mean imputation invents
# certainty that the data do not contain
```

### 2. Regression imputation

Fit the regression on the complete cases.

```{r}
m_educ <- lm(educ ~ fedu + medu + sex, data = hin95_educ)
summary(m_educ)
# lm() applies listwise deletion by default
```

Replace the missing values by the predicted values.

```{r}
hin95_educ <- hin95_educ |>
  mutate(
    educ_hat = predict(m_educ, newdata = hin95_educ),
    educ_imp3 = if_else(is.na(educ), educ_hat, educ)
  )
# predict() returns NA for any respondent with a missing predictor
# only the observed values of educ are left untouched
```

How many missing values remain?

```{r}
hin95_educ |>
  summarise(
    still_missing = sum(is.na(educ_imp3)),
    of_which_missing_predictor = sum(
      is.na(educ) & (is.na(fedu) | is.na(medu) | is.na(sex))
    )
  )
# the two numbers should be equal
```

```{r}
# Why does educ still have missing values?
#
# A prediction requires all three predictors. Respondents who are missing on
# educ AND on fedu, medu, or sex cannot be given a predicted value, so their
# missing value survives the imputation. Missingness in surveys tends to
# cluster: someone who skips questions about their own education often skips
# the parental questions as well, so this overlap is usually substantial.
#
# How to proceed?
#
# - Fit a second, simpler model using only the predictors that are observed
#   for the remaining cases, and impute from that.
# - Impute the predictors first, then impute educ.
# - Note that regression imputation puts every imputed case exactly on the
#   regression line, which understates the variance and inflates the
#   correlations. Adding a random residual, so-called stochastic regression
#   imputation, repairs part of this.
# - The proper solution is multiple imputation, for instance with the mice
#   package, which imputes all variables jointly and carries the uncertainty
#   about the imputed values through to the standard errors.
```

Add a random residual to illustrate the last point.

```{r}
set.seed(1)

sigma_educ <- sigma(m_educ)

hin95_educ <- hin95_educ |>
  mutate(
    educ_imp4 = if_else(
      is.na(educ),
      educ_hat + rnorm(n(), 0, sigma_educ),
      educ
    )
  )
# stochastic regression imputation
# set.seed() makes the result reproducible
```

```{r}
hin95_educ |>
  summarise(
    sd_observed = sd(educ, na.rm = TRUE),
    sd_regression = sd(educ_imp3, na.rm = TRUE),
    sd_stochastic = sd(educ_imp4, na.rm = TRUE)
  )
# the stochastic version preserves the spread of the original variable better
```

### 3. A series of variables

Count the number of non-missing values per respondent.

```{r}
n_items <- ncol(select(hin95, rs272:rs286))
n_items
# how many variables are in the series
```

```{r}
hin95 <- hin95 |>
  mutate(
    nvalid = rowSums(!is.na(pick(rs272:rs286))),
    nmiss = n_items - nvalid
  )
# pick() selects the columns inside mutate()
```

```{r}
table(hin95$nvalid)
# distribution of the number of valid answers
```

```{r}
hin95 |>
  summarise(
    complete = sum(nvalid == n_items),
    exactly_one_missing = sum(nmiss == 1),
    all_missing = sum(nvalid == 0)
  )
# the three counts asked for in the assignment
```

Visualise the missing value patterns.

```{r}
hin95 |>
  select(rs272:rs286) |>
  vis_miss()
# each row is a respondent, each column a variable
# a staircase shape suggests a monotone pattern
```

Study the patterns as a table.

```{r}
hin95 |>
  select(rs272:rs286) |>
  is.na() |>
  colSums()
# how many missing values does each variable have separately
```

```{r}
hin95 |>
  select(rs272:rs286) |>
  is.na() |>
  as_tibble() |>
  count(across(everything()), sort = TRUE) |>
  slice(1:10)
# the ten most frequent missing value patterns
# TRUE means missing
```

```{r}
# Is the pattern monotone?
#
# If the variables are ordered so that the number of missing values only
# increases from left to right, and if a respondent missing on one variable
# is also missing on every later variable, the pattern is monotone. This is
# typical for a questionnaire that respondents abandon partway through: they
# answer up to a point and then stop.
#
# Compare the column sums above with the patterns table. If the frequent
# patterns are all of the form "observed ... observed, missing ... missing",
# the pattern is monotone. Isolated gaps in the middle mean it is not.
```

Select the complete observations and compute the mean.

```{r}
hin95_complete <- hin95 |>
  filter(nvalid == n_items) |>
  mutate(
    rs_mean = rowMeans(pick(rs272:rs286))
  )
# no na.rm is needed here because these rows have no missing values
```

```{r}
hin95_complete |>
  summarise(
    n = n(),
    mean_of_means = mean(rs_mean)
  )
# the mean of the mean-variable over the complete cases
```

Compute the mean over the available values, for all respondents.

```{r}
hin95 <- hin95 |>
  mutate(
    rs_mean_available = if_else(
      nvalid == 0,
      NA_real_,
      rowMeans(pick(rs272:rs286), na.rm = TRUE)
    )
  )
# rowMeans(na.rm = TRUE) returns NaN, not NA, for a row with no valid values
# the if_else() turns those rows into a proper missing value
```

```{r}
hin95 |>
  summarise(
    n_valid_scale = sum(!is.na(rs_mean_available)),
    mean_available = mean(rs_mean_available, na.rm = TRUE)
  )
# compare with the complete-case mean above
# a difference between the two suggests that respondents with missing values
# differ systematically from respondents without them
```

### 4. Education of the same-sex parent

```{r}
# The construction itself is one line, but the missing values need thought.
# There are three sources of missingness, and they are not equivalent.
#
# 1. male is missing. We do not know which parent to look at, so ssedu must
#    be missing, even if both fedu and medu are observed.
# 2. The same-sex parent's education is missing. Then ssedu is missing:
#    that is precisely the value we wanted.
# 3. The OTHER parent's education is missing. This is irrelevant. A woman
#    with an observed medu should get a value for ssedu even if fedu is
#    missing. Code that is careless about this throws away usable cases.
#
# if_else() handles all three correctly: it propagates NA from the condition,
# and it evaluates only the branch that applies.
```

```{r, eval=FALSE}
data <- data |>
  mutate(
    ssedu = if_else(male == 1, fedu, medu)
  )
# male == 1 gives NA when male is missing, so ssedu becomes NA
# for men the father's education is taken, for women the mother's
```

