---
title: Adding cases and variables
---

In this chapter, you will learn about how to combine different datasets:

- Add cases to a dataset ("append cases").

- Add variables on the same cases from another dataset (“one-to-one merge”).

- Add contextual information to a dataset (“many-to-one merge”).

- Form all possible combinations of observations in two datasets.

We will elaborate on and illustrate these tasks with simple examples. As a first example, you organize an international data collection, where all countries use the same questionnaire. You obtain datasets from the Netherlands, from the UK, etc. The datasets contain (largely) the same variables. You want to make one dataset of these country datasets, with an additional variable country recording the data source.

 As a second example, we consider that the two files to be combined contain different variables. We need to distinguish two cases:

- **one-to-one merge:** Both datasets contain variables about the same units. For example, you have two datasets on countries, one with economic variables, and one with demographic variables. You want to make one dataset in which the observations are countries; the dataset should contain both the economic and the demographic variables.

- **many-to-one merge / matched merge / table lookup:** The datasets describe different units. Example: One dataset describes households, and contains a variable containing the city of residence. The second dataset describes cities, with variables for the number of inhabitants, the political composition of the city council etc. We want to form a dataset on household, in which properties of the city of residence are included. Each household is located in one city, but one city may contain more than one household in the data. This is known as a many-to-one link, table lookup or many-to-one merge.

A crucial concept in merging is a key. A key comprises one or more variables used to match observations from the two datasets. In the example of one-to-one merging, the key is likely a country identifier. In the example on many-to-one merging, the key is the city identifier. The key variable or variables should occur in both datasets. In R, key variables do not necessarily need to have the same name in both datasets, but using the same name is usually clearer and less error-prone.

PS: The first type of merge is also possible without a key variable or key variables, and strictly on sequential order of the files: the first record of file A with the first record of file B, the second record of file A with the second record of file B, etc. This type of merging is very error prone. We urge you never to merge without keys.

### Task 1: adding cases

As a pet example of adding cases, we consider two datasets. The dataset `world95_oecd_eu` describes the population size of a selection of 4 European OECD members:

| country     |   pop |
|-------------|------:|
| Austria     |  8000 |
| Belgium     | 10100 |
| Denmark     |  5200 |
| Netherlands | 15400 |

while `world95_oecd_extra` describes a selection of 3 non-European OECD members:

| country   |    pop |
|-----------|-------:|
| Australia |  17800 |
| Canada    |  29100 |
| USA       | 260800 |

The first task that we want you to master is to create a new dataset that contains all 7 OECD countries:

| country     |    pop |
|-------------|-------:|
| Austria     |   8000 |
| Belgium     |  10100 |
| Denmark     |   5200 |
| Netherlands |  15400 |
| Australia   |  17800 |
| Canada      |  29100 |
| USA         | 260800 |


### Task 2: one-to-one merge

As a pet example of 1:1 merging, we consider two datasets with characteristics of a collection of countries, namely the population size, dataset `world95_3_pop`:

| country     |   pop |
|-------------|------:|
| Afghanistan | 20500 |
| Argentina   | 33900 |
| Armenia     |  3700 |

and religion in dataset `world95_3_rel`:

| country     | religion |
|-------------|----------|
| Afghanistan | Muslim   |
| Argentina   | Catholic |

Note that the second dataset does not contain Armenia; Armenia is missing in the second dataset. As a second task we will address the creation of the one-to-one merged dataset that contains two characteristics of all countries involved:

| country     |   pop | religion |
|-------------|------:|----------|
| Afghanistan | 20500 | Muslim   |
| Argentina   | 33900 | Catholic |
| Armenia     |  3700 | NA       |

Note that in the combined dataset, Armenia is included, but the religion of Armenia is missing.

### Task 2: many-to-one merge

A paradigmatic example of many-to-one merging involves adding contextual data to information on individual subjects. We consider a pet example with 4 individuals from various countries:

|  id | male | age | country     |
|----:|-----:|----:|-------------|
|   1 |    0 |  24 | Afghanistan |
|   2 |    1 |  41 | Afghanistan |
|   3 |    0 |  58 | Armenia     |
|   4 |    0 |  35 | Netherlands |

A separate file contains characteristics of various countries:

| country     |   pop |
|-------------|------:|
| Afghanistan | 20500 |
| Argentina   | 33900 |
| Armenia     |  3700 |

Note that the individual dataset contains a subject from the Netherlands, but the population size of the Netherlands is not available. Moreover, we have contextual information about Argentina, but no subject from Argentina. As your third task, you have to learn how to create the many-to-one merge of the datasets, adding country properties to the individual data:

|  id | male | age | country     |   pop |
|----:|-----:|----:|-------------|------:|
|   1 |    0 |  24 | Afghanistan | 20500 |
|   2 |    1 |  41 | Afghanistan | 20500 |
|   3 |    0 |  58 | Armenia     |  3700 |
|   4 |    0 |  35 | Netherlands |    NA |

### Task 4: many-to-many merge
We want to analyze school choice by parents. In the pet example we have 4 parents each of whom can choose between 3 schools. We want to construct a dataset that consists of combinations of parents and schools.

parents:

| parentid | status | neighb | choice |
|---------:|-------:|-------:|-------:|
|        1 |      3 |      1 |      2 |
|        2 |      1 |      1 |      3 |
|        3 |      2 |      2 |      1 |
|        4 |      1 |      2 |      1 |

schools:

| schoolid | public |
|---------:|-------:|
|        1 |      1 |
|        2 |      1 |
|        3 |      0 |

pairs:

| parentid | status  | neighb | choice  | schoolid |        public |
|---------:|--------:|-------:|--------:|---------:|--------------:|
|        1 |       3 |      1 |       2 |        1 |             1 |
|        1 |       3 |      1 |       2 |        2 |             1 |
|        1 |       3 |      1 |       2 |        3 |             0 |
|        2 |       1 |      1 |       3 |        1 |             1 |
|        2 |       1 |      1 |       3 |        2 |             1 |
|        2 |       1 |      1 |       3 |        3 |             0 |
|        3 |       2 |      2 |       1 |        1 |             1 |
|        3 |       2 |      2 |       1 |        2 |             1 |
|        3 |       2 |      2 |       1 |        3 |             0 |
|        4 |       1 |      2 |       1 |        1 |             1 |
|        4 |       1 |      2 |       1 |        2 |             1 |
|        4 |       1 |      2 |       1 |        3 |             0 |

## R

You should read about the following R functions:

| Function      | Description                   |
|---------------|-------------------------------|
| `bind_rows()` | Append datasets, add cases    |
| `left_join()` | Merge datasets, add variables |

We present an R snippet for the four tasks:

### Add cases

```{r, eval=FALSE}
oecd <- bind_rows(
  world95_oecd_eu,
  world95_oecd_extra
)

```

### one-to-one merge

```{r, eval=FALSE}
countries <- left_join(
              world95_3_pop,
              world95_3_rel,
              by = "country" 
            )
```

Note that in `left_join()`, the argument `by` specifies the key variable. If the key variable has the same name in both datasets, you can omit the argument `by`, but it is good practice to include it for clarity. If the key variable has different names in the two datasets, you can specify them as a named vector, e.g., `by = c("country" = "country_name")`. If the key is a composite, you can specify multiple variables, e.g., `by = c("country", "year")`.

### one-to-many merge

```{r, eval=FALSE}
individuals <-  left_join(
                  merge_indiv,
                  world95_3_pop,
                  by = "country"
                )

```

### All combinations of parents and schools

Note that "many-to-many" merges are relatively rare, and creating all combinations of units (which multiplies the number of rows) is very often *not* what you want. This is why in the code below we have to explicitly tell `left_join()` that we know what we are doing and really want a many-to-many merge; otherwise it will still produce the combinations but throw a warning. 

```{r, eval=FALSE}
# Let's specify a fake join key, with the same for all values in both datasets
schools <- schools |>
            mutate(join_key = 1)

parents <- parents |>
            mutate(join_key = 1)

pairs <- left_join(
              parents,
              schools,
              by = "join_key",
              relationship = "many-to-many",
            ) |>
          select(-join_key)


# A more compact way, using tidyr::crossing():
# (will fail if there are variables with the same name in both datasets, line "join_key",
# but can be fixed with the .name_repair argument)
crossing(parents, join_key)
```



Some notes on combining datasets

- To append datasets, make sure that corresponding variables in both datasets have the same names, and are coded in the same way. For example, you do not want `sex` to be coded 0 = male and 1 = female in one dataset, and 0 = female and 1 = male in the other.
- Issues to consider with key variables for merging:
  - Make sure that the keys are coded identically in both datasets. If you merge on `country`, a given country code should obviously refer to the same country in both datasets.
  - In a 1:1 merge, make sure to always use a key, e.g., the respondent number. Do not rely on the assumption that the first observation in one dataset corresponds with the first observation in the other dataset.
  - R allows composite keys for merging, i.e., the key may consist of multiple variables, e.g., `country` and `year`.
  - Do not merge on keys with missing values, unless you really know what you are doing.
  - `left_join()` keeps all observations from the first dataset and adds matching information from the second dataset. Observations in the first dataset without a match receive NA on the variables from the second dataset.
  - As always: Do not assume that combining files was successful if R does not complain. Check the number of observations! Check the variables! Check values! Check everything!
  
  
## Assignments

### 1. Synthetic Eurobarometer data

The datasets `eurobaro2000_ge`, `eurobaro2000_nl`, and `eurobaro2000_uk` contain subsets of the Eurobarometer 2000 data, with small selections of respondents from Germany, the Netherlands, and the United Kingdom. These datasets comprise demographic information of respondents, `sex` and `age`, and their trust in two institutions of the European Union, the EU commission, `tcomm`, and the EU parliament, `tparl`, recorded using 5 point scales. Use `bind_rows()` to append these three datasets to create one dataset `eb2000` for the three countries, adding a string variable `country` recording the data source. Add a `year` variable that includes the value 2000.

The datasets `eurobaro2002` and `eurobaro2004` similarly contain Eurobarometer data for the same three countries for the years 2002 and 2004. We want you to combine these two datasets with `eb2000`. Make sure that the in the combined dataset there is a variable contaning the year of each survey. However, there are some complications that you have to deal with.

- the variable `Country` in 2002: The United Kingdom is not abbreviated to `UK` as in 2000. (Also note the variable name)
- the variable `country` in 2004: It is a labelled numeric variable, not a string variable.
- the variable `sex` in 2002 and 2004: The meaning of 1 and 2 are reversed compared to 2000. Use the coding of 2000. 
- in 2002 and 2004, trust was also measured for the European Court of Law.
- in 2004, trust was measured on a binary scale, 0 = no and 1 = yes, rather than on the 5 point scale used in 2000 and 2002. We do not want to discard information, but also should not bind columns whose values represent different things. Think of a solution and implement it. 

We also have two contextual datasets. `eudata` contains the population size (in September 2012, ignore the mismatch) of these EU countries. The dataset `euydata` describes the GDP in some EU countries in the years 2000 to 2004, rounded in 1000 EUR units. Make sure you check thge variable names and values before joining. Enrich the individual data with these two contextual datasets using `left_join()`.

### 2. Add cases: appending datasets

In this assignment we consider how to combine observations from multiple files. For instance, data entry from paper questionnaires is performed by different people, yielding different datasets with the *same variables* for *different subjects*.

- As this exercise is meant to develop your computer skills, we are going to generate two disjoint sets of records ourselves. Load `sbs399`. Select the men and save them in the object `sbs399m`. Then select the women and save them in the object `sbs399f`. 

- To reconstruct the full dataset `sbs399` from these two partial datasets on women and men respectively, append the files using `bind_rows()`. Check that the number of cases is unchanged.

- Now do a more complicated append of the files involving different subjects and partially different variables. To simulate this, delete some variables from the male file and some other variables from the female file. Append the files using `bind_rows()`. How does R deal with the variables that appear in one dataset, but not in the other?

### 3. 1:1 merge: adding variables

In this assignment, we consider how to combine multiple files that comprise different variables for the same subjects. In real life, this may for instance occur in a panel study, in which data from different waves are stored in separate files. We require you create such datasets for training purposes.

- Use `isbs399`. This is a variant of `sbs399` in which we excluded same-sex households. If you want to, you may consider constructing `isbs399` from `sbs399` yourself. This dataset comprises couples nested in households and singles. Respondents are identified with `hrespnr`. Save two subsets of `isbs399`, one with the variables `irespnr`, `hrespnr`, `sex`, `ybirth`, and `age`; the other with `irespnr`, `hrespnr`, `sex`, `fage`, and `fpres`.
- Merge the two files you just created, using `irespnr` as the key and `left_join()`. Ascertain that the merged file is correct.
- Now create the two subsets of `isbs399` again, excluding the first observation in the first file, and the second observation in the second file. Merge again on `hrespnr` using `left_join()`. Inspect the first two observations. What do you conclude?
  - Extra: experiment with `full_join()` instead of `left_join()`. What do you conclude?
- Merge the two datasets once more, this time matching on `hrespnr`. Why does it not work? Yes, `hrespnr` does not uniquely identify observations. The pair `hrespnr` and `sex` does uniquely identify subjects, and so they make a valid composite key. Merge on the pair `hrespnr` and `sex`, and convince yourself that this leads to the same result as merging on `hrespnr`.

### 4. n:1 merge: contextual enrichment

In this assignment we consider how to enrich an individual level dataset with contextual data. The dataset `sbs399` contains information on individuals nested in couples or households, identified by the household identifier `hrespnr`. The dataset `hin95_house` provides properties of the couple's house, ownership, number of rooms, etc. Thus, `hin95_house` also contains the household identifier.

Merge `sbs399` and `hin95_house` on `hrespnr` using `left_join()`.

### 5. Cities and regions

We consider a partially synthetic example with three datasets. The dataset `mm_hn95data` contains information about households, including the city of residence, and the cities in which the husband and wife work. The second dataset `mm_cities` contains information about cities, including the region to which the city belongs. The third dataset `mm_regions` contains information about regions. For your convenience, we list the variables in the three datasets.

- `mm_hn95data`
  - `hhid` household id
  - `city` city of residence
  - `hhinc` household income
  - `nkids` number of kids
  - `hrace` husband: race
  - `hwcity` work husband: city
  - `wrace` wife: race
  - `wwcity` work wife: city
  - `religion` religion of couple
  - `vchurch` frequency church visit couple

- `mm_cities`
  - `cityid` city identifier
  - `inhabit` number of inhabitants
  - `region` region identifier
  - `jobs` number of jobs
  - `vacant` number of vacancies

- `mm_regions`
  - `regid` region identifier
  - `growth` economic growth in 1990

- Start with the dataset on households `mm_hn95data`. Merge in properties of the city in which the households live from the dataset  `mm_cities` using `left_join()`. 
- Continuing with this merged dataset, now add the economic growth in the region of residence from the dataset `mm_regions` using `left_join()`.
- We start again with the unmodified dataset `mm_hn95data`. This dataset contains the ids of three cities relevant to the households: the city in which they live, `city`, the city in which the husband works, `hwcity`, and the city in which the wife works, `wwcity`. We want you to merge in information about each of the cities, adding three variables, e.g., named `res_inhabit`, `hwcity_inhabit`, and `wwcity_inhabit`, for the number of inhabitants in each of the cities. Use `left_join()` multiple times.


::: {.content-visible when-format="html"}

## Answers

::: {.callout-note appearance="simple" collapse=true}
## See answers



```{r setup-merge}
library(tidyverse)
library(haven)
```

### 1. Synthetic Eurobarometer data

Inspect the three 2000 datasets before appending them.

```{r}
eurobaro2000_ge <- read_dta("data/eurobaro2000_ge.dta")
eurobaro2000_nl <- read_dta("data/eurobaro2000_nl.dta")
eurobaro2000_uk <- read_dta("data/eurobaro2000_uk.dta")

glimpse(eurobaro2000_ge)
glimpse(eurobaro2000_nl)
glimpse(eurobaro2000_uk)
```

Append the three country files and add `country` and `year`.
```{r}
# We can add new variables on the spot, calling mutate() for each country
# inside bind_rows(). Here we even use pipes, but "mutate(eurobaro2000_ge, country = "Germany")" would of course also work.
# Or we could modify each dataset first, and only them use them in bind_rows()
eb2000 <- bind_rows(
  eurobaro2000_ge |> mutate(country = "Germany"),
  eurobaro2000_nl |> mutate(country = "Netherlands"),
  eurobaro2000_uk |> mutate(country = "UK")
) |>
  mutate(year = 2000)

eb2000
```

Similar result (only the columns order differs) with the `.id` argument, which is shorter and less error-prone.
```{r}
eb2000_alt <- bind_rows(
                Germany = eurobaro2000_ge,
                Netherlands = eurobaro2000_nl,
                UK = eurobaro2000_uk,
                .id = "country"
            ) |>
            mutate(year = 2000)

eb2000_alt
```

Now inspect the 2002 and 2004 files. Each complication has to be resolved before appending, because `bind_rows()` stacks variables by name and will happily combine variables that mean different things.

```{r}
eurobaro2002 <- read_dta("data/eurobaro2002.dta")
eurobaro2004 <- read_dta("data/eurobaro2004.dta")

eurobaro2002
eurobaro2004
```

Harmonise 2002:
```{r}
eb2002 <- eurobaro2002 |>
  rename(country = Country) |>
  mutate(
    country = replace_values(
                  country,
                  "United Kingdom" ~ "UK",
              ),
    sex = 3 - sex,
    year = 2002
  )

eb2002
```

Harmonise 2004.
```{r}
eb2004 <- eurobaro2004 |>
  mutate(
    country = recode_values(country,
                            1 ~ "Germany",
                            2 ~ "Netherlands",
                            3 ~ "UK"),
    sex = 3 - sex,
    year = 2004
  )

# Here we recoded the country variable manually. Thuis would be cumbersome and
# error-prone for a larger number of countries. If you have a 'labelled' variable 
# as we have here (these are created when data is imported with labels, e.g. from Stata),
# you can use the function labelled::to_character() to convert it to a character 
# variable with the labels as values. 
```


The additional trust measure (`tcourt`) is not a problem, it is just missing for 2000. The bigger problem is that the 2004 trust variables are on a different scale. So let's just give them a different name in 2004, and keep the original variables for 2000 and 2002. 

```{r}
eb2004 <- eb2004 |>
  rename(
    tcomm2 = tcomm,
    tparl2 = tparl,
    tcourt2 = tcourt
  )
```

Now we can append append the three years.

```{r}
eb <- bind_rows(eb2000, eb2002, eb2004)

eb 
```

Add the contextual data. 
```{r}
eudata <- read_dta("data/eudata.dta")
euydata <- read_dta("data/euydata.dta")

eudata
```

```{r}
euydata
```
Note that year is stored in the variable `yr` in `euydata`, and that the country names are in small caps in both contextual datasets. We harmonise the variable names and values before merging (in this case it would be easier to change the values in `eb` to small caps, but it is good to see other approaches).

```{r}
eudata <- eudata |>
  mutate(country = str_to_title(country),
         country = if_else(country == "Uk", "UK", country))
eudata
```

```{r}
eudata <- eudata |>
  mutate(country = str_to_title(country),
         country = if_else(country == "Uk", "UK", country))
eudata
```

```{r}
euydata <- euydata |>
  rename(year = yr) |>
  mutate(country = str_to_title(country),
         country = if_else(country == "Uk", "UK", country))
euydata
```

`eudata` varies by country only, so the key is `country`; `euydata` varies by country and year, so the key is the pair. 

```{r}
eb_enriched <- eb |>
  left_join(eudata, by = "country") |>
  left_join(euydata, by = c("country", "year"))
# if we haden't change yr to year in euydata we would have to use by = c("country", "year" = "yr") instead

eb_enriched
```

Check that the number of cases is unchanged after the enrichment.
```{r}
nrow(eb)
nrow(eb_enriched)
```

### 2. Add cases: appending datasets

Split `sbs399` into two disjoint datasets.

```{r}
sbs399 <- read_dta("data/sbs399.dta")

labelled::print_labels(sbs399$sex)
# sex = 0 for men and sex = 1 for women
```

```{r}
sbs399m <- sbs399 |> filter(sex == 0)
sbs399f <- sbs399 |> filter(sex == 1)


nrow(sbs399m) + nrow(sbs399f) == nrow(sbs399)
# the two subsets should partition the original dataset
```

Reconstruct the full dataset.

```{r}
sbs399_reconstructed <- bind_rows(sbs399m, sbs399f)

nrow(sbs399_reconstructed)
nrow(sbs399)
# the same number of cases
```


Now drop different variables from each file.

```{r}
sbs399m_part <- sbs399m |> select(-age, -educ)
sbs399f_part <- sbs399f |> select(-fage, -fpres)

combined_part <- bind_rows(sbs399m_part, sbs399f_part)
```

`bind_rows()` keeps the union of the variables and fills the gaps with NA. In this examplem, age and educ are missing for men, and fage and fpres are missing for women. Let's check that.
```{r}
combined_part |> 
  select(sex, age, educ, fage, fpres) 

combined_part |>
  summarise(
    n = n(),
    missing_age = sum(is.na(age)),
    missing_educ = sum(is.na(educ)),
    missing_fage = sum(is.na(fage)),
    missing_fpres = sum(is.na(fpres)),
    .by = sex
  )

```


### 3. 1:1 merge: adding variables

```{r}
isbs399 <- read_dta("data/isbs399.dta")
```

Create two subsets with different variables for the same respondents.

```{r}
part_a <- isbs399 |>
  select(irespnr, hrespnr, sex, ybirth, age)

part_b <- isbs399 |>
  select(irespnr, hrespnr, sex, fage, fpres)

nrow(part_a)
nrow(part_b)
```

Merge on `irespnr`.

```{r}
merged <-  left_join(
              part_a,
              part_b,
              by = c("irespnr")
            )

merged
```

Note that `hrespnr` and `sex` are duplicated in the merged dataset. R automatically appends `.x` and `.y` to duplicate variable names. We could instead drop the duplicate variables from one of the datasets before merging, or, since they content is identical across datasets, we could include them in the `by` argument of `left_join()`, which will keep only one copy of the variable in the merged dataset.

```{r}
merged |>
  summarise(
    n = n(),
    missing_fage = sum(is.na(fage)),
    missing_fpres = sum(is.na(fpres))
  )
# no rows added, no unexpected missing values
```

Now drop one observation from each file and merge again.

```{r}
part_a2 <- part_a |> slice(-1)

# Let's also drop the duplicate variables from part_b2, since we don't need them in the merged dataset.
part_b2 <- part_b |>
           slice(-2) |> 
           select(-hrespnr, -sex)

merged2 <-  left_join(
                  part_a2, 
                  part_b2, 
                  by = "irespnr"
            )

merged2 |> 
  slice(1:2)
```

The result depends on which dataset you start from. `left_join()` is not symmetric: rows unmatched on the left (first dataset, in this case `part_a2`) are kept, rows unmatched on the right (second dataset,  `part_b2`) are silently discarded. So you have to keep in mind that the first dataset is the one you want to keep all rows from, and the second dataset is the one you want to add variables from. `full_join()` keeps unmatched rows from both sides. This may or may not be what you want.

Merge on `hrespnr` instead.

```{r}
merged_wrong <-  left_join(
                        part_a,
                        part_b, 
                        by = "hrespnr"
                      )

nrow(part_a)
nrow(merged_wrong)
```

Use the composite key `hrespnr` and `sex`.

```{r}
merged_composite <-  left_join(
                            part_a,
                            part_b, 
                            by = c("hrespnr", "sex")
                      )


nrow(merged_composite)
nrow(merged)
```


### 4. n:1 merge: contextual enrichment

```{r}
hin95_house <- read_dta("data/hin95_house.dta")
```

Some checks: 
```{r}
nrow(sbs399)
nrow(hin95_house)
n_distinct(hin95_house$hrespnr) == nrow(hin95_house)
# the contextual file should have exactly one record per household
```

Merge household information into the individual-level dataset. Note that `sbs399 |> left_join(hin95_house, by = "hrespnr")` is of course equivalent to `left_join(hin95_house, sbs399, by = "hrespnr")`, but the first form helps highlight the assymmetry mentioned before: the first dataset is the one we want to keep all rows from, the second dataset is the one we want to add variables from. 

```{r}
sbs399_house <- sbs399 |>
                left_join(hin95_house, 
                          by = "hrespnr")

nrow(sbs399)
nrow(sbs399_house)
```


### 5. Cities and regions

```{r}
mm_hn95data <- read_dta("data/mm_hn95data.dta")
mm_cities <- read_dta("data/mm_cities.dta")
mm_regions <- read_dta("data/mm_regions.dta")
```

Merge the city properties into the household data. Note that the key variables have different names: city on the left, cityid on the right.

```{r}
names(mm_hn95data)
names(mm_cities)

# Check that there is exactly one record per city in mm_cities,
n_distinct(mm_cities$cityid) == nrow(mm_cities)
```

Instead of renaming the city variable in one of the datasets, we can specify the different names in the `by` argument of `left_join()`. The left-hand name is kept in the result.
```{r}
hh_city <- mm_hn95data |>
            left_join(mm_cities,
                       by = c("city" = "cityid")
                      )

nrow(mm_hn95data)
nrow(hh_city)
hh_city |> 
  slice(1:10) 
```


Add the regional data.

```{r}
n_distinct(mm_regions$regid) == nrow(mm_regions)

hh_city_region <- hh_city |>
                  left_join(mm_regions, 
                            by = c("region" = "regid")
                            )

hh_city_region |> 
  slice(1:10)
```



Merge city information three times, for the city of residence and both work cities. We can in one chain. Note that we start with the original dataset `mm_hn95data`, then join `mm_cities` three times, each time we select only the relevant columns (`cityid` and `inhabit`, while changing the name of the latter) and specify a different key in the `by` argument.

```{r}
mm_hn95data_3 <- mm_hn95data |>
  left_join(
    mm_cities |> select(cityid, res_inhabit = inhabit),
    by = c("city" = "cityid")
  ) |>
  left_join(
    mm_cities |> select(cityid, hwcity_inhabit = inhabit),
    by = c("hwcity" = "cityid")
  ) |>
  left_join(
    mm_cities |> select(cityid, wwcity_inhabit = inhabit),
    by = c("wwcity" = "cityid")
  )

nrow(mm_hn95data)
nrow(mm_hn95data_3)

mm_hn95data_3 |> 
  slice(1:10)
```

:::
:::