---
title: Restructuring data
---

We may present panel data, data on people indicated by `id`, for different years indicated by `year`, in long format. To keep things simple, suppose for now that we only measured one time-varying variable, income `inc`, alongside the time-constant variable `male`:

| id  | year | inc | male |
|-----|------|-----|------|
| 1   | 2000 | 10  | 0    |
| 1   | 2001 | 12  | 0    |
| 1   | 2002 | 13  | 0    |
| 2   | 2000 | 14  | 1    |
| 2   | 2001 | 14  | 1    |
| 2   | 2002 | 13  | 1    |
| 3   | 2000 | 20  | 0    |
| 3   | 2002 | 22  | 0    |

or in wide format:

| id  | inc_2000 | inc_2001 | inc_2002 | male |
|-----|----------|----------|----------|------|
| 1   | 10       | 12       | 13       | 0    |
| 2   | 14       | 14       | 13       | 1    |
| 3   | 20       | NA       | 22       | 0    |

Be sure that you understand that these two datasets contain the same information, but only represent the information in a different way. One can translate one form into the other form. The variable `inc` is said to be time-varying, while `male` is time-constant.

In this part you practice the R functions that restructure data between long format and wide format. You need to learn about this because your data may be in one format, while the data management or statistical method you want to use requires the data to be in the other format.

For instance, multilevel analysis typically requires data to be in long format, e.g., measurements at different time points are stored in different rows, while structural equation modeling often requires wide data, e.g., measurements stored in different variables. In SaSR7 you will learn about these statistical methods. For now, we only introduce how to get the data into the correct format.

## R

You should read about the following R functions:

| Function         | Description                           |
|------------------|---------------------------------------|
| `pivot_longer()` | Convert data from wide to long format |
| `pivot_wider()`  | Convert data from long to wide format |

Examples:

### A single time-varying variable

Restructure long to wide. Every `id` becomes one row, and the values of `year` become new column names, prefixed with `inc_`:

```{r, eval=FALSE}
data_wide <- data_long |>
  pivot_wider(
    id_cols = c(id, male),
    names_from = year,
    names_prefix = "inc_",
    values_from = inc
  )
```

Restructure wide to long. The reverse operation recovers `year` from the column names, and `inc` from the values stored in those columns:

```{r, eval=FALSE}
data_long <- data_wide |>
  pivot_longer(
    cols = -c(id, male),
    names_prefix = "inc_",
    names_to = "year",
    values_to = "inc"
  )
```

### Several time-varying variables

Things get a little more involved when more than one variable is time-varying, for instance if we also measure a person's status `stat` every year:

| id  | year | inc | stat | male |
|-----|------|-----|------|------|
| 1   | 2000 | 10  | 3    | 0    |
| 1   | 2001 | 12  | 4    | 0    |
| 1   | 2002 | 13  | 5    | 0    |
| 2   | 2000 | 14  | 2    | 1    |
| 2   | 2001 | 14  | 2    | 1    |
| 2   | 2002 | 13  | 2    | 1    |
| 3   | 2000 | 20  | 1    | 0    |
| 3   | 2002 | 22  | 3    | 0    |

or in wide format:

| id  | inc_2000 | inc_2001 | inc_2002 | stat_2000 | stat_2001 | stat_2002 | male |
|-----|---------|---------|---------|----------|----------|----------|------|
| 1   | 10      | 12      | 13      | 3        | 4        | 5        | 0    |
| 2   | 14      | 14      | 13      | 2        | 2        | 2        | 1    |
| 3   | 20      | NA      | 22      | 1        | NA       | 3        | 0    |

Now a single column name, e.g. `inc_2000`, has to be split into two pieces of information: which variable (`inc` or `stat`) and which year (`2000`). `pivot_wider()` handles this by pivoting on more than one variable at once:

```{r, eval=FALSE}
data_wide <- data_long |>
  pivot_wider(
    id_cols = c(id, male),
    names_from = year,
    values_from = c(inc, stat)
  )
```

Going back to long format, we tell `pivot_longer()` that one part of the name (`.value`) becomes the new variable name, `inc` or `stat`, and the other part becomes the value of `year`:

```{r, eval=FALSE}
data_long <- data_wide |>
  pivot_longer(
    cols = -c(id, male),
    names_sep = "_",
    names_to = c(".value", "year")
  )
```

Note that `pivot_longer()` and `pivot_wider()` expect that variable names follow a consistent naming convention. In the examples above, the variables are named using a prefix, `inc` and `stat`, followed by a year, `2000`, `2001`, and `2002`, corresponding to the level of an index variable, here the year in which income and status were measured, and `names_sep = "_"` is enough to tell R where the prefix ends and the year begins. If variables are not named consistently, you may need to rename them before restructuring the data.

Sometimes a single separator is not enough, for example when a name mixes a prefix, a single-letter sex code, and a two-digit wave without any separator at all, like `inc_m01` or `lit_f02`. In such cases `names_sep` cannot locate the boundaries, and `pivot_longer()`/`pivot_wider()` instead need a *regular expression* (via the `names_pattern` argument) to describe the naming pattern. Regular expressions are a powerful but separate topic that we do not cover in this course; for now, it is enough to know that they exist for naming conventions that `names_sep` cannot handle.

## Assignments


### 1. Longitudinal data in long and wide format

The dataset `gpa2long.dta` contains longitudinal data, taken from the book on multilevel analysis that we use in SaSR7, in long format.

- `student` student identifier
- `occas` measurement occasion, 0 = year 1 semester 1, 1 = year 1 semester 2, 2 = year 2 semester 1, 3 = year 2 semester 2, ...
- `sex` 0 = male, 1 = female
- `gpa` grade-point average
- `job` paid work hours per week
- `highgpa` high school GPA
- `admitted` admitted to university, 0 = no, 1 = yes, or missing

a. Which variables are time-constant and which are time-varying?
b. How many observations are there for student 1 and student 2?
c. Use `pivot_wider()` to convert the data to wide format using `occas` as the measurement occasion.  Save it as a new object called `gpa2wide`.
d. Which new variables are created?
e. How many observations does the wide dataset contain?
f. Convince yourself that information was preserved across the restructuring and that only the representation of the information changed.
g. Use `pivot_longer()` to convert `gpa2wide` it back to long format.

### 2. Sex-specific country indicators

In this assignment, we use the `World95` dataset. This dataset contains information for 109 countries. One observation corresponds to one country.

Some variables are sex-specific, e.g., the life expectancy for men and women (`lifeexpm` and `lifeexpf` respectively) and the literacy rates of men and women (`lit_male` and `lit_fema` respectively). Other variables, e.g., `populatn` and `density`, are not sex-specific.

We want to create a dataset in which each country is represented by two observations, one describing the men and one describing the women of that country.

A new dummy variable `female` should describe whether a record pertains to women or men.

Thus, a new dataset has to be constructed with `2 × 109 = 218` observations, with at least the following variables:

- `country`
- `female`
- `population`
- `density`
- `lifeexp`
- `lit`

where `population` and `density` do not vary with gender, while `lifeexp` and `lit` do.

- Which variables are gender-specific and which are not?
- Rename variables if necessary so that they follow a consistent naming convention.
- Use `pivot_longer()` to create the desired dataset.
- Create a variable `female`.
- Verify that the resulting dataset contains 218 observations and 109 countries.

### 3. Multilevel data in four formats

The synthetic dataset `level4.dta` contains multilevel data on three sex-role attitude measurements over time, 2002, 2004, and 2006, for both spouses of heterosexual couples.

- `hhid` household id
- `iid` individual id
- `urban` urbanized
- `male`
- `edu` highest completed education, 1 to 10
- `year`
- `att1` family value: man natural breadwinner
- `att2` family value: women are better with children
- `att3` family value: men and women equal decision makers

We say that the dataset `level4` is in long-long-long-wide format for the key variables `hhid`, `iid`, `year`, and  (not-existing) `item`. The first term *long* refers to households, and reflects that households are stored in different observations identified by `hhid`. The second term *long* refers to individuals, and reflects that partners are stored as separate observations identified by `iid`; the variable `hhid` allows linking people who form a couple. The third long refers to time: measurements at different time points are stored in different observations, with time indicated by the variable `year`. The final term wide specifies that the repeated attitude measurements are stored in different variables within an observation, not as different observations.

Make sure that you understand the data well.

a.  Use `pivot_longer()` to convert the data from long-long-long-wide to long-long-long-long format, i.e., the items are stored in different observations rather than in different variables indexed by a new variable `item`.
b. Use `pivot_wider()` to convert the data from the original long-long-long-wide format to long-long-wide-wide format. Explain what an observation in this dataset represents.
c. Use `pivot_wider()` to convert the data from the original long-long-long-wide format to long-wide-wide-wide format, identifying individuals within households by `male` rather than by `iid`. What happens if you identify people within households by `iid`? Is using `male` as a person-within-household identifier correct if the data include same-sex households?

::: {.content-visible when-format="html"}

## Answers

::: {.callout-note appearance="simple" collapse=true}
## See answers

```{r setup-restructure}
library(tidyverse)
library(haven)

gpa2long <- read_dta("data/gpa2long.dta")
world95 <- read_dta("data/world95.dta")
level4 <- read_dta("data/level4.dta")
```

### 1. Longitudinal data in long and wide format

Have a look at the data before restructuring.

```{r}
gpa2long
```

Which variables are time-constant and which are time-varying? You can answer this by thinking about the meaning of the variables and looking informally at the data. 

- student: a person identifier, so should be time-constant
- occas: the time index, so it is by definition time-varying
- sex: time-constant
- gpa: time-varying, can change each semester
- job: time-varying: working hours change between semesters
- highgpa: time-constant, as it's the final high school GPA
- admitted: time-constant, there is a single admission decision per student

But to be sure we can check the number of distinct values of each variable within each student. If a variable has more than one distinct value for a student, it is time-varying; if it has only one distinct value, it is time-constant.

```{r}
library(skimr)

gpa2long |>
  # Compute the number of dsitinct values of each variable *within each student*
  # You learned about the .by argument in "Selecting cases" and will practice it more in "Aggregating data")
  # (Tip: have a look at the results of this mutate() step before running the summarise step to see what is going on)
  mutate(
    n_occas    = n_distinct(occas),
    n_sex      = n_distinct(sex),
    n_gpa      = n_distinct(gpa),
    n_job      = n_distinct(job),
    n_highgpa  = n_distinct(highgpa),
    n_admitted = n_distinct(admitted),
    .by = student
  ) |> 
  # Now let's look at the summary stats for those counts
  skim(starts_with("n_")) 
  # Time-constant variables should have a minimum and maximum of 1 (i.e., they have only one value per student)
```


How many observations are there for student 1 and student 2?

```{r}
gpa2long |>
  filter(student %in% c(1, 2)) |>
  count(student) # could also use tabyl()
```


Convert to wide format.

```{r}
gpa2wide <- gpa2long |>
  pivot_wider(
    id_cols = c(student, sex, highgpa, admitted),
    values_from = c(gpa, job),
    names_from = occas
  )
# id_cols lists everything that identifies a student and does not vary over time
# values_from lists the time-varying variables
# names_from indicates that occas should be used to distinguish the new variable names
```

Which new variables are created?

```{r}
names(gpa2wide)
# each time-varying variable is now split into one column per occasion
```

How many observations does the wide dataset contain?

```{r}
nrow(gpa2wide)
n_distinct(gpa2long$student)
# the wide dataset has exactly one row per student
```

Was information preserved? Explore it yourself. 

Convert back to long format:
```{r}
gpa2long_again <- gpa2wide |>
  pivot_longer(
    cols = -c(student, sex, highgpa, admitted),
    names_sep = "_",
    names_to = c(".value", "occas")
  ) 
# cols indicates which columns we want to make long. We want all time-varying
# variables. Note the the minus sign, we are saying "everything minus these variables here"
# names_sep indicates that the names of the variables we want to make longer contain more than one piece of information, separated by "_" (job_0, gpa_3, etc).
# ".value" tells pivot_longer to put the value of the old variable value in new a variable that matches the first piece of the name. The second piece should be put in a variable called "occas"
# E.g.
# job_0 -> the first piece is "job", so the value of "job_0" goes to a new variable called "job". The second piece is "0", so "0" goes to a new variable called "occas"
```

Compare with the original.

```{r}
gpa2long_again |>
  arrange(student, occas) |>
  select(student, occas, gpa, job) |>
  slice(1:10)

gpa2long |>
  arrange(student, occas) |>
  select(student, occas, gpa, job) |>
  slice(1:10)

# Note that in gpa2long_again, "occas" was created as a string (because we
# got the 0 to 5 values from variable names, which are strings). We can correct
# that:
gpa2long_again |> 
  mutate(occas = as.numeric(occas))
# We still lose the labels that came with the original long dataset, but that 
# does not affect results
```


### 2. Sex-specific country indicators

Inspect the variable names.

```{r}
names(world95)
# look for the sex-specific pairs
```

```{r}
# gender-specific: lifeexpm and lifeexpf, lit_male and lit_fema
# not gender-specific: country, populatn, density, and the remaining variables
#
# The two pairs do not follow the same naming convention: one uses a suffix
# without a separator, the other an underscore and a full word. pivot_longer()
# needs a single consistent pattern, so we rename first. Using a separator like 
# "_" is always a goog idea, because it makes pivoting easier.
```

Rename to a consistent convention.

```{r}
world95_renamed <- world95 |>
  rename(
    lifeexp_male = lifeexpm,
    lifeexp_female = lifeexpf,
    lit_male = lit_male,
    lit_female = lit_fema,
    population = populatn
  )

# check
world95_renamed |>
  select(country, population, density, starts_with("lifeexp"), starts_with("lit")) |>
  slice(1:5)
```

Restructure to two rows per country.

```{r}
world95_long <- world95_renamed |>
  select(country, population, density, lifeexp_male, lifeexp_female, lit_male, lit_female) |>
  pivot_longer(
    cols = c(lifeexp_male, lifeexp_female, lit_male, lit_female),
    names_to = c(".value", "sex"),
    names_sep = "_"
  )
```

Create the dummy variable `female`.

```{r}
world95_long <- world95_long |>
  mutate(
    female = if_else(sex == "female", 1, 0)
  )
# female = 1 for the female record, 0 for the male record
```

Verify the number of observations.

```{r}
nrow(world95_long)
n_distinct(world95_long$country)
```


### 3. Multilevel data in four formats

Inspect the data.

```{r}
level4
```

```{r}
level4 |>
  arrange(hhid, iid, year) |>
  slice(1:12)
# two people per household, three years per person, three items per row
```


#### a. Long-long-long-long

```{r}
level4_llll <- level4 |>
  pivot_longer(
    cols = c(att1, att2, att3),
    names_to = "item",
    values_to = "att"
  )
# each attitude item now occupies its own row, indexed by item
```

```{r}
nrow(level4)
nrow(level4_llll)
# the number of rows is multiplied by three, one row per item
```

```{r}
level4_llll |>
  arrange(hhid, iid, year, item) |>
  slice(1:12)
# an observation is now one item, for one person, in one year
```

#### b. Long-long-wide-wide

```{r}
level4_llww <- level4 |>
  pivot_wider(
    id_cols = c(hhid, iid, urban, male, edu),
    names_from = year,
    values_from = c(att1, att2, att3)
  )
# the time dimension moves into the variable names
```

```{r}
names(level4_llww)
nrow(level4_llww)
# nine attitude variables: three items times three years
# one row per individual
```

What does an observation represent? One person.


#### c. Long-wide-wide-wide

```{r}
level4_lwww <- level4 |>
  pivot_wider(
    id_cols = c(hhid, urban),
    names_from = c(male, year),
    values_from = c(att1, att2, att3, edu),
  )
# both the person and the time dimension move into the variable names
# e.g. att1_1_2002 means att1 from the person with male=1 and year=2002
# (we could use the argument name_preffix to get more clear names like 
# "att1_male1_2002")
```

We now have one row per household.
```{r}
nrow(level4_lwww)
n_distinct(level4$hhid)
# one row per household

names(level4_lwww)
# each variable is now indexed by male and year
```

What would happen if we used `iid` instead of `male`? iid is unique across the whole dataset, not just within a household. Using it as the person-within-household index would create one pair of columns for every individual in the data, so the result would have as many column groups as there are people, almost all of them missing for any given
household. The dataset would be enormous and unusable.


```{r}
# Using iid
level4 |>
  pivot_wider(
    id_cols = c(hhid, urban),
    names_from = c(iid, year),
    values_from = c(att1, att2, att3, edu),
  )
```


Is male a correct person-within-household identifier? Only for heterosexual couples, which is what this synthetic dataset contains. In a same-sex household both partners share the same value of male, so the pair no longer identifies them uniquely. pivot_wider() wouldfind two values for one cell and return a list column with a warning about
duplicate identifiers, rather than the intend two sets of columns.

One solution would be creating a gender-neutral unique identifier within each houseold
(partner 1, partner 2), and use that instead: 

```{r}
level4 |>
  mutate(
    partner = dense_rank(iid),
    .by = hhid
  ) |> 
  pivot_wider(
    id_cols = c(hhid, urban),
    names_from = c(partner, year),
    values_from = c(att1, att2, att3, edu),
  ) 
```

Now `att1_1_2002` means att1 for partner 1 (i.e. the first row within the household) in 2004. Note that in this case we lost the information the who is male (parter 1 or 2).

:::
:::