---
title:  Getting to know and summarizing data
---

In this part you will learn about the following basic tasks:

- Load a dataset from disk into computer memory.

- Explore the variables in the data.

- List values for (some) variables.

- Sort the observations on variables.

- The concept of “key variable(s)”.

- Create frequency distributions of variables.

- Create cross tabulations of multiple variables.

- Produce a table of summary statistics (e.g., means, sd = standard deviation) for a series of variables.

- Produce a table of correlations of quantitative variables.

The concept of key variables may be new to you. A key consists of one or more variables that uniquely identify observations.

In social science datasets, we often use single observations to represent all variables measured on subjects. Each person is then a row in the data matrix. Such datasets often contain a respondent number, say `id`, usually integer valued. The values of `id` are usually unique, and then `id` qualifies as a key. If you know a subject's value of `id`, you can find the corresponding row in the data matrix, and hence know all the subject's characteristics.

If a value of `id` occurs more than once, obviously, the value of `id` does not identify a unique subject, and `id` is not a key. Observe that `id` need not take all possible values. For instance, if your `id` consists of 9 digits, likely your dataset does not contain all possible numbers of 9 digits. Many possible values of `id` will not be used. That is not a problem. `id` would still be a key if all values are distinct.

If `id` takes duplicate values, we may form a key by considering additional variables. For instance, think of a household survey with one or two adults. Observations are people nested in households. The variable `hid`, household number, would not be a key. The combination of `hid` and `sex` might be a key if the sample comprises single people and different-sex couples, but not if the sample also includes same-sex couples, cohabiting siblings, etc. In the latter case, the data may contain a variable indicating whether a person is the oldest adult in the household, and together with `hid` this variable may form a key. If you only have birth dates, the combination of `hid` and birth date is usually a key, but you could have bad luck if your data include a household with two cohabiting twin sisters.

If a dataset does not contain a case identifier such as a respondent number, we strongly suggest you create one. Be sure to do this in a reproducible way. First put the observations in a specific order, by sorting on a series of variables that uniquely identify cases. In the example below, we assume that each observation has a unique pattern on the variables `city`, `age`, `edu`, and `sex`. Obviously, you should verify this assumption.

In R, you can check duplicates and create an identifier as follows:

```{.r}      
df |> 
  count(city, age, edu, sex) |> 
  filter(n > 1)  

df <- df |>   
  arrange(city, age, edu, sex) |>   
  mutate(respnr = row_number()) 
```

After sorting the cases, a new variable is created with the value 1 in the first observation, 2 in the second observation, 3 in the third observation, etc. This is one of the occasions in which you may want to save the dataset, but we urge you not to overwrite the original file.

## R

You will learn the R functions listed below to describe and summarize data. Read the corresponding documentation pages when needed. Start with the description and examples. We do not expect you to learn all functions by heart, but you should get a rough idea of, and remember, what they can accomplish.

| Function                      | Description                                   |
|-------------------------------|-----------------------------------------------|
| `read_csv()`                  | Load a CSV file into R                        |
| `read_dta()` / `read_sav()`   | Load a Stata / SPSS file into R               |
| `read_excel()`                | Load an Excel file into R                     |
| `glimpse()`                   | Describe data contents                        |
| `skim()`                      | Describe data contents and summary statistics |
| `View()`                      | View observations and variables               |
| `arrange()`                   | Sort observations                             |
| `table()`                     | One-way and two-way tables of frequencies     |
| `prop.table()`                | Convert frequencies to proportions            |
| `tabyl()`                     | Frequency tables with percentages             |
| `chisq.test()`                | Pearson's chi-square test for independence    |
| `summarise()`                 | Summary statistics                            |
| `cor()`                       | Correlations (Pearson by default)             |
| `count()`                     | Count observations by category or group       |
| `select()`                    | Select variables                              |
| `filter()`                    | Select observations satisfying a condition    |
| `slice()`                     | Select observations by row number             |

Examples:

Load a file into memory, using a relative path:

```{r, eval=FALSE}
library(readr) # for opening CSV files 
data <- read_csv("data/mydata.csv")  

library(haven) # for opening Stata (.dta) and SPSS (.sav) files 
data <- read_dta("data/mydata.dta")  
data <- read_sav("data/mydata.sav") 

library(readxl) # for opening Excel (.xlsx) files 
data <- read_excel("data/mydata.xlsx") 
```


Describe variables in the active dataset:

```{r, eval=FALSE}
glimpse(data) # shows basic info about the dataset and 
              # a overview of all variables

data |>   
  select(edu, exp) |>   
  glimpse() # compact overview of edu and exp  

data |>   
  select(starts_with("rat")) |>   
  glimpse() # describe all variables starting with rat 

skim(data) # summary statistics  (from package 'skimr')
View(data) # open data viewer in RStudio 
```

List observations:

```{r, eval=FALSE}
data # all variables and all observations  

data |>   
  select(edu, inc) # selection of variables  

data |>   
  filter(country == 4) # selection of observations 
                       # (all rows in which variable 'country' equals 4)

data |>   
  select(country, edu, inc) |>    # selection of variables
  slice(1:20)                     # selection of observations 1 to 20 
```

Sort observations on one or more variables:

```{r, eval=FALSE}
data |>   
  arrange(edu) # sort by edu (increasing)

data |>   
  arrange(edu, inc) # sort by edu and inc  (increasing)

data |>   
  arrange(desc(edu)) # sort by edu (decreasing)
```

One-way tabulation, frequencies:

```{r, eval=FALSE}
table(data$class) # one-way table  

table(data$edu, useNA = "ifany") # one-way table, include missing values  

# usign tabyl, from package 'janitor'
data |> 
  tabyl(edu) # frequency table with missing values and percentages
             
```

Two-way tabulation of variables:

```{r, eval=FALSE}
table(data$edu, data$class)

# using tabyl:
data |> 
  tabyl(edu, class)
```

Row and column percentages:

```{r, eval=FALSE}
# row percentages:
table(data$edu, data$class) |> 
  prop.table(margin = 1)
# column percentages:
table(data$edu, data$class) |> 
  prop.table(margin = 2)

# using tabyl:
data |> 
  tabyl(edu, class) |> 
  adorn_percentages("row")

data |> 
  tabyl(edu, class) |> 
  adorn_percentages("columns") |> 
  adorn_pct_formatting() # optional
```

In `prop.table`, the argument `margin = 1` computes proportions within rows, while `margin = 2` computes proportions within columns.

Pearson's chi-square test for independence:

```{r, eval=FALSE}
table(data$edu, data$class) |> 
  chisq.test()
# Pearson's chi-square test for independence
```

The chi-square test checks whether two categorical variables are statistically independent. 

Tabulation of three or more variables:

```{r, eval=FALSE}
# three variables 
xtabs(~ edu + class + gender, data = data, addNA=TRUE) 

data |> 
  tabyl(edu, class, gender)

# four variables (not possible with tabyl)
xtabs(~ edu + class + gender + country, data = data, addNA=TRUE) 

# It is also possible to obtain frequencies using count(). 
# But it is not printed as a table and it does not show missing combinations 
data |>   
  count(gender, edu, class)

data |>   
  count(country, gender, edu, class)  

```

Summary statistics:

```{r, eval=FALSE}
# Summary statistics for income 
data |> 
  skim(income)

data |>   
  summarise(     
    mean_income = mean(income, na.rm = TRUE),     
    sd_income = sd(income, na.rm = TRUE),     
    min_income = min(income, na.rm = TRUE),     
    max_income = max(income, na.rm = TRUE)   
    ) 

```

Correlation matrix of two or more variables:

```{r, eval=FALSE}
data |>   
  select(edu, inc, age) |>   
  cor(use = "complete.obs") # Pearson correlations 
```

Count the number of observations meeting a condition (you will practice this again in [Selecting cases](dh_select.qmd)):

```{r, eval=FALSE}
# count observations with income above 50000:
data |>   
  filter(income > 50000) |>   
  nrow() 

# count observations with income between 5000 and 50000, boundaries excluded: 
data |>   
  filter(income > 5000, income < 50000) |>   
  nrow()

# count observations with income between 5000 and 50000, boundaries included:
data |>   
  filter(income >= 5000, income <= 50000) |> # alternative: filter(between(income, 5000, 50000))  
  nrow() 

# count observations with income above 50000 and income not missing:
data |>   
  filter(income > 50000, !is.na(income)) |>   
  nrow() 
```

Check that the variable `birthd` is a key:

```{r, eval=FALSE}
data |>   
  summarise(no_missing = all(!is.na(birthd))) # birthd is never missing  

data |>   
  count(birthd) |>   
  filter(n > 1) # check duplicates in birthd 
```

Check that `birthd` and `sex` form a composite key:

```{r, eval=FALSE}
data |>   
  summarise(no_missing = all(!is.na(birthd), !is.na(sex))) # birthd and sex are never missing  

data |>   
  count(birthd, sex) |>   
  filter(n > 1) # check duplicates in the combination of birthd and sex 
```

Create a key:

```{r, eval=FALSE}
data <- data |>   
  arrange(birthd, sex) |>   
  mutate(id = row_number()) # create id after sorting by birthd and sex
```

## Assignments

In these assignments we use the dataset `sbs399.dta`. 

1. How many observations and variables the dataset contains?

2. What is the variable label of `feduc`?

3. Display a frequency distribution of `educ`. What is the percentage of subjects with an education up to category 4? How many missing values are there in `educ`?

4. Select the variables `hrinc` and `pres`. Adjust the syntax to show these variables only for men among the initial 40 observations. Further adjust the syntax to restrict the sample to men with below median education (see previous question) among the initial 40 observations. Finally, show the variables `hrinc` and `pres` for men and women among the initial 5 observations and arrange the results by sex.

5. List values of `age` in increasing order. Next, list just the five smallest values, and list just the five largest values. Do not read these values from the data viewer. Give syntax to display the requested cases. 

6. Create a two-way table of `feduc` rows and `meduc` columns. Add row and column percentages. Test for independence of `feduc` and `meduc` using Pearson's chi-square test.

7. What are the mean, standard deviation, median, minimum, and maximum value of `feduc`?

8.  Verify that the number of subjects for whom the mother is higher educated than the father is 543. What is the count if you restrict to female subjects?

9.  What are the Pearson correlations of `educ`, `feduc`, and `meduc`? Use the help system to find out how to compute Spearman's rank correlations.

10. Is it true that for any variable the values are above average for 50% of cases? If you are uncertain, check this with some variables. For an explanation, think about this example: what percentage of people have more than the average number of legs? 50%? Who are these people?

11. Is `hrespnr` a key, meaning that it is never missing and all values are unique? Why? Do not rely on the data viewer. Write syntax to obtain the information on which you base your answer. What about the pair of variables `hrespnr` and `sex`? Why? What about `hrespnr`, `sex`, and `age`, and `hrespnr`, `sex`, and `fage`?

::: {.content-visible when-format="html"}

## Answers

::: {.callout-note appearance="simple" collapse=true}
## See answers


```{r setup-summary}
library(tidyverse)
library(haven)
library(janitor)
library(skimr)

# The file sbs399.dta should be available in your working directory.

sbs399 <- read_dta("data/sbs399.dta")
```

### 1
How many observations and variables the dataset contains?
```{r}
glimpse(sbs399)
# 3,350 observations
# 50 variables
```


### 2

What is the variable label of `feduc`?
```{r}
glimpse(sbs399$feduc)
#fathers highest education
```

### 3

Display a frequency distribution of `educ`. What is the percentage of subjects with an education up to category 4? How many missing values are there in `educ`?

```{r}
# frequency distribution of educ, including missing values:
table(sbs399$educ, useNA = "ifany")

sbs399 |>
  tabyl(educ)
# frequency table with percentages

# There are (0.07373134 + 0.18208955 + 0.15044776 + 0.06686567) = 0.4731343 = 47.31% of subjects with education up to category 4.
# Missings = 0

```

### 4

Select the variables `hrinc` and `pres`. Adjust the syntax to show these variables only for men among the initial 40 observations. Further adjust the syntax to restrict the sample to men with below median education (see previous question) among the initial 40 observations. Finally, show the variables `hrinc` and `pres` for men and women among the initial 5 observations and arrange the results by sex

```{r}
sbs399 |>
  select(hrinc, pres)
```

Only men among the initial 40 observations:

```{r}
sbs399 |>
  slice(1:40) |>
  filter(sex == 0) |>
  select(hrinc, pres)
```

Men with below median education among the initial 40 observations:

```{r}
sbs399 |>
  slice(1:40) |>
  filter(
    sex == 0,
    educ < 5 # median of educ is 5, see response to question 3
  ) |>
  select(hrinc, pres, educ, sex)
```

Men and women among the initial 5 observations:

```{r}
sbs399 |>
  slice(1:5) |>
  select(sex, hrinc, pres) |> 
  arrange(sex)
```

### 5

List values of `age` in increasing order. Next, list just the five smallest values, and list just the five largest values. Do not read these values from the data viewer. Give syntax to display the requested cases. 

Values of `age` in increasing order:

```{r}
sbs399 |>
  arrange(age) |>
  select(age)
```

Five smallest values:

```{r}
sbs399 |>
  arrange(age) |>
  slice(1:5) |>
  select(age)
```

Five largest values:

```{r}
sbs399 |>
  arrange(desc(age)) |>
  slice(1:5) |>
  select(age)
```


### 6

Create a two-way table of `feduc` rows and `meduc` columns. Add row and column percentages. Test for independence of `feduc`and `meduc` using Pearson’s chi-square test.

Frequency table:

```{r}
tab_feduc_meduc <- table(sbs399$feduc, sbs399$meduc)
tab_feduc_meduc
```

Row percentages:

```{r}
prop.table(tab_feduc_meduc, margin = 1) * 100
```

Column percentages:

```{r}
prop.table(tab_feduc_meduc, margin = 2) * 100
```

Pearson's chi-square test:

```{r}
chisq.test(tab_feduc_meduc)
```

### 7
 What are the mean, standard deviation, median, minimum, and maximum value of `feduc`? 

```{r}
sbs399 |>
  summarise(
    mean = mean(feduc, na.rm = TRUE),
    sd = sd(feduc, na.rm = TRUE),
    min = min(feduc, na.rm = TRUE),
    median = median(feduc, na.rm = TRUE),
    max = max(feduc, na.rm = TRUE)
  )

# Mean = 3.04
# Standard deviation = 2.20
# Minimum = 1 (elementary [lower])
# Median = 2
# Maximum = 8 (university [wo])
```

### 8

Verify that the number of subjects for whom the mother is higher educated than the father is 543. What is the count if you restrict to female subjects?

```{r}
sbs399 |>
  filter(meduc > feduc) |>
  nrow()
```

Restricted to female subjects:

```{r}
sbs399 |>
  filter(
    meduc > feduc,
    sex == 1
  ) |>
  nrow()

# Count restricted to female subjects = 262
```

### 9

What are the Pearson correlations of `educ`, `feduc`, and `meduc`? Use the help system to find out how to compute Spearman’s rank correlations.

Pearson correlations:

```{r}
sbs399 |>
  select(educ, feduc, meduc) |>
  cor(use = "complete.obs", method = "pearson")
```

Spearman rank correlations:

```{r}
sbs399 |>
  select(educ, feduc, meduc) |>
  cor(use = "complete.obs", method = "spearman")
```

### 10

Is it true that for any variable the values are above average for 50% of cases? If you are uncertain, check this with some variables. For an explanation, think about this example: what percentage of people have more than the average number of legs? 50%? Who are these people?

No answer provided. Think yourself, using the hint.


### 11

Is `hrespnr` a key, meaning that it is never missing and all values are unique? Why? Do not rely on the data viewer. Write syntax to obtain the information on which you base your answer. What about the pair of variables `hrespnr` and `sex`? Why? What about`hrespnr`, `sex`, and `age`, and `hrespnr`, `sex`, and `fage`?

Check whether `hrespnr` is never missing:

```{r}
sbs399 |>
  summarise(no_missing_hrespnr = all(!is.na(hrespnr)))
```

Check whether `hrespnr` is unique:

```{r}
sbs399 |>
  count(hrespnr) |>
  filter(n > 1)
```

`hrespnr` is never missing, but it is not unique (i.e. there are duplicates).

Check whether `hrespnr` and `sex` form a composite key:
```{r}
sbs399 |>
  summarise(no_missing = all(!is.na(hrespnr), !is.na(sex)))

sbs399 |>
  count(hrespnr, sex) |>
  filter(n > 1)
```

Again, never missing, but not unique.

Check whether `hrespnr`, `sex`, and `age` form a composite key:
```{r}
sbs399 |>
  summarise(no_missing = all(!is.na(hrespnr), !is.na(sex), !is.na(age)))

sbs399 |>
  count(hrespnr, sex, age) |>
  filter(n > 1)
```

Check whether `hrespnr`, `sex`, and `fage` form a composite key:

```{r}
sbs399 |>
  summarise(no_missing = all(!is.na(hrespnr), !is.na(sex), !is.na(fage)))

sbs399 |>
  count(hrespnr, sex, fage) |>
  filter(n > 1)
```

We conclude that  `hrespnr`, `sex`, and `fage` form a composite key. 
:::
:::