---
title: Selecting cases
---

In this part you will learn about selecting cases (observations).


| Function       | Description                                 |
|----------------|---------------------------------------------|
| `filter()`     | Select observations satisfying a condition  |
| `slice()`      | Select observations by row number           |
| `slice_head()` | Select the first observations               |
| `slice_tail()` | Select the last observations                |
| `group_by()`   | Define groups for subsequent operations     |
| `.by`          | Apply an operation separately within groups |

## R

Examples:

Select observations using logical conditions:

```{r, eval=FALSE}
data |>   
  filter(sex == 1)
# select observations for which sex equals 1

data |>   
  filter(sex == 1, edu > 3)
# select observations for which sex equals 1 AND edu is larger than 3

data |>   
  filter(sex == 1 | edu >= 3)
# select observations for which sex equals 1 OR edu is at least 3
# the vertical bar means OR, so respondents satisfying either condition are counted


data |> 
  filter(edu != 3)
# select observations in which edu is different than 3


data |> 
  filter((sex == 1 & edu > 3) | (sex == 0 & edu < 3))
# select observations in which one of this is true:
# 1. sex equals 1 and edu is larger than 3
# OR
# 2. sex equals 0 and education is lower than 3

data |> 
  filter(age == max(age))
# select observation(s) whose age equals the maximum age in the dataset, i.e. the 
# oldest person(s). Should use na.rm=TRUE is there are missing values in age
```

Select observations based on row number:
```{r, eval=FALSE}
data |> 
   slice(5) 
# select 5th row

data |>   
  slice(1:20) 
# select rows 1 to 20

data |>   
  slice_head(20)
# select first 20 observations (same as above)

data |>   
  slice_tail(20)
# select the last 20 observations  
```

The same can also be done with `filter()` and `row_number()`:
```{r, eval=FALSE}
data |>   
  filter(row_number() == 5) 
# select 5th row 

data |>   
  filter(row_number() <= 20)  
# select the first 20 observations

data |>   
  filter(row_number() > (n() - 20))
# select the last 20 observations 
# n() gives the total number of rows
```

Keep a subset of observations:

```{r, eval=FALSE}
data_male <- data |>   
  filter(male == 1)  
# create a new dataset containing only observations for which male==1

data_first100 <- data |>   
  slice(1:100)
# create a new dataset containing only the first 100 observations
```

The argument `.by` is to apply a filter within groups. This is useful when you want to compare an observation with other observations in each group, and filter based on that. You will learn more about grouped operations in [Aggregating data](dh_aggregate.qmd). Suppose each observation belongs to a household, indicated by the id variable `hh_id`:

```{r, eval=FALSE}
data |>   
  filter(row_number() == 1, .by = hh_id)  
# select the first row from *each household*. It is as if we create one dataset per 
# household, selected the first row from each of them, them put them together again

data |>   
  filter(age == max(age), .by = hh_id)  
# select observations where age equals the maximum age *within each household*, i.e. 
# the oldest in each household. Note that this can be more than one person per household
# if they have the same age.
``` 


## Assignments

In these assignments, you should use the dataset `sbs399`. 

1.  Using `filter()` and `nrow()`, count the number of observations in `sbs399`:

    - with age at most 30;

    - with age at least 20;

    - with age between 20 and 30, boundaries included;

    - with age between 20 and 30, boundaries excluded;

    - males over 25;

    - people who are either below 68 or female.

2.  Select the subjects with age at most 30 and store this subset in a new object. Do not overwrite the original dataset `sbs399`.

    - Count the number of subjects of age at least 20 in this subset.

    - Count how many subjects of age at least 20 in this subset are men and how many are women.

3.  Using the full dataset `sbs399` again.

    - Create a selection variable, say `touse`, that takes the value 1 for males over 30 and 0 otherwise.

    - For the selected sample:

      - create a two-way table for the respondent's education, `educ`, and the respondent's father's education, `feduc`.

      - compute the correlation between `educ` and `feduc`

4.  What is the difference between the following R commands?

```r        
data |>   
  select(-sex)  

data |>   
  filter(sex != 1)  

data |>   
  filter(sex == 1) 
```

5.  Using the household identifier `hrespnr` in the dataset `sbs399`, use `.by` or `group_by()` to:

    - Select the oldest people in households.
      - Bonus: try to figure out in how many households there are "ties", i.e. more than one person has the same maximum age.  

    - List `sex` and `age` in increasing order of age in households where the age difference between the oldest and the youngets person is larger than 5, without creating new variables;

    - Drop people living alone in a household. How many cases remain in the dataset?

## Answers

```{r setup-select}
library(tidyverse)
library(haven)
library(janitor)

sbs399 <- read_dta("data/sbs399.dta")
```

Check the coding of `sex` before using it in any condition.

```{r}
print_labels(sbs399$sex)
# sex = 0 for men and sex = 1 for women
```

### 1. Counting observations with `filter()` and `nrow()`

Age at most 30.

```{r}
sbs399 |>
  filter(age <= 30) |>
  nrow()
# number of respondents aged 30 or younger
```

Age at least 20.

```{r}
sbs399 |>
  filter(age >= 20) |>
  nrow()
# number of respondents aged 20 or older
```

Age between 20 and 30, boundaries included.

```{r}
sbs399 |>
  filter(age >= 20, age <= 30) |>
  nrow()
# between(age, 20, 30) would give the same result
```

Age between 20 and 30, boundaries excluded.

```{r}
sbs399 |>
  filter(age > 20, age < 30) |>
  nrow()
# respondents of exactly 20 or exactly 30 are now excluded
```

Males over 25.

```{r}
sbs399 |>
  filter(sex == 0, age > 25) |>
  nrow()
# men older than 25
```

People who are either below 68 or female.

```{r}
sbs399 |>
  filter(age < 68 | sex == 1) |>
  nrow()
```

Note that `filter()` drops rows for which the condition evaluates to `NA`. In this dataset `age` and `sex` are complete, so nothing is lost here, but with incomplete data the counts above would exclude those respondents.

### 2. Working with a subset

Select the subjects with age at most 30 and store them in a new object.

```{r}
sbs399_young <- sbs399 |>
  filter(age <= 30)
# the original dataset sbs399 is left untouched
```

Count the number of subjects of age at least 20 in this subset.

```{r}
sbs399_young |>
  filter(age >= 20) |>
  nrow()
# respondents aged between 20 and 30 in the subset
```

The same count separately for men and women.

```{r}
sbs399_young |>
  filter(age >= 20) |>
  tabyl(sex) # could also be count(sex), as well as other solutions
# counts by sex using .by
```


### 3. A selection variable

Create the selection variable `touse`.

```{r}
sbs399 <- sbs399 |>
  mutate(
    touse = if_else(sex == 0 & age > 30, 1, 0)
  )
# touse = 1 for men older than 30, 0 otherwise

# check how many cases
sbs399 |> 
  tabyl(touse)
```

Create the two-way table of education and father's education for the selected sample.

```{r}
sbs399 |>
  filter(touse == 1) |> 
  tabyl(educ, feduc)
```

Compute the correlation between `educ` and `feduc` in the selected sample.
```{r}
# Saving object with selected sample
sbs399_selected <- sbs399 |>
                    filter(touse == 1)

cor(sbs399_selected$educ, sbs399_selected$feduc)  

# Solutions without creating a new object
sbs399 |>
  filter(touse == 1) |> 
  with(cor(educ, feduc))

sbs399 |>
  filter(touse == 1) |> 
  summarise(r = cor(educ, feduc))
```



### 4. Comparing three commands

```{r}
# select(-sex) removes the COLUMN sex; all rows are kept.
# filter(sex != 1) removes ROWS for which sex equals 1; the column stays.
# filter(sex == 1) keeps only the rows for which sex equals 1.
#
# The two filter() commands look like opposites, but they are not exact
# complements in general: rows with a missing value on sex are dropped by
# both, because NA != 1 and NA == 1 both evaluate to NA rather than TRUE.
```

Demonstrate this with the number of rows and columns.

```{r}
dim(sbs399)
dim(select(sbs399, -sex))
dim(filter(sbs399, sex != 1))
dim(filter(sbs399, sex == 1))
# the first command changes the number of columns, the other two the number of rows
```

### 5. Working within households

Select the oldest people in households:
```{r}
sbs399 |>
  filter(age == max(age), .by = hrespnr) 
```

How many households with ties?
```{r}
# Using only filter()
sbs399 |>
  filter(age == max(age), .by = hrespnr) |> # select oldest
  filter(n() > 1, .by = hrespnr) |> # keep rows for households with more than one row
  filter(row_number() == 1, .by = hrespnr) |> # keep only one row per household, to have the correct count |> 
  nrow()

# Another  solution
sbs399 |>
  filter(age == max(age), .by = hrespnr) |> 
  filter(n() > 1, .by = hrespnr) |> 
  select(hrespnr) |> 
  n_distinct()
```


List `sex` and `age` in increasing order of age, in households where the age difference is larger than 5.

```{r}
sbs399 |>
  filter(max(age) - min(age) > 5, .by = hrespnr) |>
  arrange(hrespnr, age) |>
  select(hrespnr, sex, age)
```

Drop people living alone. How many respondents remain?

```{r}
sbs399 |>
  filter(n() > 1, .by = hrespnr) |> 
  nrow()
```


