---
title: Selecting cases and repeating analyses for groups
---

In this part you will learn about the following basic tasks:

- How to select cases (observations), either permanently or temporarily, for one or more commands.

- How to repeat a command (or collection of commands) for various subgroups.

| Function       | Description                                 |
|----------------|---------------------------------------------|
| `filter()`     | Select observations satisfying a condition  |
| `slice()`      | Select observations by row number           |
| `slice_head()` | Select the first observations               |
| `slice_tail()` | Select the last observations                |
| `group_by()`   | Define groups for subsequent operations     |
| `.by`          | Apply an operation separately within groups |

## R

Examples:

Select observations using logical conditions:

```{r, eval=FALSE}
data |>   
  filter(sex == 1)
# select observations for which sex equals 1

data |>   
  filter(sex == 1, edu > 3)
# select observations for which sex equals 1 and edu is larger than 3

data |>   
  filter(sex == 1 | edu >= 3)
# select observations for which sex equals 1 or edu is at least 3
```

Select observations based on row number:

Use `slice()` to select observations by their row number.

```{r, eval=FALSE}
data |>   
  slice(1:20) 
# select the first 20 observations  

data |>   
  slice((n() - 19):n()) 
# select the last 20 observations  
```

The same can also be done with `filter()` and `row_number()`:

```{r, eval=FALSE}
data |>   
  filter(row_number() <= 20)  
# select the first 20 observations

data |>   
  filter(row_number() >= n() - 19) 
# select the last 20 observations 
```

Keep a subset of observations:

```{r, eval=FALSE}
data_male <- data |>   
  filter(sex == 2)  
# create a new dataset containing only observations for which sex equals 2

data_first100 <- data |>   
  slice(1:100)
# create a new dataset containing only the first 100 observations
```

Remove observations:

```{r, eval=FALSE}
data <- data |>   
  filter(sex != 1)
# remove observations for which sex equals 1

data <- data |>   
  slice(-(1:100))
# remove the first 100 observations
```

Compute statistics within groups:

```{r, eval=FALSE}
data |>   
  summarise(     
    mean_income = mean(income, na.rm = TRUE),     
    .by = country   ) # mean income by country 
```

Compute statistics within combinations of groups:

```{r, eval=FALSE}
data |>   
  summarise(     
    mean_income = mean(income, na.rm = TRUE),     
    .by = c(country, sex)   ) # mean income by country and sex 
```

Create variables within groups:

```{r, eval=FALSE}
data <- data |>   
  mutate(     
    hinc = max(income, na.rm = TRUE),     
    .by = hrespnr   ) # highest income within household 
```

The `.by` argument is often a convenient alternative to `group_by()`. It applies the operation separately within groups without permanently changing the grouping structure of the dataset.

## Assignments

In these assignments, you should use the dataset `sbs399`. Solutions can be found in `dh_select.Rmd`.

1.  Using `filter()` and `nrow()`, count the number of observations in `sbs399`:

    - with age at most 30;

    - with age at least 20;

    - with age between 20 and 30, boundaries included;

    - with age between 20 and 30, boundaries excluded;

    - males over 25;

    - people who are either below 68 or female, with `sex` equal to 1.

2.  Select the subjects with age at most 30 and store this subset in a new object. Do not overwrite the original dataset `sbs399`.

    Count the number of subjects of age at least 20 in this subset.

    Separately for men and women, count the number of subjects of age at least 20. Use `.by` or `group_by()` for this last question.

3.  Reload the full dataset `sbs399`.

    Create a selection variable, say `touse`, that takes the value 1 for males over 30 and 0 otherwise.

    For the selected sample:

    - create a two-way table;

    - compute the correlation between the respondent's education, `educ`, and the respondent's father's education, `feduc`.

4.  What is the difference between the following R commands?

```         
data |>   
  select(-sex)  

data |>   
  filter(sex != 1)  

data |>   
  filter(sex == 1) 
```

5.  Using the household identifier `hrespnr` in the dataset `sbs399`, use `.by` or `group_by()` to:

    - define the number of participants in a household;

    - list the oldest people in households without creating a new variable;

    - list the ages of the same-sex households;

    - list `sex` and `age` in increasing order of age in households where the age difference is larger than 5, without creating new variables;

    - drop the singles.

## Answers

Make sure this file is in the same R Project as the `data` folder. Use relative file paths, not full paths to your computer.

```{r setup-select}
library(tidyverse)
library(haven)

sbs399 <- read_dta("data/sbs399.dta")
```

Check the coding of `sex` before using it in any condition.

```{r}
table(sbs399$sex, useNA = "ifany")
# the answers below assume sex = 1 for women and sex = 2 for men
# swap the codes if your frequency table shows the opposite
```

### 1. Counting observations with `filter()` and `nrow()`

Age at most 30.

```{r}
sbs399 |>
  filter(age <= 30) |>
  nrow()
# number of respondents aged 30 or younger
```

Age at least 20.

```{r}
sbs399 |>
  filter(age >= 20) |>
  nrow()
# number of respondents aged 20 or older
```

Age between 20 and 30, boundaries included.

```{r}
sbs399 |>
  filter(age >= 20, age <= 30) |>
  nrow()
# a comma between conditions means AND
# between(age, 20, 30) would give the same result
```

Age between 20 and 30, boundaries excluded.

```{r}
sbs399 |>
  filter(age > 20, age < 30) |>
  nrow()
# respondents of exactly 20 or exactly 30 are now excluded
```

Males over 25.

```{r}
sbs399 |>
  filter(sex == 2, age > 25) |>
  nrow()
# men older than 25
```

People who are either below 68 or female.

```{r}
sbs399 |>
  filter(age < 68 | sex == 1) |>
  nrow()
# the vertical bar means OR, so respondents satisfying either condition are counted
```

Note that `filter()` drops rows for which the condition evaluates to `NA`. A respondent with a missing age is therefore never counted, not even in the `OR` condition above.

```{r}
sum(is.na(sbs399$age))
sum(is.na(sbs399$sex))
# check how many observations are lost for this reason
```

### 2. Working with a subset

Select the subjects with age at most 30 and store them in a new object.

```{r}
sbs399_young <- sbs399 |>
  filter(age <= 30)
# the original dataset sbs399 is left untouched
```

Count the number of subjects of age at least 20 in this subset.

```{r}
sbs399_young |>
  filter(age >= 20) |>
  nrow()
# respondents aged between 20 and 30 in the subset
```

The same count separately for men and women.

```{r}
sbs399_young |>
  filter(age >= 20) |>
  summarise(
    n = n(),
    .by = sex
  )
# counts by sex using .by
```

The same result with `group_by()`.

```{r}
sbs399_young |>
  filter(age >= 20) |>
  group_by(sex) |>
  summarise(n = n())
# group_by() followed by summarise() gives the same table
```

### 3. A selection variable

Reload the full dataset.

```{r}
sbs399 <- read_dta("data/sbs399.dta")
# start again from the complete dataset
```

Create the selection variable `touse`.

```{r}
sbs399 <- sbs399 |>
  mutate(
    touse = if_else(sex == 2 & age > 30, 1, 0)
  )
# touse = 1 for men older than 30, 0 otherwise
# touse is missing if sex or age is missing
```

Check the selection variable.

```{r}
table(sbs399$touse, useNA = "ifany")
# how many observations are selected
```

Create the two-way table of education and father's education for the selected sample.

```{r}
sbs399 |>
  filter(touse == 1) |>
  count(educ, feduc) |>
  pivot_wider(
    names_from = feduc,
    values_from = n,
    values_fill = 0
  )
# cross-tabulation of educ and feduc within the selected sample
```

The base R equivalent.

```{r}
sbs399_selected <- sbs399 |>
  filter(touse == 1)

table(sbs399_selected$educ, sbs399_selected$feduc, useNA = "ifany")
# same two-way table using table()
```

Compute the correlation between `educ` and `feduc` in the selected sample.

```{r}
sbs399 |>
  filter(touse == 1) |>
  summarise(
    r = cor(educ, feduc, use = "complete.obs")
  )
# use = "complete.obs" drops respondents with a missing value on either variable
```

### 4. Comparing three commands

```{r}
# select(-sex) removes the COLUMN sex; all rows are kept.
# filter(sex != 1) removes ROWS for which sex equals 1; the column stays.
# filter(sex == 1) keeps only the rows for which sex equals 1.
#
# The two filter() commands are not exact complements: rows with a missing
# value on sex are dropped by both, because NA != 1 and NA == 1 both
# evaluate to NA rather than to TRUE.
```

Demonstrate this with the number of rows and columns.

```{r}
dim(sbs399)
dim(select(sbs399, -sex))
dim(filter(sbs399, sex != 1))
dim(filter(sbs399, sex == 1))
# the first command changes the number of columns, the other two the number of rows
```

### 5. Working within households

Define the number of participants in a household.

```{r}
sbs399 <- sbs399 |>
  mutate(
    n_household = n(),
    .by = hrespnr
  )
# number of respondents sharing the same household identifier
```

Check the household sizes.

```{r}
table(sbs399$n_household, useNA = "ifany")
# frequency table of household sizes
```

List the oldest people in households, without creating a new variable.

```{r}
sbs399 |>
  filter(age == max(age), .by = hrespnr) |>
  select(hrespnr, sex, age)
# filter() with .by compares each respondent to the maximum age in their own household
# households with a tie return more than one row
```

List the ages of the same-sex households.

```{r}
sbs399 |>
  filter(n_distinct(sex) == 1, .by = hrespnr) |>
  select(hrespnr, sex, age)
# keep only households in which every member has the same value on sex
# note that single-person households satisfy this condition as well
```

List `sex` and `age` in increasing order of age, in households where the age difference is larger than 5.

```{r}
sbs399 |>
  filter(max(age) - min(age) > 5, .by = hrespnr) |>
  arrange(hrespnr, age) |>
  select(hrespnr, sex, age)
# the range of age is computed within each household without storing it
```

Drop the singles.

```{r}
sbs399 <- sbs399 |>
  filter(n() > 1, .by = hrespnr)
# keep only respondents living in a household with at least two participants
```

Check the result.

```{r}
table(sbs399$n_household, useNA = "ifany")
# households of size 1 should no longer appear
```
