---
title: Creating new variables
---

In this part you will learn about the following basic tasks:

- Creating a dummy (0/1) variable.

- Recoding a continuous variable.

- Recoding a continuous variable into groups.

- Centering and standardizing variables.

- Reverse coding of variables.

- Creating additive scales.

## R

Learn the following R commands to create new or modify existing variables.

table

Examples:

Create a variable as a mathematical expression of existing variables:

```{r, eval=FALSE}
data <- data |>
  mutate(
    hhinc = h_inc + w_inc,
    minc = (h_inc + w_inc) / 2,
    adiff = abs(age - page),
    inc2 = inc^2
  )
# sum, mean, absolute difference, and squared income
```

Change negative values of an existing variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    inc = if_else(inc < 0, 0, inc)
  )
# replace negative values by 0
```

Create a dummy variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    male = if_else(sex == 2, 1, 0)
  )
# create dummy variable male
```

Preserve missing values:

```{r, eval=FALSE}
data <- data |>
  mutate(
    male = if_else(
      is.na(sex),
      NA_real_,
      if_else(sex == 2, 1, 0)
    )
  )
# sex = NA implies male = NA
```

Recode a dummy variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    female = if_else(is.na(male), NA_real_, 1 - male)
  )
# male = 1 becomes female = 0 and vice versa
```

Recode values into categories:

```{r, eval=FALSE}
data <- data |>
  mutate(
    region = case_when(
      country %in% c(71, 72, 79) ~ 1,
      between(country, 73, 78) ~ 2,
      !is.na(country) ~ 3,
      TRUE ~ NA_real_
    )
  )
# create region variable
```

Count the number of variables equal to a specific value:

```{r, eval=FALSE}
data <- data |>
  mutate(
    ngood = rowSums(
      across(good01:good05, ~ .x == 2),
      na.rm = TRUE
    )
  )
# count how many variables equal 2
```

Create a standardized variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    zedu = (edu - mean(edu, na.rm = TRUE)) /
      sd(edu, na.rm = TRUE)
  )
# standardize edu
```

Create a rank variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    rinc = rank(inc)
  )
# rank order statistics of income
```

## Assignments

### Create dummy variables.

In the dataset `sbs399` the variable `class` marks the respondent's social class. How often is this variable non-missing?

We want to create a dummy variable for being of white collar. This variable, which we will call `white`, is 1 for subjects belonging to the white-collar class and 0 otherwise. We interpret missing values in `class` to denote "unknown". Thus, the dummy `white` should also be missing if `class` is missing.

- Create this dummy variable using `if_else()`. Make sure that missing values are handled correctly.
- Create the same dummy variable using `case_when()`.
- Produce a cross-tabulation of the two newly created variables and verify that they are identical.

### Recoding a categorical variable.

In the dataset `sbs399` the variable `polit` marks the political party the respondent intended to vote for in a national election.

(a) Generate a new variable `P` in which political preference is expressed on a left-to-right scale:

1 = SP

2 = GroenLinks

3 = PvdA

4 = D66

5 = CDA

6 = VVD

7 = GPV

8 = RPF

9 = SGP

10 = Unie 55+

11 = AOV

12 = CD

Use `case_when()` to construct the variable.

(b) Recode `P` into a new variable `P3` in which political preference is expressed in three categories:

1 = Left (SP, GroenLinks, PvdA, D66)

2 = Center (CDA)

3 = Right (all remaining parties)

### Recode a continuous variable into groups.

We consider the hourly wage variable `hrinc`.

Generate a new variable `Q` for the four quartiles of `hrinc`:

- Q = 1 for the lowest 25%
- Q = 2 for the next 25%
- Q = 3 for the next 25%
- Q = 4 for the highest 25%

Verify that the groups contain approximately equal numbers of observations.

### Centering and standardizing variables.

Use the dataset `sbs399`. We focus on `hrinc`.

(a) Create a variable `c_hrinc` that contains the centered hourly wage rate by subtracting the sample mean of `hrinc`.

Check that:

- the mean of `c_hrinc` is approximately 0;
- the standard deviations of `c_hrinc` and `hrinc` are the same.

(b) Create a variable `z_hrinc` that contains the standardized hourly wage rate by subtracting the mean of `hrinc` and dividing by its standard deviation.

Check that:

- the mean of `z_hrinc` is approximately 0;
- the standard deviation of `z_hrinc` is approximately 1.

(c) Create a variable `z2_hrinc` that standardizes hourly income separately for men and women.

Ensure that:

- you obtain one variable, not two;
- the variable is not missing unless `hrinc` itself is missing.

(d) Apply a unit-range transformation to `hrinc`:

$$U(x) = \frac{x - \min(x)}{\max(x) - \min(x)}$$

Verify that the transformed minimum equals 0 and the transformed maximum equals 1.

Which method do you prefer for income? And for number of hours worked per week?

### Creating additive scales.

We consider the five items `sexrol1` to `sexrol5` in the dataset `sbs399`.

(a) Explain why it is not sensible to simply add the five items.

(b) One of the items should be reverse coded. Which one?

Construct a new variable for this reversed item using:

- a recoding approach with `case_when()`;
- a mathematical expression.

Verify that both constructed variables are identical.

(c) Create the attitude measure from the five items.

Hint: use `rowMeans()`.

What is the mean of the measure? Can you interpret this mean substantively?

(d) The data used so far are complete, meaning that respondents answered all items. In many surveys, however, respondents do not answer every question. This situation is illustrated by the dataset `sbs399_sexroles`.

We will now discuss a simple ad hoc method for dealing with missing values when constructing a scale. Although more advanced methods are available and will be discussed later in the course, this approach remains useful for understanding the basic idea.

We simply take the mean of all valid item responses for each respondent. Thus, if a respondent answered items 1, 2, and 3, but not items 4 and 5, we calculate the scale score as the mean of items 1, 2, and 3. If a respondent answered all items, we take the mean across all items. If a respondent did not answer any of the items, the scale score should be missing as well.

In R, this can be accomplished using `rowMeans()` together with the argument `na.rm = TRUE`, which computes the mean using all available values while ignoring missing values.

## Answers

Make sure this file is in the same R Project as the `data` folder. Use relative file paths, not full paths to your computer.

```{r setup}
library(tidyverse)
library(haven)

sbs399 <- read_dta("data/sbs399.dta")
```

### Create dummy variables

How often is `class` non-missing?

```{r}
sbs399 |>
  filter(!is.na(class)) |>
  nrow()
# number of non-missing values in class
```

Check the categories of `class` before creating the dummy variable.

```{r}
table(sbs399$class, useNA = "ifany")
# frequency table of class, including missing values
```

Create the dummy variable `white` using `if_else()`.

```{r}
sbs399 <- sbs399 |>
  mutate(
    white = if_else(
      is.na(class),
      NA_real_,
      if_else(class == 2, 1, 0)
    )
  )
# white = 1 for white-collar class, white = 0 otherwise, class = NA implies white = NA
```

Create the same dummy variable using `case_when()`.

```{r}
sbs399 <- sbs399 |>
  mutate(
    white2 = case_when(
      class == 2 ~ 1,
      !is.na(class) ~ 0,
      TRUE ~ NA_real_
    )
  )
# same dummy variable, created with case_when()
# change class == 2 if your frequency table shows a different code for white collar
```

Produce a cross-tabulation of the two newly created variables.

```{r}
table(sbs399$white, sbs399$white2, useNA = "ifany")
# the two variables should be identical
```

### Recoding a categorical variable

Check the original party codes before recoding.

```{r}
table(sbs399$polit, useNA = "ifany")
# frequency table of polit; verify which code belongs to which party
```

Generate a new variable `P` in which political preference is expressed on a left-to-right scale.

```{r}
sbs399 <- sbs399 |>
  mutate(
    P = case_when(
      polit == 1 ~ 1,   # SP
      polit == 2 ~ 2,   # GroenLinks
      polit == 3 ~ 3,   # PvdA
      polit == 4 ~ 4,   # D66
      polit == 5 ~ 5,   # CDA
      polit == 6 ~ 6,   # VVD
      polit == 7 ~ 7,   # GPV
      polit == 8 ~ 8,   # RPF
      polit == 9 ~ 9,   # SGP
      polit == 10 ~ 10, # Unie 55+
      polit == 11 ~ 11, # AOV
      polit == 12 ~ 12, # CD
      TRUE ~ NA_real_
    )
  )
# create left-to-right political preference scale
# here the original codes already run from left to right, so P equals polit
# adjust the mapping if your frequency table shows a different ordering
```

Check the new variable.

```{r}
table(sbs399$P, useNA = "ifany")
# frequency table of P
```

Recode `P` into a new variable `P3` with three categories.

```{r}
sbs399 <- sbs399 |>
  mutate(
    P3 = case_when(
      P %in% c(1, 2, 3, 4) ~ 1,
      P == 5 ~ 2,
      !is.na(P) ~ 3,
      TRUE ~ NA_real_
    )
  )
# P3 = 1 Left, P3 = 2 Center, P3 = 3 Right
```

Check the new variable.

```{r}
table(sbs399$P3, useNA = "ifany")
# frequency table of P3
```

### Recode a continuous variable into groups

Generate a new variable `Q` for the four quartiles of `hrinc`.

```{r}
sbs399 <- sbs399 |>
  mutate(
    hrinc_rank = rank(hrinc, na.last = "keep") / sum(!is.na(hrinc)),
    Q = case_when(
      hrinc_rank <= .25 ~ 1,
      hrinc_rank <= .50 ~ 2,
      hrinc_rank <= .75 ~ 3,
      !is.na(hrinc_rank) ~ 4,
      TRUE ~ NA_real_
    )
  )
# create four approximately equal groups of hrinc
# na.last = "keep" makes sure missing values on hrinc stay missing on Q
```

Verify that the groups contain approximately equal numbers of observations.

```{r}
table(sbs399$Q, useNA = "ifany")
# frequency table of Q
```

Remove the helper variable.

```{r}
sbs399 <- sbs399 |>
  select(-hrinc_rank)
# hrinc_rank was only needed to construct Q
```

Note that `ntile()` does the same in one step.

```{r}
table(ntile(sbs399$hrinc, 4), useNA = "ifany")
# dplyr::ntile() is a shortcut for the same quartile split
```

### Centering and standardizing variables

Create a variable `c_hrinc` that contains the centered hourly wage rate.

```{r}
sbs399 <- sbs399 |>
  mutate(
    c_hrinc = hrinc - mean(hrinc, na.rm = TRUE)
  )
# center hrinc by subtracting the mean
```

Check that the mean of `c_hrinc` is approximately 0 and that the standard deviations of `c_hrinc` and `hrinc` are the same.

```{r}
sbs399 |>
  summarise(
    mean_c_hrinc = mean(c_hrinc, na.rm = TRUE),
    sd_c_hrinc = sd(c_hrinc, na.rm = TRUE),
    sd_hrinc = sd(hrinc, na.rm = TRUE)
  )
# mean of c_hrinc should be approximately 0; standard deviations should be the same
```

Create a variable `z_hrinc` that contains the standardized hourly wage rate.

```{r}
sbs399 <- sbs399 |>
  mutate(
    z_hrinc = (hrinc - mean(hrinc, na.rm = TRUE)) /
      sd(hrinc, na.rm = TRUE)
  )
# standardize hrinc
```

Check that the mean of `z_hrinc` is approximately 0 and the standard deviation is approximately 1.

```{r}
sbs399 |>
  summarise(
    mean_z_hrinc = mean(z_hrinc, na.rm = TRUE),
    sd_z_hrinc = sd(z_hrinc, na.rm = TRUE)
  )
# mean should be approximately 0; standard deviation should be approximately 1
```

Check that `sex` has no missing values before grouping.

```{r}
table(sbs399$sex, useNA = "ifany")
# if sex were missing for some respondents, those cases would form their own group
# and z2_hrinc would be missing for them even though hrinc is observed
```

Create a variable `z2_hrinc` that standardizes hourly income separately for men and women.

```{r}
sbs399 <- sbs399 |>
  group_by(sex) |>
  mutate(
    z2_hrinc = (hrinc - mean(hrinc, na.rm = TRUE)) /
      sd(hrinc, na.rm = TRUE)
  ) |>
  ungroup()
# standardize hrinc separately by sex
# this yields one variable, and it is missing only where hrinc is missing
```

Check the means and standard deviations by sex.

```{r}
sbs399 |>
  group_by(sex) |>
  summarise(
    mean_z2_hrinc = mean(z2_hrinc, na.rm = TRUE),
    sd_z2_hrinc = sd(z2_hrinc, na.rm = TRUE)
  )
# within each sex, mean should be approximately 0 and standard deviation approximately 1
```

Apply a unit-range transformation to `hrinc`.

```{r}
sbs399 <- sbs399 |>
  mutate(
    u_hrinc = (hrinc - min(hrinc, na.rm = TRUE)) /
      (max(hrinc, na.rm = TRUE) - min(hrinc, na.rm = TRUE))
  )
# unit-range transformation of hrinc
```

Verify that the transformed minimum equals 0 and the transformed maximum equals 1.

```{r}
sbs399 |>
  summarise(
    min_u_hrinc = min(u_hrinc, na.rm = TRUE),
    max_u_hrinc = max(u_hrinc, na.rm = TRUE)
  )
# minimum should be 0 and maximum should be 1
```

Which method do you prefer for income? And for number of hours worked per week?

```{r}
# For income, standardization is often useful because income can be very skewed.
# For hours worked per week, unit-range transformation can be easier to interpret.
# The best choice depends on the goal of the analysis.
```

### Creating additive scales

Explain why it is not sensible to simply add the five items.

```{r}
# It is not sensible to simply add the five items if one item is coded in the opposite direction.
# In that case, high values do not mean the same thing for all items.
# The reversed item should first be recoded.
```

Which item should be reverse coded?

```{r}
sbs399 |>
  select(sexrol1:sexrol5) |>
  cor(use = "complete.obs")
# the reversed item is the one correlating negatively with the other four
# here this is sexrol5
```

Reverse item 5 using `case_when()`.

```{r}
sbs399 <- sbs399 |>
  mutate(
    sexrol5r_case = case_when(
      sexrol5 == 1 ~ 5,
      sexrol5 == 2 ~ 4,
      sexrol5 == 3 ~ 3,
      sexrol5 == 4 ~ 2,
      sexrol5 == 5 ~ 1,
      TRUE ~ NA_real_
    )
  )
# reverse code sexrol5 using case_when()
```

Reverse item 5 using a mathematical expression.

```{r}
sbs399 <- sbs399 |>
  mutate(
    sexrol5r_math = 6 - sexrol5
  )
# reverse code sexrol5 using a mathematical expression
```

Verify that both constructed variables are identical.

```{r}
table(sbs399$sexrol5r_case, sbs399$sexrol5r_math, useNA = "ifany")
# the two reverse-coded variables should be identical
```

Create the attitude measure from the five items.

```{r}
sbs399 <- sbs399 |>
  mutate(
    sexrole_scale = rowMeans(
      cbind(
        sexrol1,
        sexrol2,
        sexrol3,
        sexrol4,
        sexrol5r_math
      )
    )
  )
# create additive scale as row mean of the five items
# no na.rm is needed here because this dataset is complete
```

What is the mean of the measure?

```{r}
sbs399 |>
  summarise(
    mean_sexrole_scale = mean(sexrole_scale, na.rm = TRUE)
  )
# mean of the sex role attitude scale
```

Can you interpret this mean substantively?

```{r}
# The scale mean gives the average position on the sex role attitude items.
# A substantive interpretation depends on the direction of the item coding.
```

Open the dataset with missing values.

```{r}
sbs399_sexroles <- read_dta("data/sbs399_sexroles.dta")
# open the dataset with missing values on the sex role items
```

Create the attitude measure using all available item responses.

```{r}
sbs399_sexroles <- sbs399_sexroles |>
  mutate(
    sexrol5r_math = 6 - sexrol5,
    n_valid = rowSums(
      !is.na(
        cbind(
          sexrol1,
          sexrol2,
          sexrol3,
          sexrol4,
          sexrol5r_math
        )
      )
    ),
    sexrole_scale = if_else(
      n_valid == 0,
      NA_real_,
      rowMeans(
        cbind(
          sexrol1,
          sexrol2,
          sexrol3,
          sexrol4,
          sexrol5r_math
        ),
        na.rm = TRUE
      )
    )
  )
# create scale using the mean of available item responses
# if all items are missing, the scale is missing
```

Check the scale.

```{r}
sbs399_sexroles |>
  summarise(
    mean_sexrole_scale = mean(sexrole_scale, na.rm = TRUE),
    min_sexrole_scale = min(sexrole_scale, na.rm = TRUE),
    max_sexrole_scale = max(sexrole_scale, na.rm = TRUE)
  )
# summary statistics for the scale with missing values handled using rowMeans(na.rm = TRUE)
```
