---
title: Creating new variables
---

In this part you will learn about the following basic tasks:

- Creating a dummy (0/1) variable.

- Recoding a continuous variable.

- Recoding a continuous variable into groups.

- Centering and standardizing variables.

- Reverse coding of variables.

- Creating additive scales.

## R

Learn the following R commands to create new or modify existing variables.

table

Examples:

Create a variable as a mathematical expression of existing variables:

```{r, eval=FALSE}
data <- data |>
  mutate(
    hhinc = h_inc + w_inc,
    minc = (h_inc + w_inc) / 2,
    adiff = abs(age - page),
    inc2 = inc^2
  )
# sum, mean, absolute difference, and squared income
```

Change negative values of an existing variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    inc = if_else(inc < 0, 0, inc)
  )
# replace negative values by 0
```

Create a dummy variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    male = if_else(sex == 2, 1, 0)
  )
# create dummy variable male
```

Preserve missing values:

```{r, eval=FALSE}
data <- data |>
  mutate(
    male = if_else(
      is.na(sex),
      NA_real_,
      if_else(sex == 2, 1, 0)
    )
  )
# sex = NA implies male = NA
```

Recode a dummy variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    female = if_else(is.na(male), NA_real_, 1 - male)
  )
# male = 1 becomes female = 0 and vice versa
```

Recode values into categories:

```{r, eval=FALSE}
data <- data |>
  mutate(
    region = case_when(
      country %in% c(71, 72, 79) ~ 1,
      between(country, 73, 78) ~ 2,
      !is.na(country) ~ 3,
      TRUE ~ NA_real_
    )
  )
# create region variable
```

Count the number of variables equal to a specific value:

```{r, eval=FALSE}
data <- data |>
  mutate(
    ngood = rowSums(
      across(good01:good05, ~ .x == 2),
      na.rm = TRUE
    )
  )
# count how many variables equal 2
```

Create a standardized variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    zedu = (edu - mean(edu, na.rm = TRUE)) /
      sd(edu, na.rm = TRUE)
  )
# standardize edu
```

Create a rank variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    rinc = rank(inc)
  )
# rank order statistics of income
```

## Assignments

See `dh_variables_assignments.Rmd` for solutions.

### Create dummy variables.

In the dataset `sbs399` the variable `class` marks the respondent's social class. How often is this variable non-missing?

We want to create a dummy variable for being of white collar. This variable, which we will call `white`, is 1 for subjects belonging to the white-collar class and 0 otherwise. We interpret missing values in `class` to denote "unknown". Thus, the dummy `white` should also be missing if `class` is missing.

- Create this dummy variable using `if_else()`. Make sure that missing values are handled correctly.
- Create the same dummy variable using `case_when()`.
- Produce a cross-tabulation of the two newly created variables and verify that they are identical.

### Recoding a categorical variable.

In the dataset `sbs399` the variable `polit` marks the political party the respondent intended to vote for in a national election.

(a) Generate a new variable `P` in which political preference is expressed on a left-to-right scale:

1 = SP

2 = GroenLinks

3 = PvdA

4 = D66

5 = CDA

6 = VVD

7 = GPV

8 = RPF

9 = SGP

10 = Unie 55+

11 = AOV

12 = CD

Use `case_when()` to construct the variable.

(b) Recode `P` into a new variable `P3` in which political preference is expressed in three categories:

1 = Left (SP, GroenLinks, PvdA, D66)

2 = Center (CDA)

3 = Right (all remaining parties)

3.  Recode a continuous variable into groups.

We consider the hourly wage variable `hrinc`.

Generate a new variable `Q` for the four quartiles of `hrinc`:

- Q = 1 for the lowest 25%
- Q = 2 for the next 25%
- Q = 3 for the next 25%
- Q = 4 for the highest 25%

Verify that the groups contain approximately equal numbers of observations.

### Centering and standardizing variables.

Use the dataset `sbs399`. We focus on `hrinc`.

(a) Create a variable `c_hrinc` that contains the centered hourly wage rate by subtracting the sample mean of `hrinc`.

Check that:

- the mean of `c_hrinc` is approximately 0;
- the standard deviations of `c_hrinc` and `hrinc` are the same.

(b) Create a variable `z_hrinc` that contains the standardized hourly wage rate by subtracting the mean of `hrinc` and dividing by its standard deviation.

Check that:

- the mean of `z_hrinc` is approximately 0;
- the standard deviation of `z_hrinc` is approximately 1.

(c) Create a variable `z2_hrinc` that standardizes hourly income separately for men and women.

Ensure that:

- you obtain one variable, not two;
- the variable is not missing unless `hrinc` itself is missing.

(d) Apply a unit-range transformation to `hrinc`:

$$U(x) = \frac{x - \min(x)}{\max(x) - \min(x)}$$

Verify that the transformed minimum equals 0 and the transformed maximum equals 1.

Which method do you prefer for income? And for number of hours worked per week?

### Creating additive scales.

We consider the five items `sexrol1` to `sexrol5` in the dataset `sbs399`.

(a) Explain why it is not sensible to simply add the five items.

(b) One of the items should be reverse coded. Which one?

Construct a new variable for this reversed item using:

- a recoding approach with `case_when()`;
- a mathematical expression.

Verify that both constructed variables are identical.

(c) Create the attitude measure from the five items.

Hint: use `rowMeans()`.

What is the mean of the measure? Can you interpret this mean substantively?

(d) The data used so far are complete, meaning that respondents answered all items. In many surveys, however, respondents do not answer every question. This situation is illustrated by the dataset `sbs399_sexroles`.

We will now discuss a simple ad hoc method for dealing with missing values when constructing a scale. Although more advanced methods are available and will be discussed later in the course, this approach remains useful for understanding the basic idea.

We simply take the mean of all valid item responses for each respondent. Thus, if a respondent answered items 1, 2, and 3, but not items 4 and 5, we calculate the scale score as the mean of items 1, 2, and 3. If a respondent answered all items, we take the mean across all items. If a respondent did not answer any of the items, the scale score should be missing as well.

In R, this can be accomplished using `rowMeans()` together with the argument `na.rm = TRUE`, which computes the mean using all available values while ignoring missing values.

# Answers

```{r setup, include=FALSE}
library(tidyverse)
library(haven)
library(janitor)
library(skimr)

# Make sure this R Markdown file is saved in the same working directory as the data file.
# The file sbs399.dta should be available in your working directory.

sbs399 <- read_dta("data/sbs399.dta")
```

## 1. What is the variable label of `feduc?`

```{r}
glimpse(sbs399$feduc)
#fathers highest education
```

## 2. Create an overview of the variables `feduc`and `meduc`.

```{r}
sbs399 |>
  select(feduc, meduc) |>
  glimpse()
# compact overview of feduc and meduc

skim(sbs399 |> select(feduc, meduc))
# extended summary statistics for feduc and meduc
```

## 3. Display a frequency distribution of `educ`. What is the percentage of subjects with aneducation up to category 4? How many missing values are there in `educ`?

```{r}
table(sbs399$educ, useNA = "ifany")
# frequency distribution of educ, including missing values

sbs399 |>
  tabyl(educ)
# frequency table with percentages

# There are (0.07373134 + 0.18208955 + 0.15044776 + 0.06686567) = 0.4731343 = 47.31% of subjects with education up to category 4.
# Missings = 0

```

## 4. List the values of the variables `hrinc`and `pres . Adjust the syntax to list only the males among the initial 40 observations. Further adjust the syntax to restrict the sample tomales with below average education among the initial 40 observations. Finally, adjust the syntax to list males and females separately among the initial 5 observations.

```{r}
sbs399 |>
  select(hrinc, pres)
```

Only males among the initial 40 observations:

```{r}
sbs399 |>
  slice(1:40) |>
  filter(sex == 1) |>
  select(hrinc, pres)
```

Males with below average education among the initial 40 observations:

```{r}
sbs399 |>
  slice(1:40) |>
  filter(
    sex == 1,
    educ < mean(educ, na.rm = TRUE)
  ) |>
  select(hrinc, pres, educ, sex)
```

Males and females separately among the initial 5 observations:

```{r}
sbs399 |>
  slice(1:5) |>
  arrange(sex) |>
  select(sex, hrinc, pres)
```

## 5. List values of `age in increasing order. Next, list just the five smallest values, and list justthe five largest values. Do not read these values from the data viewer. Give syntax todisplay the requested cases. Then list all observations except the observations between 6and 3345, both numbers included.

Values of `age` in increasing order:

```{r}
sbs399 |>
  arrange(age) |>
  select(age)
```

Five smallest values:

```{r}
sbs399 |>
  arrange(age) |>
  slice(1:5) |>
  select(age)
```

Five largest values:

```{r}
sbs399 |>
  arrange(desc(age)) |>
  slice(1:5) |>
  select(age)
```

All observations except observations 6 to 3345:

```{r}
sbs399 |>
  slice(-c(6:3345))
```

## 6. Create a two-way table of `feduc` rows and `meduc` columns. Add row and column percent-ages. Test for independence of `feduc`and `meduc` using Pearson’s chi-square test.

Frequency table:

```{r}
tab_feduc_meduc <- table(sbs399$feduc, sbs399$meduc)
tab_feduc_meduc
```

Row percentages:

```{r}
prop.table(tab_feduc_meduc, margin = 1) * 100
```

Column percentages:

```{r}
prop.table(tab_feduc_meduc, margin = 2) * 100
```

Pearson's chi-square test:

```{r}
chisq.test(tab_feduc_meduc)
```

## 7. What are the mean, standard deviation, minimum, and maximum value of `feduc`? And what is the median?

```{r}
sbs399 |>
  summarise(
    mean = mean(feduc, na.rm = TRUE),
    sd = sd(feduc, na.rm = TRUE),
    min = min(feduc, na.rm = TRUE),
    median = median(feduc, na.rm = TRUE),
    max = max(feduc, na.rm = TRUE)
  )

# Mean = 3.04
# Standard deviation = 2.20
# Minimum = 1 (elementary [lower])
# Median = 2
# Maximum = 8 (university [wo])
```

## 8. Verify that the number of subjects for whom the mother is higher educated than the father is 543. What is the count if you restrict to female subjects?

```{r}
sbs399 |>
  filter(meduc > feduc) |>
  nrow()
```

Restricted to female subjects:

```{r}
sbs399 |>
  filter(
    meduc > feduc,
    sex == 1
  ) |>
  nrow()

# Count restricted to female subjects = 262
```

## 9. What are the Pearson correlations of `educ`, `feduc`, and `meduc`? Use the help system to
find out how to compute Spearman’s rank correlations.

Pearson correlations:

```{r}
sbs399 |>
  select(educ, feduc, meduc) |>
  cor(use = "complete.obs", method = "pearson")
```

Spearman rank correlations:

```{r}
sbs399 |>
  select(educ, feduc, meduc) |>
  cor(use = "complete.obs", method = "spearman")
```

## 10. Is it true that for any variable the values are above average for 50% of cases? If you are uncertain, check this with some variables. For an explanation, think about this example: what percentage of people have more than the average number of legs? 50%? Who are these people?

No, this is not necessarily true. The mean is not the same as the median. A variable can be skewed, and then the percentage of cases above the mean can be much lower or much higher than 50%.

Check with some variables:

```{r}
sbs399 |>
  summarise(
    pct_age_above_mean = mean(age > mean(age, na.rm = TRUE), na.rm = TRUE) * 100,
    pct_educ_above_mean = mean(educ > mean(educ, na.rm = TRUE), na.rm = TRUE) * 100,
    pct_hrinc_above_mean = mean(hrinc > mean(hrinc, na.rm = TRUE), na.rm = TRUE) * 100
  )
```

## 11. Is `hrespnr` a key, meaning that it is never missing and all values are unique? Why? Do not rely on the data viewer. Write syntax to obtain the information on which you base your answer. What about the pair of variables `hrespnr` and `sex`? Why? What about`hrespnr`, `sex`, and `age`, and `hrespnr`, `sex`, and `fage`?

Check whether `hrespnr` is never missing:

```{r}
sbs399 |>
  summarise(no_missing_hrespnr = all(!is.na(hrespnr)))
```

Check whether `hrespnr` is unique:

```{r}
sbs399 |>
  count(hrespnr) |>
  filter(n > 1)
```

If this returns rows, `hrespnr` is not a key.

Check whether `hrespnr` and `sex` form a composite key:

```{r}
sbs399 |>
  summarise(no_missing = all(!is.na(hrespnr), !is.na(sex)))

sbs399 |>
  count(hrespnr, sex) |>
  filter(n > 1)
```

Check whether `hrespnr`, `sex`, and `age` form a composite key:

```{r}
sbs399 |>
  summarise(no_missing = all(!is.na(hrespnr), !is.na(sex), !is.na(age)))

sbs399 |>
  count(hrespnr, sex, age) |>
  filter(n > 1)
```

Check whether `hrespnr`, `sex`, and `fage` form a composite key:

```{r}
sbs399 |>
  summarise(no_missing = all(!is.na(hrespnr), !is.na(sex), !is.na(fage)))

sbs399 |>
  count(hrespnr, sex, fage) |>
  filter(n > 1)
```

A set of variables is a key if it is never missing and if there are no duplicate combinations.