---
title: Creating new variables
---

In this part you will learn about the following basic tasks:

- Creating a binary (0/1) variable.

- Recoding a categorical variable.

- Recoding a continuous variable into groups.

- Centering and standardizing variables.

- Reverse coding of variables.

- Creating additive scales.

## R

To create or modify existing variables in a dataset, we will use the function `mutate()`. 

The table below shows useful functions for recoding variables. These functions are normally used within `mutate()`.

| Function                      | Description                                        |
|-------------------------------|----------------------------------------------------|
| `if_else()`                   | Create/modify based on simple logical condition    |
| `case_when()`                 | Create/modify based on multiple logical conditions |
| `recode_values()`             | Recode values  (`dplyr` version >= 1.2)            |
| `replace_values()`            | Replace values (`dplyr` version >= 1.2)            |
| `across()`                    | Applies function to multiple columns               |



Examples:

Create a variable as a mathematical expression of existing variables:

```{r, eval=FALSE}
data <- data |>
  mutate(
    hhinc = h_inc + w_inc,
    minc = (h_inc + w_inc) / 2,
    adiff = abs(age - page),
    inc2 = inc^2
  )
# sum, mean, absolute difference, and squared income (^ denotes "power")
```

Change negative values of an existing variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    inc = if_else(inc < 0, 0, inc)
  )
# replace negative of values of inc by 0
```

Create a binary ("dummy") variable:
```{r, eval=FALSE}
# Create binary variable male. 
# Suppose that sex is coded as female=1, male=2, and there are no missings

# The variable can be numerical (0/1)..
data |>
  mutate(
    male = if_else(sex == 2, 1, 0) # male = 1, female = 0. In thise case, we could also simply do `male = sex - 1`
  )

# or logical (TRUE/FALSE)...
data |>
  mutate(
    male = (sex == 2) # male = TRUE, female = FALSE
  )
```

For many uses in R, having a binary variable as numeric (1/0) or logical (TRUE/FALSE) makes little difference. For example, R will treat TRUE/FALSE as 1/0 if you compute the mean of a logical vector. Logical binary variables can make code a bit more succint when they are used in other operations: with logical vectors you can write e.g `filter(male, employed)` instead of `filter(male == 1, employed == 1)`. But be aware that some mathematical and modelling functions expect only numerical variables. 

In the code above, we presumed thhat there were no missing values. If there were missing values, they would be coded as `0`, which is obviously wrong. To deal with with missing values:
```{r, eval=FALSE}
data <- data |>
  mutate(
    male = recode_values(sex,
                         1 ~ 0,
                         2 ~ 1) # Change 1 into 0, 2 into 1
  )
# By default, recode_values() assigns missing (NA) to any values we did not specify. So any NAs would stay as NAs, and any values otehr than 1 and 2 would become NA).
```

Recode a dummy variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    female = if_else(is.na(male), NA_real_, 1 - male)
  )
# male = 1 becomes female = 0 and vice versa
```

Recode values into categories:

```{r, eval=FALSE}
# create region variable
# 71, 72, 79 = 1
# 73 to 78 = 2
# all other valide (non NA) codes = 3
data <- data |>
  mutate(
    region = recode_values(country,
                            c(71, 72, 79)  ~ 1,
                            73:78          ~ 2,
                            NA ~ NA,
                            default = 3 # this applies to all other values not speficied above. 
                                        # Attention: if we had nto specified NA ~ NA above, 
                                        # NA values would also have been recoded to 3
    )
  )
```

You can also recode using `case_when()`, specifying logical conditions instead of original values on the lef-hand side. `case_when()` is thus more flexible (e.g. it can handle conditions based on multiple original variables, see below) but tends to more verbose.

```{r, eval=FALSE}
# Same task as above, using case_when().
# Note that we explicitly refer to the original variable in each condition
data <- data |>
  mutate(
    region = case_when(
      country %in% c(71, 72, 79) ~ 1,
      country %in% 73:78         ~ 2,
      !is.na(country)            ~ 3,
      is.na(country)             ~ NA
    )
  )


# Recoding based on multiple existing variables
data <- data |>
  mutate(
    group = case_when(
      (male==1 & age < 18)  ~ "male_minor",
      (male==1 & age >= 18) ~ "male_adult"
      (male==0 & age < 18)  ~ "female_minor",
      (male==0 & age >= 18) ~ "female_adult"
      .default              ~ NA  # Everything else as NA. Note the dot in .default.
    )
  )
```


Create a standardized variable (z-scores):

```{r, eval=FALSE}
data <- data |>
  mutate(
    zedu = (edu - mean(edu, na.rm = TRUE)) /
      sd(edu, na.rm = TRUE)
  )
# standardize edu
```

Create a rank variable:

```{r, eval=FALSE}
data <- data |>
  mutate(
    rinc = rank(inc)
  )
# rank order statistics of income
```

With `across()` you can repeat operations across several columns: 

```{r, eval=FALSE}
# Create a new variable counting the number of variables with value == 2:
data <- data |>
  mutate(
    ngood = rowSums(
      across(good01:good05, ~ .x == 2), # for each variable good1 to good5, check if the value is 2
      na.rm = TRUE
    )
  )

# Change 9999 to NA in all variables whose names start with "income" 
data <- data |>
  mutate(
         across(starts_with("income"), ~if_else(.x == 9999, NA, .x))
         )
  
```

## Assignments

### Create dummy variables.

In the dataset `sbs399` the variable `class` marks the respondent's social class. How often is this variable non-missing?

We want to create a dummy variable for being of white collar. This variable, which we will call `white`, is 1 for subjects belonging to the white-collar class and 0 otherwise. We interpret missing values in `class` to denote "unknown". Thus, the dummy `white` should also be missing if `class` is missing.

- Create this dummy variable using `if_else()`. Call it `white1`.
- Create the same dummy variable using `recode_values()`. Call it `white2`.
- Produce a cross-tabulation of the two newly created variables and verify that they are identical.

### Recoding a categorical variable.

In the dataset `sbs399` the variable `polit` marks the political party the respondent intended to vote for in a national election.

(a) Generate a new variable `pol_pref` in which political preference is expressed on a left-to-right scale:

1 = SP
2 = GroenLinks
3 = PvdA
4 = D66
5 = CDA
6 = VVD
7 = GPV
8 = RPF
9 = SGP
10 = Unie 55+
11 = AOV
12 = CD

Use `case_when()` to construct the variable.

(b) Recode `pol_pref` into a new variable `pol_pref3` in which political preference is expressed in three categories:

1 = Left (SP, GroenLinks, PvdA, D66)

2 = Center (CDA)

3 = Right (all remaining parties)

### Recode a continuous variable into groups.

We consider the hourly wage variable `hrinc`.

Generate a new variable `hrinc_quart` for the four quartiles of `hrinc`:

- hrinc_quart = 1 for the lowest 25%
- hrinc_quart = 2 for the next 25%
- hrinc_quart = 3 for the next 25%
- hrinc_quart = 4 for the highest 25%

Verify that the groups contain approximately equal numbers of observations.

### Centering and standardizing variables.

We focus on `hrinc` again.

(a) Create a variable `c_hrinc` that contains the centered hourly wage rate by subtracting the sample mean of `hrinc`.

Check that:

- the mean of `c_hrinc` is approximately 0;
- the standard deviations of `c_hrinc` and `hrinc` are the same.

(b) Create a variable `z_hrinc` that contains the standardized hourly wage rate by subtracting the mean of `hrinc` and dividing by its standard deviation.

Check that:

- the mean of `z_hrinc` is approximately 0;
- the standard deviation of `z_hrinc` is approximately 1.

(c) Create a variable `z2_hrinc` that standardizes hourly income separately for men and women. Within each sex, the mean of `z2_hrinc` should be approximately 0 and standard deviation approximately 1.

Ensure that:

- you obtain one variable, not two;
- the variable is not missing unless `hrinc` itself is missing.

(d) Apply a unit-range transformation to `hrinc`:

$$U(x) = \frac{x - \min(x)}{\max(x) - \min(x)}$$

Verify that the transformed minimum equals 0 and the transformed maximum equals 1.

Which method do you prefer for income? And for number of hours worked per week?

### Creating scales

We consider the five items `sexrol1` to `sexrol5` in the dataset `sbs399`. These items measure attitudes towards women’s emancipation by (5 point) Likert items. We want to average the 5 items and treat it as a measurement of attitude, with a larger value indicating a more traditional attitude.

(a) Explain why it is not sensible to simply add the five items.

(b) You likely want to reverse one of the items. Which one?

Construct a new variable for this reversed item using:

- a recoding approach with `recode_values()` (or `case_when()`);
- a mathematical expression.

Verify that both constructed variables are identical.

(c) Create the attitude measure from the five items.

Hint: use `rowMeans()`.

What is the mean of the measure? Can you interpret this mean substantively?

(d) The data used so far are complete, meaning that respondents answered all items. In many surveys, however, respondents do not answer every question. This situation is illustrated by the dataset `sbs399_sexroles`.

We will now discuss a simple ad hoc method to deal with this. This method is quick-and-dirty; it used to be widely used, but theoretical statistics has yielded better methods. We will discuss these more advanced methods later on.

For now  we simply take the mean of all valid item responses for each respondent. Thus, if a respondent answered items 1, 2, and 3, but not items 4 and 5, we calculate the scale score as the mean of items 1, 2, and 3. If a respondent answered all items, we take the mean across all items. If a respondent did not answer any of the items, the scale score should be missing as well.

In R, this can be accomplished using the argument `na.rm = TRUE` in `rowMeans()` (or in `mean()` used together with `rowwise()`), which computes the mean using all available values while ignoring missing values.

## Answers

Make sure this file is in the same R Project as the `data` folder. Use relative file paths, not full paths to your computer.

```{r setup-variables}
library(tidyverse)
library(haven)
library(tabyl)
library(skimr)

sbs399 <- read_dta("data/sbs399.dta")
```

### Create dummy variables

How often is `class` non-missing?

```{r}
sbs399 |>
  filter(!is.na(class)) |>
  nrow()
# number of non-missing values in class
```

Check the categories of `class` before creating the dummy variable. One way of doing this is using `haven::print_labels()`. Also check if there missing values

```{r}
print_labels(sbs399$class)
# 1 = white collar 

table(sbs399$class, useNA = "ifany")
# There are missing values
```

Create the dummy variable `white` using `if_else()`.

```{r}
sbs399 <- sbs399 |>
  mutate(
    white1 = if_else(
      is.na(class),
      NA_real_,
      if_else(class == 1, 1, 0)
    )
  )
# white = 1 for white-collar class, white = 0 otherwise, class = NA implies white = NA
```

Create the same dummy variable using `recode_values()`.

```{r}
sbs399 <- sbs399 |>
  mutate(
    white2 = recode_values(class,
                           1 ~ 1,
                           NA ~ NA,
                           default = 0
                          )
  )
# same dummy variable, created with recode_values()
```

Produce a cross-tabulation of the two newly created variables.
```{r}
sbs399  |> 
  tabyl(white1, white2)

# Also good to check against the original variable:
sbs399  |> 
   tabyl(white1, class)
```

### Recoding a categorical variable

Check the original party codes before recoding.

```{r}
sbs399 |> 
  tabyl(polit)
print_labels(sbs399$polit)
```

Generate a new variable `pol_pref` in which political preference is expressed on a left-to-right scale.

```{r}
sbs399 <- sbs399 |>
  mutate(
    pol_pref = recode_values(
                polit,
                1 ~ 3,   # PvdA
                2 ~ 5,   # CDA
                3 ~ 4,   # D66
                4 ~ 6,   # VVD
                5 ~ 2,   # GroenLinks
                6 ~ 9,   # SGP
                7 ~ 8,   # RPF
                8 ~ 7,   # GPV
                9 ~ 11,  # AOV
                10 ~ 10, # Unie 55+
                11 ~ 1,  # SP
                12 ~ 12, # CD
                13 ~ NA
    )
  )


```

Check the new variable.

```{r}
sbs399 |> 
  tabyl(pol_pref)

# Compare with original variable
sbs399 |> 
  tabyl(polit, pol_pref)
```

Recode `pol_pref` into a new variable `pol_pref3` with three categories.

```{r}
sbs399 <- sbs399 |>
  mutate(
    pol_pref3 = recode_values(pol_pref,
                              1:4  ~ 1,
                              5    ~ 2,
                              6:12 ~ 3
                              )
  )
# pol_pref3 = 1 Left, pol_pref3 = 2 Center, pol_pref3 = 3 Right
```

Check the new variable.

```{r}
sbs399 |> 
  tabyl(pol_pref3)
```

### Recode a continuous variable into groups

Generate a new variable `hrinc_quart` for the four quartiles of `hrinc`.

```{r}
sbs399 <- sbs399 |>
  mutate(
    hrinc_rank = rank(hrinc, na.last = "keep") / sum(!is.na(hrinc)),
    hrinc_quart = case_when(
      hrinc_rank <= .25                    ~ 1,
      hrinc_rank  > .25 & hrinc_rank <=.50 ~ 2,
      hrinc_rank  > .50 & hrinc_rank <=.75 ~ 3,
      hrinc_rank  > .75 ~ 4,
      NA ~ NA
    )
  )
# create four approximately equal groups of hrinc
# na.last = "keep" makes sure missing values on hrinc stay missing on hrinc_rank
```

Verify that the groups contain approximately equal numbers of observations.

```{r}
sbs399 |> 
  tabyl(hrinc_quart)
```

Other solutions below. Note that they give slightly different results. The computation of quantile groups if often very ambiguous, because results differ based on how you handle ties, how you claddify cases right on the quantile value, etc. 

```{r}
sbs399 <- sbs399  |> 
   mutate(# Using cut() and quantile()
          hrinc_quart2 = cut(hrinc, 
                            breaks = quantile(hrinc,
                                              probs = c(0, 0.25, 0.5, 0.75, 1), 
                                              na.rm = TRUE), 
                            include.lowest = TRUE),

          # Using dplyr::ntile() 
          hrinc_quart3 = ntile(hrinc, 4)
    )  

# Compare:
sbs399 |>  tabyl(hrinc_quart, hrinc_quart2)
sbs399 |>  tabyl(hrinc_quart, hrinc_quart3)
```

### Centering and standardizing variables

Create a variable `c_hrinc` that contains the centered hourly wage rate.

```{r}
sbs399 <- sbs399 |>
  mutate(
    c_hrinc = hrinc - mean(hrinc, na.rm = TRUE)
  )
# center hrinc by subtracting the mean
```

Check that the mean of `c_hrinc` is approximately 0 and that the standard deviations of `c_hrinc` and `hrinc` are the same.

```{r}
sbs399 |>
  skim(c_hrinc, hrinc)
```

Create a variable `z_hrinc` that contains the standardized hourly wage rate. Remeber that we already computed the deviation from the mean as `c_hrinc`, so we can use it here. 

```{r}
sbs399 <- sbs399 |>
  mutate(
    z_hrinc = c_hrinc / sd(hrinc, na.rm = TRUE)
  )

# You can also create standardized values using the function scale()
```

Check that the mean of `z_hrinc` is approximately 0 and the standard deviation is approximately 1.

```{r}
sbs399 |>
  skim(z_hrinc)
```

Create a variable `z2_hrinc` that standardizes hourly income separately for men and women.

```{r}
sbs399 <- sbs399 |>
  group_by(sex) |>
  mutate(
    z2_hrinc = (hrinc - mean(hrinc, na.rm = TRUE)) /
      sd(hrinc, na.rm = TRUE)
  ) |>
  ungroup()

# Check missings
sbs399 |> 
  skim(hrinc, z2_hrinc) # same number of missing
sbs399 |> 
  filter(is.na(z2_hrinc), !is.na(hrinc))
  # No cases in which z2_hrinc is missing and hrinc is not
```

Check the means and standard deviations by sex.

```{r}
sbs399 |>
  group_by(sex) |>
  skim(z2_hrinc) |>
  ungroup()
```


Apply a unit-range transformation to `hrinc`.

```{r}
sbs399 <- sbs399 |>
  mutate(
    u_hrinc = (hrinc - min(hrinc, na.rm = TRUE)) /
      (max(hrinc, na.rm = TRUE) - min(hrinc, na.rm = TRUE))
  )
# unit-range transformation of hrinc
```

Verify that the transformed minimum equals 0 and the transformed maximum equals 1.

```{r}
sbs399 |>
  skim(u_hrinc)
```

Which method do you prefer for income? And for number of hours worked per week? 

### Creating  scales

Which item should be reverse coded? 

```{r}
# Look at variable labels. We can dot it separelty for each variable using glimpse(), but it faster with var_label() from pkg labelled
sbs399 |>
  select(sexrol1:sexrol5) |>
  labelled::var_label()

# Looking at correlations between all variables is also informative
sbs399 |>
  select(sexrol1:sexrol5) |>
  cor()
# Note that item 5 correlates negatively with all the other
```

Reverse item 5:

```{r}
# Using recode_values()
sbs399 <- sbs399 |>
  mutate(
      sexrol5rev = recode_values(sexrol5, 
                                  1 ~ 5,
                                  2 ~ 4,
                                  3 ~ 3,
                                  4 ~ 2,
                                  5 ~ 1,
                                  default =  NA
                                )
  )

sbs399 <- sbs399 |>
  mutate(
    sexrol5rev_math = 6 - sexrol5
  )

# Verify that constructed variables are identical
sbs399 |> 
  tabyl(sexrol5rev, sexrol5rev_math)
```


Create the attitude measure from the five items.

```{r}
# Possible solutions, all with the same result IN THIS CASE (because of no missing)
sbs399 <- sbs399 |> 
            mutate(sexrole_scale1 = (sexrol1 + sexrol2 + sexrol3 + sexrol4 + sexrol5rev)/5)

sbs399 <- sbs399 |>
            mutate(sexrole_scale2 = rowMeans(
                                    cbind(sexrol1, sexrol2, sexrol3, sexrol4, sexrol5rev)
                                    )
              ) 
sbs399 <- sbs399 |>            
            rowwise()  |> 
            mutate(sexrole_scale3 = 
                        mean(
                           c(sexrol1, sexrol2, sexrol3, sexrol4, sexrol5rev)
                            )
            ) |> 
            ungroup()

```

What is the mean of the measure?

```{r}
sbs399 |>
  skim(sexrole_scale1, 
       sexrole_scale2,
       sexrole_scale3)

```

Open the dataset with missing values.

```{r}
sbs399_sexroles <- read_dta("data/sbs399_sexroles.dta")
```

Create the attitude measure using all available item responses.

```{r}
sbs399_sexroles <- sbs399_sexroles |>
  mutate(
    sexrol5rev = 6 - sexrol5,
    
    sexrole_scale =  rowMeans(
                          cbind(sexrol1, sexrol2, sexrol3, sexrol4, sexrol5rev),
                            na.rm = TRUE
                             )
    ) 


# OR
sbs399_sexroles <- sbs399_sexroles |>            
                    rowwise()  |> 
                    mutate(sexrole_scale2 = 
                                mean(
                                    c(sexrol1, sexrol2, sexrol3, sexrol4, sexrol5rev),
                                    na.rm = TRUE
                                    ),
                                    
                    ) |> 
                    ungroup()
```

Check the scale.

```{r}
sbs399_sexroles |>
  skim(sexrole_scale, sexrole_scale2)
```
