---
title: Factor analysis (FA)
---

## Introduction

See slides.

## R

You need to learn about the following R functions for factor analysis. They are part of the `psych` package, which needs to be installed and loaded. Optionally you may want to check principal component analysis.

| R             | Description                                        |
|---------------|----------------------------------------------------|
| `fa()`        | factor analysis by maximum likelihood              |
| `scree()`     | plot of eigenvalues to determine number of factors |


Examples:

Setup data on sex role attitudes

```{r, eval=FALSE}
library(psych)
library(haven)

data <- read_dta("data/sbs399_sexroles.dta")
# this dataset contains 5 items about sex role attitudes, in a 5-point Likert format
# Important: in the code below we use the entire data object because it already contains only the 5 items. 
# If you have a larger dataset, you need to select the relevant items first.
```

Examine correlation between items:

```{r, eval=FALSE}
cor(
  data,
  use = "pairwise.complete.obs"
)
```

Determine the number of factors graphically with a scree plot:

```{r, eval=FALSE}
scree(data)
```

Factor analysis using maximum likelihood:

```{r, eval=FALSE}
fa_fit <- fa(
  data,
  nfactors = 2,
  fm = "ml"
)
```

Rotate common factors to ease interpretation. 

```{r, eval=FALSE}
fa_varimax <- fa(
  data,
  nfactors = 2,
  fm = "ml",
  rotate = "varimax"
)

fa_oblique <- fa(
  data,
  nfactors = 2,
  fm = "ml",
  rotate = "promax"
)

```

Print the factor loadings (pattern matrix):
```{r, eval=FALSE}
fa_varimax$loadings
```

Print the structure matrix:
```{r, eval=FALSE}
fa_varimax$Structure
```

Get factor scores: 
```{r, eval=FALSE}
fa_scores <- fa_varimax$scores
```


## Assignments

### Muslim rights

Use the dataset `muslimrights`. The dataset contains two sets of items.

Attitudes towards Muslim expressive rights was measured with seven statements. Four example items were:

- The right to establish Islamic schools should always exist in the Netherlands.
- Some Islamic holidays should become official Dutch holidays.
- Dutch TV should broadcast more programmes by and for Muslims.
- In the Netherlands the wearing of a headscarf should not be forbidden.

National identification was measured by six statements:

- My Dutch identity is an important part of my self (iden1);
- I identify strongly with the Netherlands (iden2);
- I feel a strong attachment to the Netherlands (iden3);
- I am proud of my Dutch background (iden4);
- Being Dutch is a very important part of how I see myself (iden5); and
- I feel a strong sense of belonging to the Netherlands (iden6).

We will focus on the first set of items, `right1` to `right7`.

a. Check the descriptives. Check the frequencies. Are the variables approximately normally distributed? Why is this important?
b. Do an EFA/ML of the data, first with an unrotated solution (`rotate = "none"`). Then apply a varimax rotation.
c. Provide interpretations:
  - For the moment we choose two dimensions, even though the scree plot might suggest a different number. In the unrotated solution, how much variance is explained by factor 1 and factor 2? And in the rotated solution? Verify that the sum of the variances of factors 1 and 2 is the same for the unrotated and rotated solutions.
  - Which variable is not very well explained by the two factors?
  - Interpret the two factors, using the varimax rotation. 
    - Tip: plotting the factor loadings may help you interpret the factors. Assuming the object created with `fa()` is called `fa_varimax`, you can plot the loadings with `plot(fa_varimax, labels = rownames(fa_varimax$loadings))`.
  - If you would make two sum scores for the items, which variables go into sum score 1 and which in sum score 2?
  - Apply a promax rotation. Provide interpretations:
    - What is the correlation between the factors?
    - Interpret the two factors.
    - Which variables go into each of the factors?
d. Oversee all analyses that you carried out. What is your final conclusion with respect to the number of dimensions (factors)? Assume for the moment that you choose two dimensions, what is your final conclusion with respect to the interpretation of the dimensions?
e. Optional. Analyze the 6 items about national identification in the dataset.


### Bem Sex Role Inventory

BSRI stands for the Bem Sex Role Inventory (Bem, 1974). We have data from a sample of 369 middle-class, English speaking women between the ages of 21 and 60 who were interviewed face-to-face. The data are provided by Tabacknick and Fidell (2007), *Using multivariate statistics*, 5nd ed., Boston: Pearson. The full inventory takes 60 items. Tabachnick and Fidell state that in our data there are 20 items measuring the overall trait femininity, 20 items the overall trait masculinity, and 5 items measuring the overall trait social desirability. The question format is that respondents are asked to attribute the 60 traits to themselves using scores going from 1 (“never or almost never true of me”) to 7 (“always or almost always true of me”). It turns out that the data are not thoroughly cleaned, as occasionally there appears a 0 and a 9 in the data file. We ignore this problem. 

- Do an initial check on the data using frequencies and descriptives. Notice that some of the variables are skewed. Skewness of items lowers the maximum that correlations can attain.
- Produce a scree plot.  It is clear that that the elbow criterion is not easy to use here. Clearly there are two outstanding factors. If we take factor 3, then we should also take 4 and 5 as their eigenvalues are close to the eigenvalue of 3. We may even consider factors 6 and 7. We will have a look at two different solutions: two factors and five factors.
- Do an EFA/ML with two factors with a promax rotation. Notice that the correlation between the rotated factors is relatively low. Interpret the two factors.
- Do another EFA/ML, now with five factors and a promax rotation. Interpret the five factors using only those variables that have loadings larger than .40. You can do this by using  `print(your_fa_object$loadings, cutoff = .40)`.

We stop here. We have seen that interpreting the factor structure for such a large list of items is difficult and sometimes somewhat arbitrary. More insight could be found from validating the factors in experimental research and by correlating the factors to other variables. Of course, this is beyond the scope of this course.

::: {.content-visible when-format="html"}

## Answers

::: {.callout-note appearance="simple" collapse=true}
## See answers

```{r setup-factor-analysis}
library(tidyverse)
library(haven)
library(psych)
library(skimr)
```

### Muslim rights

```{r}
muslimrights <- read_dta("data/muslimrights.dta")

rights_items <- muslimrights |> 
               select(right1:right7)
```

Descriptives and frequencies. ML factor analysis assumes multivariate normality, so with clearly skewed the ML significance tests should be interpreted with some caution.

```{r}
rights_items |> 
  skim()
```

Determine the number of factors with a scree plot. One might be enough by the eigenvalue > 1 criterion, but we will use two factors for the sake of illustration.

```{r}
cor(
  rights_items,
  use = "pairwise.complete.obs"
)

scree(rights_items)
```


```{r}
fa_muslim_unrotated <- fa(
  rights_items,
  nfactors = 2,
  fm = "ml",
  rotate = "none"
)

fa_muslim_varimax <- fa(
  rights_items,
  nfactors = 2,
  fm = "ml",
  rotate = "varimax"
)
```

Compare the unrotated and rotated (varimax) solutions:
```{r}
fa_muslim_unrotated
fa_muslim_varimax
```

Note that the sum of the variances of factors 1 and 2 is the same for the unrotated and rotated solutions. Also note that item `right5` has the lowest communality (`h2`), and correspondingly the highest uniqueness (`u2`), indicating that this item is the least explained by the two factors.

Plot the factor loadings for the varimax-rotated solution:
```{r}
fa.plot(fa_muslim_varimax, labels = rownames(fa_muslim_varimax$loadings))
```


Oblique (promax) rotation:

```{r}
fa_muslim_promax <- fa(
  rights_items,
  nfactors = 2,
  fm = "ml",
  rotate = "promax"
)

fa_muslim_promax
```

Correlation between factors is `{r} round(fa_muslim_promax$Phi[1,2], 2)`


### Bem Sex Role Inventory 

```{r}
bsri <- read_dta("data/bsri.dta")
```


```{r}
bsri |>
  skim()
```

Use a scree plot to judge how many factors to retain:

```{r}
scree(bsri)
```

Two-factor solution with a promax rotation:

```{r}
fa_bsri_2 <- fa(
  bsri,
  nfactors = 2,
  fm = "ml",
  rotate = "promax"
)

fa_bsri_2
```

Five-factor solution with a promax rotation, showing only loadings larger than .40:

```{r}
fa_bsri_5 <- fa(
  bsri,
  nfactors = 5,
  fm = "ml",
  rotate = "promax"
)
print(fa_bsri_5$loadings, cutoff = .40)
```

:::
:::