---
title: Some programming features in R
---

## Objects

Objects allow you to abbreviate commands considerably, making code more readable, and improving consistency. For instance, we can define an object `C` with a set of variables to be used as controls in a series of models. Using an object ensures you use the same control variables everywhere; changing the controls in all models is accomplished by editing one line of code.

```{r, eval=FALSE}
C <- c("sex", "age", "married")
```

## Repeating commands

Often, you want to apply the same operation to several variables. You can write out the commands by hand, but this is repetitive and easy to mistype.

1a. Repeat a variable construction for multiple variables

```{r, eval=FALSE}
data <- data |>
  mutate(
    rgood1 = 1 - good1,
    rgood2 = 1 - good2,
    rgood3 = 1 - good3,
    rgood4 = 1 - good4,
    rgood5 = 1 - good5
  )
```

This works, but it is not very efficient. If you later discover that the transformation should be different, you have to change five lines of code.

1b. Equivalent, using `across()`

```{r, eval=FALSE}
data <- data |>
  mutate(
    across(
      good1:good5,
      ~ 1 - .x,
      .names = "r{.col}"
    )
  )
```

The function `across()` applies the same operation to several variables. Here, `good1:good5` selects the variables from `good1` to `good5`. The expression `~ 1 - .x` means: take each selected variable and compute `1 - variable`. The argument `.names = "r{.col}"` creates new variable names by adding `r` before the original variable name.

1c. Equivalent, selecting variables by name pattern

```{r, eval=FALSE}
data <- data |>
  mutate(
    across(
      starts_with("good"),
      ~ 1 - .x,
      .names = "r{.col}"
    )
  )
```

This does the same thing, but selects variables by name. All variables whose names start with good are transformed. This is useful when there are many variables with a shared naming pattern.

1d. Equivalent, looping over a numerical range

```{r, eval=FALSE}
for (i in 1:5) {
  data[[paste0("rgood", i)]] <- 1 - data[[paste0("good", i)]]
}
```

This loop repeats the same command for `i = 1, 2, 3, 4, 5`. The function `paste0()` creates variable names such as `good1` and `rgood1`. This approach is useful, but for data manipulation in the tidyverse, `across()` is often easier to read.

2.

```{r, eval=FALSE}
for (i in 1:5) {

  data |>
    filter(country == i) |>
    summarise(
      mean_income = mean(income, na.rm = TRUE)
    )

  data |>
    filter(country == i) |>
    count(religion)

}
```

This loop repeats two analyses for each value of `country` from 1 to 5. For each country, it computes the mean income and creates a frequency table of religion. This is useful when you want to run the same checks separately for multiple groups.

## Writing your own functions

If you repeat the same block of code many times, it can be useful to write your own function. A function takes inputs, does something with them, and returns an output.

```{r, eval=FALSE}
rob_regress <- function(formula, data) {

  lm(
    formula = formula,
    data = data
  )

}
```

This function is called `rob_regress`. It takes two inputs: a model formula and a dataset. Inside the function, `lm()` estimates a linear regression model.

```{r, eval=FALSE}
rob_regress(
  inc ~ edu + age,
  data = data
)
```

This runs the function with income as the dependent variable and education and age as explanatory variables.
